### 2017-01-16 Paper_v1.tex
What the what

\documentclass[10pt,twocolumn,twoside,final]{IEEEtran}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{amssymb,amsmath}

\bibliographystyle{plain}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% *** SUBFIGURE PACKAGES ***
\ifCLASSOPTIONcompsoc
  \usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
\else
  \usepackage[caption=false,font=footnotesize]{subfig}
\fi

\begin{document}

\title{An Effective Model of HL-60 Differentiation}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Ryan Tasseff\IEEEauthorrefmark{1},
Holly A. Jensen\IEEEauthorrefmark{1},
Johanna Congleton\IEEEauthorrefmark{2},
Andrew Yen\IEEEauthorrefmark{2},
Jeffrey D. Varner\IEEEauthorrefmark{1}}\\
\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Chemical and Biomolecular Engineering,}
\IEEEauthorblockA{\IEEEauthorrefmark{2}Department of Biomedical Sciences, Cornell University, Ithaca, NY 14850 USA}
%\thanks{Manuscript received  1, 2012; revised September 17, 2014.
Corresponding author: J. Varner (email:jdv27@cornell.edu).}

%One of the defining features of this program is slowly induced, persistent MAPK activation.
%The molecular mechanisms of ATRA-induced commitment, arrest and functional differentiation are only partially understood.
%In this study, we explored the ATRA-inducible c-RAF, also known as RAF1 (RAF),
%interactome to determine the functional and regulatory architecture responsible for persistent MAPK activation in HL-60 cells.
%We surveyed a panel of 19 possible RAF interaction partners in the presence or absence of ATRA and the RAF inhibitor GW5074.
%We found five proteins (Akt, CK2, 14-3-3, Src and Vav1) that interacted with RAF under different conditions.
%In particular, the interaction between RAF and Vav1 demonstrated a relationship with ATRA-inducible MAPK activation and aspects of functional differentiation.

%this hypothesis using a combination of experimental and computational tools.
%First, we explored the  RAF interactome
%by surveying a panel of 19 possible binding partners using immunoprecipitation (IP),
%with and without ATRA and the RAF inhibitor GW5074.
%Initially, we expected increased ATRA-dependent association between RAF and kinases linked to BLR1 activity;
%however, this was not supported by data.
%Instead, we found that the interaction between the guanine nucleotide exchange factor Vav1 and RAF was both ATRA-inducible and simultaneously sensitive to RAF inhibition.
%Next, we considered how MAPK activation and differentiation were affected by the inhibition of RAF kinase activity in the presence and absence of ATRA.
%We showed that RAF activity was directly proportional to ERK phosphorylation and to functional differentiation processes such as the generation of reactive oxygen species (ROS).
%Moreover, interactions between RAF and kinase partners such as Akt or CK2, or the scaffolding protein 14-3-3 were largely insensitive to ATRA treatment.
%These studies established the working hypothesis that Vav1 (or potentially other ATRA-inducible proteins) acted as
%limiting members of a constitutively assembled trigger complex that propelled sustained MAPK activation, arrest and differentiation.
%We tested this hypothesis by constructing a mechanistic mathematical model of the RAF-Vav1 circuit,
%based on the IP and Western blot data presented in this study,
%and from previous literature.
%Interestingly, cells treated with ATRA for time periods shorter than the commitment phase retain a limited inheritable memory,
%which reduces the time required to reach commitment during subsequent ATRA exposure \cite{YEN1984}.

\maketitle

\begin{abstract}
We present an effective model All-Trans Retinoic Acid (ATRA)-induced differentiation of HL-60 cells.
The model describes a key architectural feature of ATRA-induced differentiation,
positive feedback between an ATRA-inducible signalsome complex involving many proteins including Vav1, a guanine nucleotide exchange factor,
and the activation of the mitogen activated protein kinase (MAPK) cascade.
The model, which was developed by integrating logical rules with kinetic modeling, was significantly smaller than previous models.
However, despite its simplicity, it captured key features of ATRA induced differentiation of HL-60 cells.
We identified an ensemble of effective model parameters using measurements taken from ATRA-induced HL-60 cells.
Using these parameters, model analysis predicted that MAPK activation was bistable as a function of ATRA exposure.
Conformational experiments supported ATRA-induced bistability.
These findings, combined with other literature evidence, suggest that positive feedback is central to a diversity of cell fate programs.
\end{abstract}

\begin{IEEEkeywords}
Mathematical modeling, systems biology
\end{IEEEkeywords}

%If differentiation programs could be rationally manipulated, regenerative treatments might be possible for cancers, spinal cord injuries and neurodegenerative disorders.
%However, we must first understand the organization and operation of these programs \cite{Young2011}.

\section{Introduction}
Understanding differentiation programs is an important
therapeutic challenge. Toward this challenge, lessons learned
in model systems, such as the lineage-uncommitted human
myeloblastic cell line HL-60, informs our analysis of more
complex therapeutically important programs. Patient derived HL-60 leukemia  cells have been
a durable experimental model since the 1970’s \cite{Breitman1980}.
HL-60 undergoes cell cycle arrest and either myeloid or monocytic differentiation following stimulation;
All-Trans Retinoic Acid (ATRA) induces G1/G0-arrest and myeloid differentiation in HL-60 cells,
while 1,25-dihydroxy vitamin D3 (D3) induces arrest and monocytic differentiation.
Commitment to cell cycle arrest and differentiation requires approximately 48 hr of treatment,
during which HL-60 cells undergo two division cycles.

\begin{figure}[!t]
\includegraphics[width=0.50\textwidth]{./figs/Fig-1-Network_v2.pdf}
\caption{Schematic of the effective differentiation circuit.
A: Above a critical threshold, ATRA activates an upstream Trigger, which induces signalsome complex formation.
Signalsome activates the mitogen-activated protein kinase (MAPK) cascade which in turn
drives differentiation program and signalsome formation.
B: Signalsome components that interact with RAF:
first column shows Western blot analysis on total RAF immunoprecipitation with and without 24 hr ATRA treatment,
and the second on total lysate.
C: Effect of the RAF inhibitor GW5074 on RAF interactions as determined by Western blot analysis of total RAF immunoprecipitation.}\label{fig:network}
\end{figure}

Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
ATRA drives sustained MEK-dependent activation of the RAF/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
MEK inhibition results in the loss of ERK and RAF phosphorylation, and the failure to arrest and differentiate \cite{Hong2001}.
ATRA (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
RAR/RXR activation is necessary for ATRA-induced RAF phosphorylation  \cite{Hong2001},
and the formation of the ATRA-inducible signalsome complex which drives differentiation.
The signalsome is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, as well as IRF-1 transcription factors \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
This signaling is driven by ATRA-induced expression of CD38 and the putative heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
BLR1, identified as an early ATRA (or D3)-inducible gene \cite{YEN1990},
is necessary for MAPK activation and differentiation \cite{WANG2004}.
Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
BLR1 overexpression enhanced RAF phosphorylation and accelerated terminal differentiation, while RAF inhibition reduced BLR1 expression and differentiation \cite{Wang2008}.
BLR1 knock-out cells failed to activate RAF or differentiate in the presence of ATRA \cite{Wang2008}.

In this study, we developed a mathematical model of the key architectural feature of ATRA induced differentiation of HL-60 cells,
namely positive feedback between an ATRA-inducible signalsome complex and MAPK activation.
Previously Tasseff et al. hypothesized that signalsome-MAPK positive feedback was essential for ATRA-induced
cell cycle arrest and differentiation \cite{Tasseff2011}.
We explored this hypothesis by constructing a minimal model containing only signalsome and MAPK components (Fig. \ref{fig:network}A).
The effective model was developed using a novel framework which integrated logical rules with kinetic modeling.
This formulation significantly reduced the size of the model compared to the previous study of Tasseff et al., while maintaining similar model performance \cite{Tasseff2011}.
The effective model, despite its simplicity, captured key features of ATRA induced differentiation of HL-60 cells.
We identified an ensemble of effective model parameters using measurements taken from ATRA-induced HL-60 cells.
Using these parameters, model analysis predicted the bistability of MAPK activation as a function of ATRA exposure.
Conformational experiments supported ATRA-induced bistability.
These findings, combined with other literature evidence, suggests that positive feedback architectures are central to
many cell fate programs.

%We tested this prediction using ATRA washout experiments.
%The effective model encoded the positive feedback between the ATRA-inducible membrane localized signalsome complex and MAPK activation.
%The effective model was identified and tested using measurements from ATRA-induced HL-60 cells.
%The effective model, despite its simplicity, captured key features of the ATRA response of HL-60 cells such as
%sustained MAPK activation, the impact of gene deletion and inhibition of the MAPK pathway.

%based on a comprehensive mathematical model of ATRA-induced siganling

%\begin{figure*}[!t]
%	\centering
%	\includegraphics[width=\textwidth]{./figs/Fig-1-Network.pdf}
%	\caption{Schematic of the two component ATRA induced HL-60 differentiation circuit.}
%\end{figure*}

\begin{figure}[!t]\centering
\includegraphics[width=0.48\textwidth]{./figs/Fig-2-cRAF-BLR1-Fit-Analysis.pdf}
\caption{Model analysis for ATRA-induced HL-60 differentiation.
A: BLR1 mRNA versus time following exposure to 1$\mu$M ATRA at t = 0 hr.
B: cRAF-pS621 versus time following exposure to 1$\mu$M ATRA at t = 0 hr.
Points denote experimental measurements, solid lines denote the mean model performance. Shaded regions denote the 99\% confidence interval calculated over the parameter ensemble.
C: Signalsome and cRAF-pS621 nullclines for ATRA below the critical threshold.
The model had two stable steady states and a single unstable state in this regime.
D: Signalsome and cRAF-pS621 nullclines for ATRA above the critical threshold.
In this regime the model had only a single stable steady state.
E: Morphology of HL-60 as a function of ATRA concentration (t = 72 hr). }\label{fig:model-fitting}
\end{figure}

%Furthermore, the Vav1 and Src associations were correlated with cRAF phosphorylated at Serine 621 (activated; cRAF-pS621) abundance in the precipitate.
%Interestingly, the cRAF-Akt interaction qualitatively increased following treatment with GW5074;
%however, it remained unaffected by treatment with ATRA.

\section{Results}

Vav1 is a member of an ATRA-inducible signalsome complex which propels sustained MAPK activation, arrest and differentiation (Fig. \ref{fig:network}B).
We conducted immunoprecipitation studies and identified a limited number of ATRA-dependent and -independent RAF interaction partners.
Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with cRAF,
suggesting a direct physical interaction was possible.
However, only the associations between cRAF and Vav1 and Src were ATRA-inducible (Fig. \ref{fig:network}B).
Others proteins e.g., CK2, Akt and 14-3-3, generally bound cRAF regardless of phosphorylation status or ATRA treatment.
Treatment with the RAF kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 and Src with cRAF (Fig. \ref{fig:network}C),
although the signal intensity for Src was weak.
However, GW5074 did not influence the association of CK2 or 14-3-3 with cRAF,
further demonstrating their independence from cRAF phosphorylation.
Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with cRAF activation following ATRA-treatment.

The model recapitulated sustained signalsome/MAPK activation following exposure to 1$\mu$M ATRA (Fig. \ref{fig:model-fitting}A-B).
An ensemble of effective model parameters was estimated by minimizing the difference between simulations and time-series measurements of BLR1 mRNA and
cRAF-pS621 following the addition of 1$\mu$M ATRA using particle swarm optimization (PSO).
We focused on the S621 phosphorylation site of cRAF since enhanced phosphorylation at this site is a defining characteristic of  sustained MAPK activation in HL-60.
Each particle in the swarm contributed a member to the parameter ensemble.
The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A) and sustained phosphorylation of cRAF-pS621 (Fig. \ref{fig:model-fitting}B)
in a growing population of HL-60 cells.
However, the effective model failed to capture the decline of BLR1 expression after 48 hr of ATRA exposure.

The model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
Nullcline analysis predicted two stable steady-states and a single unstable state when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C).
In the lower stable state, neither the signalsome nor cRAF-pS621 were present (thus, the differentiation program was deactivated).
However, at the high stable state, both the signalsome and cRAF-pS621 were present, allowing for sustained activation and differentiation.
Interestingly, when ATRA was above a critical threshold, only the activated state was accessible (Fig. \ref{fig:model-fitting}D).
To test these findings, we first identified the ATRA threshold. We exposed HL-60 cells to different ATRA concentrations for 72 hr (Fig. \ref{fig:model-fitting}E).
Morphological changes associated with differentiation were visible for ATRA $\geq$ 0.25 $\mu$M, suggesting the critical ATRA
threshold was near this concentration.

%\begin{figure}[!t]\centering
%\includegraphics[width=0.5\textwidth]{./figs/Fig-3-QualitativeBehavior.pdf}
%\caption{Qualitative analysis of the effective HL-60 differentiation model. A: Signalsome and cRAF-pS621 nullclines for ATRA below the critical limit. The reduced order model has two stable steady states and a single unstable state in this regime. B: Signalsome and cRAF-pS621 nullclines for ATRA above the critical limit. The reduced order model has only a single stable steady state in this regime. C: HL-60 morphological response to increasing ATRA concentration. }\label{fig:bifurcation-results}
%\end{figure}
%Taken together, nullcline analysis suggested qualitatively different behavior was possible depending upon the
%degree of ATRA forcing. Below a critical threshold, both deactivated and activated behavior were possible, while above the threshold, only activated behavior was possible.
%However, beyond 144 hr the activated MAPK signal (as measured by phosphorylated ERK1/2) decayed. The decreasing MAPK signal indicated additional factors and connections were likely involved in the ATRA circuit beyond the signalsome and MAPK.

Next, we tested that a cell locked into an activated state remained activated following ATRA withdraw.
Sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
After activation, if we inhibited or removed elements from the effective circuit we expected the siganlsome and MAPK signals to decay.
We simulated ATRA induced activation in the presence of kinase inhibitors, and without key circuit elements (Fig. \ref{fig:model-predictions}).
Consistent with experimental results using multiple MAPK inhibitors, ATRA activation in the presence of MAPK inhibitors lowered the steady-state value of signalsome (Fig. \ref{fig:model-predictions}A). In the presence of BLR1, the signalsome and cRAF-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, blue).
On the other hand, BLR1 deletion removed the ability of the circuit to maintain a sustained MAPK response following the withdraw of ATRA (Fig. \ref{fig:model-predictions}B, gray).
Lastly, washout experiments in which cells were exposed to 1$\mu$M ATRA for 24 hr, and then transferred to fresh media without ATRA, confirmed the persistence of the self sustaining activated state for up to 144 hr (Fig. \ref{fig:model-predictions}C).

\section{Discussion}
We presented an effective model of ATRA-inducible differentiation of HL-60 cells which
encoded positive feedback between the ATRA-inducible signalsome complex and the MAPK pathway.
Despite its simplicity, the model captured key features of the ATRA induced differentiation such as
sustained MAPK activation, and bistability with respect to ATRA exposure. We also reported a new ATRA-inducible component of the signalsome, Vav1.
Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
The Vav1/RAF association correlated with RAF activity, was ATRA-inducible and decreased after treatment with GW5074.
The presence of Vav1 in RAF/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
While its function in the signalsome is unclear, Vav1 has been shown to associate with a Cbl-Slp76-CD38 complex in an ATRA-dependent manner;
furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevented
ATRA-induced MAPK activation \cite{Shen2009}. Thus, interaction of Cbl-Slp76-Vav1 and CD38 appears to be required for transmission of the ATRA signal by the
signalsome.

Several engineered, or naturally occurring systems involved in cell fate decisions incorporate positive feedback and bistability \cite{Ferrell:2002aa}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the differentiation circuit presented here; ATRA drives signalsome which activates MAPK, cell-cycle arrest, differentiation and signalsome.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, they share similar cell fate decision architectures.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while positive feedback is sometimes not desirable in man made systems, it may be at the core of a diverse variety of
cell fate programs and other networks important to human health.

Model performance was impressive given its limited size. However, there were several issues to explore further.
First, there was likely missing connectivity in the effective differentiation circuit.
Decreasing BLR1 expression with simultaneously sustained cRAF-pS261 activation was not captured by the current network architecture.
This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRAF-pS261 abundance.
We could model this by separating siganlsome formation into an inactive precursor pool that is transformed to a long-lived activated siganlsome by MAPK activation.
We should also explore adding additional downstream biological modules to this skeleton model, for example the upregulation of reactive oxygen markers such as p47Phox or
cell cycle arrest components to capture the switch from an actively proliferating population to a population in G0-arrest.
Next, the choice of max/min integration rules or the particular form of the transfer functions could also be explored.
Integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules might have different properties which could influence model identification or performance.
For example, a mean integration rule would be differentiable, allowing derivative-based optimization approaches to be used.
The form of the transfer function could also be explored. We choose hill-like functions because of their
prominence in the systems and synthetic biology community. However, many other transfer functions are possible.

\section{Materials and Methods}

\begin{figure}[!t]\centering
\includegraphics[width=0.50\textwidth]{./figs/Fig-4-Predictions.pdf}
\caption{Model simulation follow exposure to 1$\mu$M ATRA.
A: BLR1 mRNA versus time with and without MAPK inhibitor.
B: cRAF-pS621 versus time following pulsed exposure to 1$\mu$M ATRA with and without BLR1.
Solid lines denote the mean model performance, while shaded regions denote the 99\% confidence interval calculated over the parameter ensemble.
C: Western blot analysis of phosphorylated ERK1/2 in ATRA washout experiments.
Experimental data in panels A and B were reproduced from Wang and Yen \cite{Wang2008}, data in panel C is reported in this study. }\label{fig:model-predictions}
\end{figure}

\noindent\subsubsection*{Effective model equations}
The model was constructed using the hybrid approach of Wayman et al. \cite{pr3010138}.
The abundance of species $i$ ($x_{i}$) is governed by:
\begin{equation}
	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
\end{equation}
where $\mathcal{R}$ and $\mathcal{M}$ denotes the number of reactions and species in the model, $\mu$ denotes the specific growth rate and
$k_{d,i}$ denotes a degradation constant for species $i$. The $\mu x_{i}$ term, which accounts for dilution due to cell growth, decays to zero following ATRA exposure.
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$,
$r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$, and $\mathbf{k}$ ($\mathcal{K}\times{1}$) denotes the unknown kinetic parameter vector.
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$,
while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right) = \bar{r}_{j}v_{j}$.
In this study, we used either zero- or first-order kinetics.
The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j} = \mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. Each regulatory transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right) = k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
\end{equation}where $\mathcal{Z}_{i}$ denotes the abundance factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
In this study, we used $\mathcal{I}_{j}\in\left\{\max,\min\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j} = 1$.

\noindent\subsubsection*{Estimation of model parameters}
Model parameters were estimated by minimizing the squared difference between simulations and experimental data set $j$:
\begin{equation}\label{eqn:objective-function}
	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
\end{equation}
The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs at time $i$ from training set $j$,
where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
The first term in Eqn. \eqref{eqn:objective-function} quantified the relative simulation error.
We used immunoblot intensity measurements for model training.
Thus, we trained the model on the \emph{relative} change between bands within each data set.
Suppose we have the intensity of species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$.
The scaled value $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ is given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
\end{equation}
where $\hat{\mathcal{M}}_{ij}=0$ and $\hat{\mathcal{M}}_{ij}=1$  describe the lowest (highest) intensity bands.
A similar scaling was used for the simulation output. The second term in the objective function ensured a realistic concentration scale was estimated by the model.
We set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
We minimized the total model residual $\sum_{j}E_{j}$ using particle swarm optimization (PSO).
All code was implemented in the Octave programming language, and is available for
download from http://www.varnerlab.org.

%Model parameters were estimated by minimizing the difference between simulations and experimental measurements (squared residual):
%\begin{equation}\label{eqn:objective-function}
%	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
%\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
%value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.

\subsubsection*{Cell culture and treatment}
Human myeloblastic leukemia cells (HL-60 cells) were grown in a humidified atmosphere of 5\% CO$_2$ at 37$^{o}$C and maintained in RPMI 1640 from Gibco (Carlsbad, CA)
supplemented with 5\% heat inactivated fetal bovine serum from Hyclone (Logan, UT) and 1$\times$ antibiotic/antimicotic (Gibco, Carlsbad, CA).
Cells were cultured in constant exponential growth \cite{Brooks1996}.
Experimental cultures were initiated at $0.1\times10^6$ cells/mL 24 hr prior to ATRA treatment;
if indicated, cells were also treated with GW5074 (2$\mu$M) 18 hr before ATRA treatment.
For the cell culture washout experiments, cells were treated with ATRA for 24 hr,
washed 3x with prewarmed serum supplemented culture medium to remove ATRA,
and reseeded in ATRA-free media as described.
Western blot analysis was performed at incremental time points after removal of ATRA.

\subsubsection*{Chemicals}
All-Trans Retinoic Acid (ATRA) from Sigma-Aldrich (St. Louis, MO) was dissolved in
100\% ethanol with a stock concentration of 5mM, and used at a final concentration of 1$\mu$M (unless otherwise noted).
The cRAF inhibitor GW5074 from Sigma-Aldrich (St. Louis, MO) was dissolved in DMSO with a stock concentration of 10mM,
and used at a final concentration of 2$\mu$M.
HL-60 cells were treated with 2$\mu$M GW5074 with or without ATRA (1$\mu$M) at 0 hr.
This GW5074 dosage had a negligible effect on the cell cycle distribution, compared to ATRA treatment alone.

\subsubsection*{Immunoprecipitation and western blotting}
Approximately $1.2 \times 10^7$ cells were lysed using 400$\mu$L of M-Per lysis buffer from Thermo Scientific (Waltham, MA).
Lysates were cleared by centrifugation at 16,950 $\times$ g in a micro-centrifuge for 20 min at 4$^{o}$C.
Lysates were pre-cleared using 100$\mu$L protein A/G Plus agarose beads from Santa Cruz Biotechnology (Santa Cruz, CA) by
inverting overnight at 4$^{o}$C.  Beads were cleared by centrifugation and total protein concentration
was determined by a BCA assay (Thermo Scientific, Waltham, MA).  Immunoprecipitations were setup by bringing
lysate to a concentration of 1g/L in a total volume of 300$\mu$L (M-Per buffer was used for dilution).
The anti-RAF antibody was added at 3$\mu$L.
A negative control with no bait protein was also used to exclude the direct interaction of proteins with the A/G beads.
After 1 hr of inversion at 4$^{o}$C, 20$\mu$L of agarose beads was added and samples were left
to invert overnight at 4$^{o}$C. Samples were then washed three times with M-Per buffer by centrifugation.
Finally proteins were eluted from agarose beads using a laemmli loading buffer.
Eluted proteins were resolved by SDS-PAGE and Western blotting.  Total
lysate samples were normalized by total protein concentration (20$\mu$g per sample)
and resolved by SDS-PAGE and Western blotting.  Secondary HRP bound antibody was used for visualization.
All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of $\alpha$-p621 RAF which was purchased from Biosource/Invitrogen (Carlsbad, CA),
and $\alpha$-CK2 from BD Biosciences (San Jose, CA).

\subsubsection*{Morphology assessment}
Untreated and ATRA-treated HL-60 cells were collected after 72 hr and cytocentrifuged for 3 min at 700 rpm onto glass slides.
Slides were air-dried and stained with Wright’s stain. Slide images were captured at 40X (Leica DM LB 100T microscope, Leica Microsystems).

% References -
\bibliography{Paper_v1}

\end{document}

### 2017-01-16 Paper_v2.tex
What the what

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[round,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\usepackage{algorithm2e}
\usepackage{algpseudocode}
%\usepackage[space]{cite}
\urlstyle{rm}

%\textwidth = 6.50 in
%\textheight = 9.5 in
%\oddsidemargin =  0.0 in
%\evensidemargin = 0.0 in
%\topmargin = -0.50 in
%\headheight = 0.0 in
%\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
\doublespace

%\bibliographystyle{chicago}
\bibliographystyle{plos2009}

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\large\bfseries}}
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

% Change the number format in the ref list -
\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Fig.
\renewcommand{\figurename}{Fig.}

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {An Effective Model of the Retinoic Acid Induced HL-60 Differentiation Program}}}
\vspace{0.05in}
{\par \centering \large{Ryan Tasseff, Holly A. Jensen, Johanna Congleton$^{\dag}$, Wei Dai, Andrew Yen$^{\dag}$, and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering} and {$^{\dag}$Department of Biomedical Sciences, {Cornell University, Ithaca NY 14853}}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Effective modeling of HL-60 differentiation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Frontiers~in~Systems~Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak

\section*{Abstract}
In this study, we present an effective model All-Trans Retinoic Acid (ATRA)-induced differentiation of HL-60 cells.
The model describes a key architectural feature of ATRA-induced differentiation,
positive feedback between an ATRA-inducible signalsome complex involving many proteins including Vav1, a guanine nucleotide exchange factor,
and the activation of the mitogen activated protein kinase (MAPK) cascade.
The model, which was developed by integrating logical rules with kinetic modeling, was significantly smaller than previous models.
However, despite its simplicity, it captured key features of ATRA induced differentiation of HL-60 cells.
We identified an ensemble of effective model parameters using measurements taken from ATRA-induced HL-60 cells.
Using these parameters, model analysis predicted that MAPK activation was bistable as a function of ATRA exposure.
Conformational experiments supported ATRA-induced bistability.
These findings, combined with other literature evidence, suggest that positive feedback is central to a diversity of cell fate programs.

\clearpage

\linenumbers

\section*{Introduction}
Understanding differentiation programs is an important
therapeutic challenge. Toward this challenge, lessons learned
in model systems, such as the lineage-uncommitted human
myeloblastic cell line HL-60, informs our analysis of more
complex therapeutically important programs. Patient derived HL-60 leukemia cells have been
a durable experimental model since the 1970’s \cite{Breitman1980}.
HL-60 undergoes cell cycle arrest and either myeloid or monocytic differentiation following stimulation;
All-Trans Retinoic Acid (ATRA) induces G1/G0-arrest and myeloid differentiation in HL-60 cells,
while 1,25-dihydroxy vitamin D3 (D3) induces arrest and monocytic differentiation.
Commitment to cell cycle arrest and differentiation requires approximately 48 hr of treatment,
during which HL-60 cells undergo two division cycles.

Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
ATRA drives sustained MEK-dependent activation of the RAF/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
MEK inhibition results in the loss of ERK and RAF phosphorylation, and the failure to arrest and differentiate \cite{Hong2001}.
ATRA (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
RAR/RXR activation is necessary for ATRA-induced RAF phosphorylation  \cite{Hong2001},
and the formation of the ATRA-inducible signalsome complex which drives differentiation.
The signalsome is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, as well as IRF-1 transcription factors \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
This signaling is driven by ATRA-induced expression of CD38 and the putative heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
BLR1, identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
is necessary for MAPK activation and differentiation \cite{WANG2004}.
Studies of the BLR1 promoter identified a 5' 17bp GT box approximately 1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
BLR1 overexpression enhanced RAF phosphorylation and accelerated terminal differentiation, while RAF inhibition reduced BLR1 expression and differentiation \cite{Wang2008}.
BLR1 knock-out cells failed to activate RAF or differentiate in the presence of ATRA \cite{Wang2008}.
Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient to explain ATRA-induced sustained MAPK activation and the expression of differentiation markers \cite{Tasseff2011}.
Model analysis also suggested that Raf was the most distinct of the MAPK proteins.

A critical question is what other components of the MAPK positive feedback circuit are required to drive ATRA-induced functional differentiation of HL-60 cells.
Wang and Yen showed that ectopic expression of the constitutively active CR3 domain of Raf1 restored ATRA-induced G0 arrest and differentiation in BLR1 knock-out cells \cite{Wang2008}.
However, ectopic expression of Raf1 CR3 alone, in the absence of ATRA, failed to induce arrest or differentiation.
Thus, additional ATRA-inducible components must exist,
which independently promote arrest and differentiation in the absence of BLR1.
In this study, we developed a mathematical model of the key architectural feature of ATRA induced differentiation of HL-60 cells,
namely positive feedback between an ATRA-inducible signalsome complex and MAPK activation.
Previously Tasseff et al. hypothesized that signalsome-MAPK positive feedback was essential for ATRA-induced
cell cycle arrest and differentiation \cite{Tasseff2011}.
We explored this hypothesis by constructing a minimal model containing only signalsome and MAPK components (Fig. \ref{fig:network}A).
The effective model was developed using a novel framework which integrated logical rules with kinetic modeling.
This formulation significantly reduced the size of the model compared to the previous study of Tasseff et al., while maintaining similar model performance \cite{Tasseff2011}.
The effective model, despite its simplicity, captured key features of ATRA induced differentiation of HL-60 cells.
We identified an ensemble of effective model parameters using measurements taken from ATRA-induced HL-60 cells.
Using these parameters, model analysis predicted the bistability of MAPK activation as a function of ATRA exposure.
Conformational experiments supported ATRA-induced bistability.
These findings, combined with other literature evidence, suggests that positive feedback architectures are central to
many cell fate programs.

\clearpage

\section*{Results}

Interrogation of the Raf interactome suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
The architecture ultimately responsible for mediating ATRA-induced Raf activation is not clear.
To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
A panel of 19 possible Raf interaction partners (kinases, GTPases, scaffolding proteins etc)
was constructed based upon known signaling pathways.
We did not consider the most likely binding partner,
the small GTPase RAS, as previous studies have ruled it out in MAPK activation in HL-60 cells \cite{Wang2008,Katagiri1994}.
Total Raf was used as a bait protein for the immunoprecipitation studies.
Western blot analysis using total Raf and pS621 Raf specific antibodies confirmed the
presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf,
suggesting a direct physical interaction was possible.
However, only the associations between Raf and Vav1 and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
Furthermore, the Vav1 and Src associations were correlated with pS621 Raf abundance in the precipitate.
Others proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
The remaining 14 proteins were expressed in whole cell lysate (Fig. \ref{fig:ip-data}B),
but were not detectable in the precipitate of Raf IP.
Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
although the signal intensity for Src was notably weak.
However, GW5074 did not influence the association of CK2 or 14-3-3 with Raf, further demonstrating their independence from Raf phosphorylation.
Interestingly, the Raf-Akt interaction qualitatively increased following treatment with GW5074;
however, it remained unaffected by treatment with ATRA.
Src family kinases are known to be important in myeloid differentiation \cite{Miranda2007} and their role in HL-60 differentiation has been investigated elsewhere \cite{Congleton:2012fk}.
Given the existing work and variable reproducibility in the context of the Raf immunoprecipitate,
we will not investigate the role of Src further in this study.
Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with Raf activation following ATRA-treatment.
Previous studies demonstrated that a Vav1-Slp76-Cbl-CD38 complex plays an important role in ATRA-induced MAPK activation and differentiation of HL-60 cells \cite{Shen2009}.
Here we did not observe direct interaction of Raf with Cbl or Slp76;
however, this complex could be involved upstream.

%For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{fig:ip-data}A.

Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers following ATRA exposure (Fig. \ref{fig:ip-data}D-F).
Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
ATRA treatment alone statistically significantly increased the G1/G0 percentage over the untreated control,
while GW5074 alone had a negligible effect on the cell cycle distribution (Fig. \ref{fig:ip-data}D).
Surprisingly, the combination of GW5074 and ATRA statistically significantly increased the G1/G0 population (82 $\pm$ 1\%)
compared with ATRA alone (61 $\pm$ 0.5\%).
Increased G1/G0 arrest following the combined treatment with GW5074 and ATRA was unexpected,
as the combination of ATRA and the MEK inhibitor (PD98059) has been shown previously to decrease ATRA-induced growth arrest \cite{Yen1998}.
However, growth arrest is not the sole indication of functional differentiation.
Expression of the cell surface marker CD11b has also been shown to coincide with HL-60 cells myeloid differentiation \cite{Hickstein1989}.
We measured CD11b expression, for the various treatment groups, using immuno-fluorescence flow cytometry 48 hr post-treatment.
As with G1/G0 arrest, ATRA alone increased CD11b expression over the untreated control,
while GW5074 further enhanced ATRA-induced CD11b expression (Fig. \ref{fig:ip-data}E).
GW5074 alone had no statistically significant effect on CD11b expression, compared with the untreated control.
Lastly, the inducible reactive oxygen species (ROS) response was used as a functional marker of differentiated neutrophils \cite{Congleton2011}.
We measured the ROS response induced by the phorbol ester 12-O-tetradecanoylphorbol-13-acetate (TPA) using flow cytometry.
Untreated cells showed no discernible TPA response,
with only 7.0 $\pm$ 3.0\% ROS induction (Fig. \ref{fig:ip-data}F).
Cells treated with ATRA had a significantly increased TPA response, 53 $\pm$ 7\% ROS induction 48 hr post-treatment.
Treatment with both ATRA and GW5074 statistically significantly reduced ROS induction (22 $\pm$ 0.6\%) compared to ATRA alone.
Interestingly, Western blot analysis did not detect a GW5074 effect on ATRA-induced expression of p47phox,
a required upstream component of the ROS response  (Fig. \ref{fig:ip-data}F, bottom).
Thus, the inhibitory effect of GW5074 on inducible ROS might occur downstream of p47phox expression.
However, the ROS producing complex is MAPK dependent,
therefore it is also possible that GW5074 inhibited ROS production by interfering
with MAPK activation (in which case the p47Phox marker might not accurately reflect phenotypic conversion and differentiation).

We constructed an effective model of the ATRA-induced HL-60 differentiation circuit which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
The model connectivity was developed from literature and the studies presented here.
The signaling model recapitulated sustained signalsome/MAPK activation following exposure to 1$\mu$M ATRA (Fig. \ref{fig:model-fitting}A-B).
An ensemble of effective model parameters was estimated by minimizing the difference between simulations and time-series measurements of BLR1 mRNA and
cRAF-pS621 following the addition of 1$\mu$M ATRA.
We focused on the S621 phosphorylation site of cRAF since enhanced phosphorylation at this site is a defining characteristic of  sustained MAPK activation in HL-60.
The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A) and sustained phosphorylation of cRAF-pS621 (Fig. \ref{fig:model-fitting}B)
in a growing population of HL-60 cells. However, the effective model failed to capture the decline of BLR1 message after 48 hr of ATRA exposure.
Next, we tested the response of the signalsome/MAPK signaling model to different ATRA dosages.

% Secondly, ATRA treatment alone increased both Raf phosphorylation at S621/S259 and Raf expression 24 hr post-treatment,
% while ATRA-induced phosphorylation of ERK was blocked by GW5074, consistent with the loss of Raf kinase activity (Fig. \ref{fig:ip-data}B).
% Lastly, both the phosphorylation and expression of Raf was inhibited by GW5074.
% This suggested a positive feedback between Raf kinase activity, Raf phosphorylation and ERK action on Raf transcription factors.

The signalsome/MAPK signaling model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
Nullcline analysis predicted two stable steady-states and a single unstable state when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C).
In the lower stable state, neither the signalsome nor cRAF-pS621 were present (thus, the differentiation program was deactivated).
However, at the high stable state, both the signalsome and cRAF-pS621 were present, allowing for sustained activation and differentiation.
Interestingly, when ATRA was above a critical threshold, only the activated state was accessible (Fig. \ref{fig:model-fitting}D).
To test these findings, we first identified the ATRA threshold. We exposed HL-60 cells to different ATRA concentrations for 72 hr (Fig. \ref{fig:model-fitting}E).
Morphological changes associated with differentiation were visible for ATRA $\geq$ 0.25 $\mu$M, suggesting the critical ATRA
threshold was near this concentration. Next, we conducted washout ATRA washout experiments to determine if activated cells remained activated even in the absence of ATRA.
HL-60 cells locked into an activated state remained activated following ATRA withdraw (Fig. \ref{fig:model-predictions}).
Sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
Thus, following activation, if we inhibited or removed elements from the effective circuit we expected the siganlsome and MAPK signals to decay.
We simulated ATRA induced activation in the presence of kinase inhibitors, and without key circuit elements.
Consistent with experimental results using multiple MAPK inhibitors, ATRA activation in the presence of MAPK inhibitors lowered the steady-state value of signalsome (Fig. \ref{fig:model-predictions}A).
In the presence of BLR1, the signalsome and cRAF-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, blue).
On the other hand, BLR1 deletion removed the ability of the circuit to maintain a sustained MAPK response following the withdraw of ATRA (Fig. \ref{fig:model-predictions}B, gray).
Lastly, washout experiments in which cells were exposed to 1$\mu$M ATRA for 24 hr, and then transferred to fresh media without ATRA,
confirmed the persistence of the self sustaining activated state for up to 144 hr (Fig. \ref{fig:model-predictions}C).
Thus, these experiments and simulations confirmed that reinforcing positive feedback likely drives the ATRA-induced differentiation program.
Next, we analyzed the ATRA-induced downstream gene expression program following signalsome and cRaf activation.

The reduced order gene expression model described signal integration and ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signalsome-MAPK model produced two outputs, Trigger and cRAF-S621 which drove the downstream differentiation program.
In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex, regulated the expression of the transcription factors
CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1. In turn, these transcription factors, in combination with cRaf-S621, regulated
the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
We assembled the connectivity of the signal integration program driven by Trigger, and the phenotypic program from literature (supplemental materials).
We estimated the parameters of the signal integration and phenotype programs from previous studies which contained both steady-state and dynamic measurements
of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
The model simulations captured the time dependent expression of both CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A), and
steady-state values for upstream members of the signal integration unit (Fig. \ref{fig:model-grn-simulations}B).

% Vav1 is a member of an ATRA-inducible signalsome complex which propels sustained MAPK activation, arrest and differentiation (Fig. \ref{fig:network}B).
% We conducted immunoprecipitation studies and identified a limited number of ATRA-dependent and -independent RAF interaction partners.
% Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with cRAF,
% suggesting a direct physical interaction was possible.
% However, only the associations between cRAF and Vav1 and Src were ATRA-inducible (Fig. \ref{fig:network}B).
% Others proteins e.g., CK2, Akt and 14-3-3, generally bound cRAF regardless of phosphorylation status or ATRA treatment.
% Treatment with the RAF kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 and Src with cRAF (Fig. \ref{fig:network}C),
% although the signal intensity for Src was weak.
% However, GW5074 did not influence the association of CK2 or 14-3-3 with cRAF,
% further demonstrating their independence from cRAF phosphorylation.
% Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with cRAF activation following ATRA-treatment.

\clearpage

\section*{Discussion}
In this study, we presented an effective model of ATRA-inducible differentiation of HL-60 cells which
encoded positive feedback between the ATRA-inducible signalsome complex and the MAPK pathway.
Despite its simplicity, the model captured key features of the ATRA induced differentiation such as
sustained MAPK activation, and bistability with respect to ATRA exposure. We also reported a new ATRA-inducible component of the signalsome, Vav1.
Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
The Vav1/RAF association correlated with RAF activity, was ATRA-inducible and decreased after treatment with GW5074.
The presence of Vav1 in RAF/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
While its function in the signalsome is unclear, Vav1 has been shown to associate with a Cbl-Slp76-CD38 complex in an ATRA-dependent manner;
furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevented
ATRA-induced MAPK activation \cite{Shen2009}. Thus, interaction of Cbl-Slp76-Vav1 and CD38 appears to be required for transmission of the ATRA signal by the
signalsome.

We conducted immunoprecipitation studies and identified a limited number of ATRA-dependent and -independent Raf interaction partners.
While we were unable to detect the association of Raf with common kinases and GTPases such as PKC, PKA, p38, Rac and Rho,
we did establish potential interactions between Raf and key partners such as Vav1, Src, Akt, CK2 and 14-3-3.
All of these partners are known to be associated with Raf activation or function.
Src is known to bind Raf through an SH2 domain, and this association has been shown to be dependent of the serine phosphorylation of Raf \cite{Cleghon1994}.
Thus, an ATRA inducible Src/Raf association may be a result of ATRA-induced Raf phosphorylation at S259 or S621.
We also identified an interaction between Raf and the Ser/Thr kinases Akt and CK2.
Akt can phosphorylate Raf at S259, as demonstrated by studies in a human breast cancer line \cite{Zimmermann1999}.
CK2 can also phosphorylate Raf, although the literature has traditionally focused on S338 and not S621 or S259\cite{Ritt2007}.
However, neither of these kinase interactions were ATRA-inducible, suggesting their association with Raf alone was not associated with ATRA-induced Raf phosphorylation.
The adapter protein 14-3-3 was also constitutively associated with Raf.
The interaction between Raf and 14-3-3 has been associated with both S621 and S259 phosphorylation and activity \cite{Hekman2004}.
Additionally, the association of Raf with 14-3-3 not only stabilized S621 phosphorylation,
but also reversed the S621 phosphorylation from inhibitory to activating \cite{Dhillon2009}.
Finally, we found that Vav1/Raf association correlated with Raf activity, was ATRA-inducible and decreased after treatment with GW5074.
The presence of Vav1 in Raf/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
Interestingly, while an integrin ligand-induced ROS response was blocked in Vav1 knockout neutrophils, TPA was able to bypass the Vav1 requirement and stimulate both ERK phosphorylation
and ROS induction \cite{Graham2007}.
In this study, the TPA-induced ROS response was dependent upon Raf kinase activity, and was mitigated by the addition of GW5074.
It is possible that Vav1 is downstream of various integrin receptors but upstream of Raf in terms of inducible ROS responses.
Vav1 has also been shown to associate with a Cbl-Slp76-CD38 complex in an ATRA-dependent manner;
furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevents
ATRA-induced MAPK activation \cite{Shen2009}.
The literature suggest a variety of possible receptor-signaling pathways, which involve Vav1, for MAPK activation; moreover,
given the ATRA-inducible association Vav1 may play a direct role in Raf activation.

We hypothesized that Vav1 is a member of an ATRA-inducible complex which propels sustained MAPK activation, arrest and differentiation.
Initially, ATRA-induced Vav1 expression drives increased association between Vav1 and Raf.
This increased interaction facilitates phosphorylation and activation of Raf by pre-bound Akt and/or CK2 at S621 or perhaps S259.
Constitutively bound 14-3-3 may also stabilize the S621 phosphorylation, modulate the activity and/or up-regulate autophosphorylation.
Activated Raf can then drive ERK activation, which in turn closes the positive feedback loop by activating Raf transcription factors,
e.g. Sp1 and/or STAT1 \cite{Kim2009,Milanini-Mongiat2002,Zhang2004,Li2006}.
We tested this working hypothesis using mathematical modeling.
The model recapitulated both ATRA time-course data as well as the GW5074 inhibitor effects.
This suggested the proposed Raf-Vav1 architecture was at least consistent with the experimental studies.
Further, analysis of the Raf-Vav1 model identified bistability in ppERK levels.
Thus, two possible MAPK activation branches were possible for experimentally testable ATRA values. The analysis also suggested the ATRA-induced Raf-Vav1 architecture could be locked into a
sustained signaling mode (high ppERK) even in the absence of a ATRA signal.
This locked-in property could give rise to an ATRA-induction memory.
We validated the treatment memory property predicted by the Raf-Vav1 circuit experimentally using ATRA-washout experiments.
ERK phosphorylation levels remained high for more then 96 hr after ATRA was removed.
Previous studies demonstrated that HL-60 cells possessed an inheritable memory of ATRA stimulus \cite{YEN1984}.
Although the active state was self-sustaining, the inactive state demonstrated considerable robustness to perturbation.
For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on ppERK levels
over the entire ensemble.
CD38 expression correlated with the ppERK, suggesting its involvement in the signaling complex.
Our computational and experimental results showed that positive feedback,
through ERK-dependent Raf expression, could sustain MAPK signaling through many division cycles.
Such molecular mechanisms could underly aspects of cellular memory associated to consecutive ATRA treatments.

Several engineered, or naturally occurring systems involved in cell fate decisions incorporate positive feedback and bistability \cite{Ferrell:2002aa}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the differentiation circuit presented here; ATRA drives signalsome which activates MAPK, cell-cycle arrest, differentiation and signalsome.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, they share similar cell fate decision architectures.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while positive feedback is sometimes not desirable in man made systems, it may be at the core of a diverse variety of
cell fate programs and other networks important to human health.

Model performance was impressive given its limited size. However, there were several issues to explore further.
First, there was likely missing connectivity in the effective differentiation circuit.
Decreasing BLR1 expression with simultaneously sustained cRAF-pS261 activation was not captured by the current network architecture.
This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRAF-pS261 abundance.
We could model this by separating siganlsome formation into an inactive precursor pool that is transformed to a long-lived activated siganlsome by MAPK activation.
We should also explore adding additional downstream biological modules to this skeleton model, for example the upregulation of reactive oxygen markers such as p47Phox or
cell cycle arrest components to capture the switch from an actively proliferating population to a population in G0-arrest.
Next, the choice of max/min integration rules or the particular form of the transfer functions could also be explored.
Integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules might have different properties which could influence model identification or performance.
For example, a mean integration rule would be differentiable, allowing derivative-based optimization approaches to be used.
The form of the transfer function could also be explored. We choose hill-like functions because of their
prominence in the systems and synthetic biology community. However, many other transfer functions are possible.

\clearpage

\section*{Materials and Methods}

\subsubsection*{Effective ATRA differentiation model.}
ATRA induced signaling events were modeled using saturation kinetics within an ordinary differential equation (ODE) framework:
\begin{equation}
	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\left(\mu + k_{d}\right) x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}
\end{equation}
The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model, respectively.
The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of process $j$.
Typically, process $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
Lastly, $\mu$ denotes the specific growth rate, and $k_{d}$ denotes the rate constant controlling cell death.
Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.

Signaling rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the HL-60 model.
The rate of an enzyme catalyzed process was modeled using saturation kinetics:
\begin{equation}\label{eqn:rate-saturation}
	\bar{r}_{j} = k_{j}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\left(\frac{x_{s}}{K_{js} + x_{s}}\right)
\end{equation}
where $k_{j}$ denotes the catalytic rate constant for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
and $K_{js}$ denotes the saturation constant for species $s$ and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$.
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
If a process had no modifying factors, $v_{j}=1$.

% The model was constructed using the hybrid modeling approach of Wayman et al. \cite{pr3010138}.
% The abundance of species $i$ ($x_{i}$) is governed by:
% \begin{equation}
% 	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
% \end{equation}
% where $\mathcal{R}$ and $\mathcal{M}$ denotes the number of reactions and species in the model, $\mu$ denotes the specific growth rate and
% $k_{d,i}$ denotes a degradation constant for species $i$. The $\mu x_{i}$ term, which accounts for dilution due to cell growth, decays to zero following ATRA exposure.
% The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$,
% $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$, and $\mathbf{k}$ ($\mathcal{K}\times{1}$) denotes the unknown kinetic parameter vector.
% If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$,
% while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
% Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
% The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right) = \bar{r}_{j}v_{j}$.
% In this study, we used either zero- or first-order kinetics.
% The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
% For each rate, we used a rule-based approach to select from competing control factors.
% If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
% $v_{j} = \mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
% where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
% The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
% variable. Each regulatory transfer function took the form:
% \begin{equation}\label{eqn:control-factor}
% 	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right) = k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
% \end{equation}where $\mathcal{Z}_{i}$ denotes the abundance factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
% In this study, we used $\mathcal{I}_{j}\in\left\{\max,\min\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j} = 1$.

The HL-60 model described both signal transduction and gene expression events following the addition of ATRA.
The output of the signal transduction model was the input to the gene expression model.
For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$):
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
\end{eqnarray}
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
The specific transcription rate was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic rate of transcription was modeled as:
\begin{equation}
	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$. The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths,
where $L_{T,o}$ is a characteristic gene length, and $L_{T,j}$ denotes the length of gene $j$.
Thus, the $\left(L_{T,o}/L_{T,j}\right)$ term is gene specific correction to the characteristic transcription rate.
The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.

The gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$,
and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule.
In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations resulting in gene expression.
Thus, the control variable $u_{j}$ took the form:
\begin{equation}
	{u}_{j} = \frac{W_{R_{1},j}+\displaystyle\sum_{n}W_{nj}f_{nj}}{1+W_{R_{1},j}+\displaystyle\sum_{d}W_{dj}f_{dj}}
\end{equation}where the numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using hill like functions), while
the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
If a gene expression process had no modifying factors, $u_{j}=1$.
Lastly, the specific translation rate was modeled as:
\begin{equation}
	r_{X,j} = V^{max}_{X}\left(\frac{m_{j}}{K_{X}+m_{j}}\right)
\end{equation}
where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, and $K_{X}$ denotes a translation saturation constant.
The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
and the Ribosome abundance ($R_{2}$), $V^{max}_{X} = k_{X}\left(R_{2}\right)$.

In this study, we estimated the $W_{ij}$ parameters, and the parameters in the transfer functions $f_{dj}$ from gene expression data sets.
On the other hand, we estimated $k_{T}, k_{X}, \theta_{m,j}, \theta_{p,j}, R_{1}$ and $R_{2}$ using estimates of transcription and translation rates,
the half-life of a typical mRNA and protein, and a typical value for the copies per cell of RNA polymerase and ribosomes from literature \cite{Milo:2010aa}.
The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
The values used for the characteristic transcription/translation parameters, degradation constants and
macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.

\noindent\subsubsection*{Estimation of signaling and gene expression model parameters.}
Signal and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental data set $j$:
\begin{equation}\label{eqn:objective-function}
	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
\end{equation}
The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs at time $i$ from training set $j$,
where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
The first term in Eqn. \eqref{eqn:objective-function} quantified the relative simulation error.
We used immunoblot intensity measurements for model training.
Thus, we trained the model on the \emph{relative} change between bands within each data set.
Suppose we have the intensity of species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$.
The scaled value $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ is given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
\end{equation}
where $\hat{\mathcal{M}}_{ij}=0$ and $\hat{\mathcal{M}}_{ij}=1$  describe the lowest (highest) intensity bands.
A similar scaling was used for the simulation output. The second term in the objective function ensured a realistic concentration scale was estimated by the model.
We set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
We minimized the total model residual $\sum_{j}E_{j}$ using heuristic optimization starting from a random initial parameter guess.

The signaling and gene expression model equations were implemented in Julia and solved using the \texttt{CVODE} routine of the Sundials package \cite{Julia,Hindmarsh2005}.
The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.

% All code was implemented in the Octave programming language, and is available for
% download from http://www.varnerlab.org.

%Model parameters were estimated by minimizing the difference between simulations and experimental measurements (squared residual):
%\begin{equation}\label{eqn:objective-function}
%	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
%\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
%value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.

\subsubsection*{Cell culture and treatment}
Human myeloblastic leukemia cells (HL-60 cells) were grown in a humidified atmosphere of 5\% CO$_2$ at 37$^{o}$C and maintained in RPMI 1640 from Gibco (Carlsbad, CA)
supplemented with 5\% heat inactivated fetal bovine serum from Hyclone (Logan, UT) and 1$\times$ antibiotic/antimicotic (Gibco, Carlsbad, CA).
Cells were cultured in constant exponential growth \cite{Brooks1996}.
Experimental cultures were initiated at $0.1\times10^6$ cells/mL 24 hr prior to ATRA treatment;
if indicated, cells were also treated with GW5074 (2$\mu$M) 18 hr before ATRA treatment.
For the cell culture washout experiments, cells were treated with ATRA for 24 hr,
washed 3x with prewarmed serum supplemented culture medium to remove ATRA,
and reseeded in ATRA-free media as described.
Western blot analysis was performed at incremental time points after removal of ATRA.

\subsubsection*{Chemicals}
All-Trans Retinoic Acid (ATRA) from Sigma-Aldrich (St. Louis, MO) was dissolved in
100\% ethanol with a stock concentration of 5mM, and used at a final concentration of 1$\mu$M (unless otherwise noted).
The cRAF inhibitor GW5074 from Sigma-Aldrich (St. Louis, MO) was dissolved in DMSO with a stock concentration of 10mM,
and used at a final concentration of 2$\mu$M.
HL-60 cells were treated with 2$\mu$M GW5074 with or without ATRA (1$\mu$M) at 0 hr.
This GW5074 dosage had a negligible effect on the cell cycle distribution, compared to ATRA treatment alone.

\subsubsection*{Immunoprecipitation and western blotting}
Approximately $1.2 \times 10^7$ cells were lysed using 400$\mu$L of M-Per lysis buffer from Thermo Scientific (Waltham, MA).
Lysates were cleared by centrifugation at 16,950 $\times$ g in a micro-centrifuge for 20 min at 4$^{o}$C.
Lysates were pre-cleared using 100$\mu$L protein A/G Plus agarose beads from Santa Cruz Biotechnology (Santa Cruz, CA) by
inverting overnight at 4$^{o}$C.  Beads were cleared by centrifugation and total protein concentration
was determined by a BCA assay (Thermo Scientific, Waltham, MA).  Immunoprecipitations were setup by bringing
lysate to a concentration of 1g/L in a total volume of 300$\mu$L (M-Per buffer was used for dilution).
The anti-RAF antibody was added at 3$\mu$L.
A negative control with no bait protein was also used to exclude the direct interaction of proteins with the A/G beads.
After 1 hr of inversion at 4$^{o}$C, 20$\mu$L of agarose beads was added and samples were left
to invert overnight at 4$^{o}$C. Samples were then washed three times with M-Per buffer by centrifugation.
Finally proteins were eluted from agarose beads using a laemmli loading buffer.
Eluted proteins were resolved by SDS-PAGE and Western blotting.  Total
lysate samples were normalized by total protein concentration (20$\mu$g per sample)
and resolved by SDS-PAGE and Western blotting.  Secondary HRP bound antibody was used for visualization.
All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of $\alpha$-p621 RAF which was purchased from Biosource/Invitrogen (Carlsbad, CA),
and $\alpha$-CK2 from BD Biosciences (San Jose, CA).

\subsubsection*{Morphology assessment}
Untreated and ATRA-treated HL-60 cells were collected after 72 hr and cytocentrifuged for 3 min at 700 rpm onto glass slides.
Slides were air-dried and stained with Wright’s stain. Slide images were captured at 40X (Leica DM LB 100T microscope, Leica Microsystems).

\clearpage

\section*{Competing interests}
The authors declare that they have no competing interests.

\section*{Author's contributions}
J.V and A.Y directed the study. R.T, H.J and J.C conducted the cell culture measurements.
J.V and W.D developed the reduced order HL-60 models and the parameter ensemble.
W.D analyzed the model ensemble, and generated figures for the manuscript.
The manuscript was prepared and edited for publication by W.D, A.Y and J.V.

\section*{Acknowledgements}
We gratefully acknowledge the suggestions from the anonymous reviewers to improve this manuscript.

\section*{Funding}
We acknowledge the financial support to J.V. by the National Science Foundation CAREER (CBET-0846876) for the support of R.T. and H.J.
In addition, we acknowledge support to A.Y. from the National Institutes of Health (CA 30555, CA152870)
and a grant from New York State Stem Cell Science.
Lastly, we acknowledge the financial support to J.V. and A.Y. from the National Cancer Institute (\#U54~CA143876).
The content is solely the responsibility of the authors and does not necessarily
represent the official views of the National Cancer Institute or the National Institutes of Health.

\clearpage

% References -
\bibliography{Paper_v1}

\clearpage

\begin{figure}[!t]
\includegraphics[width=0.9\textwidth]{./figs/Fig-1-IP-ExperimentalStudies.pdf}
\caption{Investigation of a panel of possible Raf interaction partners in the presence and absence of ATRA.
A: Species identified to precipitate out with Raf:
first column shows Western blot analysis on total Raf immunoprecipitation with and without 24 hr ATRA treatment
and the second on total lysate.
B: The expression of species considered that did not precipitate out with Raf at levels detectable by Western blot analysis on total lysate.
C: Effect of the Raf inhibitor GW5074 on Raf interactions as determined by Western blot analysis of total Raf immunoprecipitation.
The Authors note the the signal associated with Src was found to be weak.
D: Cell Cycle distribution as determined by flow cytometry indicated arrest induced by ATRA, which was increased by the addition of GW5074.
E: Expression of the cell surface marker CD11b as determined by flow cytometry indicated increased expression induced by ATRA,
which was enhanced by the addition of GW5074.
F: Inducible reactive oxygen species (ROS) as determined by DCF flow cytometry. The functional differentiation response of ATRA treated cells
was mitigated by GW5074.}\label{fig:ip-data}
\end{figure}

\begin{figure}[!t]
\includegraphics[width=1.0\textwidth]{./figs/Fig-1-Network_v3.pdf}
\caption{Schematic of the effective ATRA differentiation circuit.
Above a critical threshold, ATRA activates an upstream Trigger, which induces signalsome complex formation.
Signalsome activates the mitogen-activated protein kinase (MAPK) cascade which in turn
drives the differentiation program and signalsome formation.
Both Trigger and activated cRAF-S621 drive a phenotype gene expression program responsible for differentiation.
Trigger activates the expression of a series of transcription factors which in combination with cRAF-S621 result in phenotypic change.}\label{fig:network}
\end{figure}

\begin{figure}[!t]\centering
\includegraphics[width=0.90\textwidth]{./figs/Fig-2-cRAF-BLR1-Fit-Analysis.pdf}
\caption{Model analysis for ATRA-induced HL-60 differentiation.
A: BLR1 mRNA versus time following exposure to 1$\mu$M ATRA at t = 0 hr.
B: cRAF-pS621 versus time following exposure to 1$\mu$M ATRA at t = 0 hr.
Points denote experimental measurements, solid lines denote the mean model performance. Shaded regions denote the 99\% confidence interval calculated over the parameter ensemble.
C: Signalsome and cRAF-pS621 nullclines for ATRA below the critical threshold.
The model had two stable steady states and a single unstable state in this regime.
D: Signalsome and cRAF-pS621 nullclines for ATRA above the critical threshold.
In this regime the model had only a single stable steady state.
E: Morphology of HL-60 as a function of ATRA concentration (t = 72 hr). }\label{fig:model-fitting}
\end{figure}

\begin{figure}[!t]\centering
\includegraphics[width=1.0\textwidth]{./figs/Fig-3-Predictions.pdf}
\caption{Model simulation following exposure to 1$\mu$M ATRA.
A: BLR1 mRNA versus time with and without MAPK inhibitor.
B: cRAF-pS621 versus time following pulsed exposure to 1$\mu$M ATRA with and without BLR1.
Solid lines denote the mean model performance, while shaded regions denote the 99\% confidence interval calculated over the parameter ensemble.
C: Western blot analysis of phosphorylated ERK1/2 in ATRA washout experiments.
Experimental data in panels A and B were reproduced from Wang and Yen \cite{Wang2008}, data in panel C is reported in this study. }\label{fig:model-predictions}
\end{figure}

\begin{figure}[!t]\centering
\includegraphics[width=0.85\textwidth]{./figs/Fig-5-GRN-Simulations.pdf}
\caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
A: CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
B: Gene expression at t = 48 hr following ATRA exposure.
Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-grn-simulations}
\end{figure}

\clearpage

% Supplemental figures -
% Set the S-
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{T\arabic{table}}
\renewcommand\thepage{S-\arabic{page}}
\renewcommand\theequation{S\arabic{equation}}

% Reset the counters -
\setcounter{equation}{0}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{page}{1}

\end{document}

### 2017-01-16 tmp.tex
What the what

PSO uses a \textit{swarming} metaheuristic to explore parameter spaces.
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$.

### 2017-01-19 Paper_v2.tex
Working on the introduction

@@
 \vspace{0.05in}
-{\par \centering \large{Ryan Tasseff, Holly A. Jensen, Johanna Congleton$^{\dag}$, Wei Dai, Andrew Yen$^{\dag}$, and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Ryan Tasseff, Holly A. Jensen, Wei Dai, Rodica P. Bunaciu$^{\dag}$, Johanna Congleton$^{\dag}$, Andrew Yen$^{\dag}$, and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 
+\setcounter{page}{1}
+
 \linenumbers
@@
 \section*{Introduction}
-Understanding differentiation programs is an important
-therapeutic challenge. Toward this challenge, lessons learned
-in model systems, such as the lineage-uncommitted human
-myeloblastic cell line HL-60, informs our analysis of more
-complex therapeutically important programs. Patient derived HL-60 leukemia cells have been
-a durable experimental model since the 1970’s \cite{Breitman1980}.
-HL-60 undergoes cell cycle arrest and either myeloid or monocytic differentiation following stimulation;
-All-Trans Retinoic Acid (ATRA) induces G1/G0-arrest and myeloid differentiation in HL-60 cells,
+Understanding the architecture of differentiation programs is an important therapeutic challenge.
+Differentiation induction chemotherapy (DIC), using agents such as the vitamin A derivative all-trans retinoic acid (ATRA),
+is a promising approach for the treatment of many cancers \cite{Bushue:2010aa,Tang:2011aa,Cheung:2012aa}.
+For example, ATRA treatment induces remission in 80–90\% of promyelocytic leukemia (APL) PML-RAR$\alpha$-positive patients \cite{Nilsson:1984aa}, thereby transforming a fatal diagnosis into a manageable disease.
+However, remission is sometimes not durable and relapsed cases exhibit emergent ATRA resistance \cite{Warrell:1993aa,Freemantle:2003aa}.
+To understand the basis of this resistance, we must first understand the ATRA-induced differentiation program.
+Toward this challenge, lessons learned in model systems, such as the lineage-uncommitted human myeloblastic cell line HL-60, could inform our analysis of the more
+complex differentiation programs occurring in patients. Patient derived HL-60 leukemia cells have been a durable experimental model since the 1970’s to study differentiation \cite{Breitman1980}.
+HL-60 undergoes cell cycle arrest and either myeloid or monocytic differentiation following stimulation; ATRA induces G1/G0-arrest and myeloid differentiation in HL-60 cells,
 while 1,25-dihydroxy vitamin D3 (D3) induces arrest and monocytic differentiation.
@@
 Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
-ATRA drives sustained MEK-dependent activation of the RAF/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
-MEK inhibition results in the loss of ERK and RAF phosphorylation, and the failure to arrest and differentiate \cite{Hong2001}.
+ATRA drives sustained MEK-dependent activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
+MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate \cite{Hong2001}.
 ATRA (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
-RAR/RXR activation is necessary for ATRA-induced RAF phosphorylation  \cite{Hong2001},
-and the formation of the ATRA-inducible signalsome complex which drives differentiation.
-The signalsome is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, as well as IRF-1 transcription factors \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
-This signaling is driven by ATRA-induced expression of CD38 and the putative heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
+RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation  \cite{Hong2001},
+and the formation of an ATRA-inducible signalsome complex at the membrane which drives MAPK activation through a yet to be identified kinase activity.
+While the makeup of the signalsome complex is not yet known,
+we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, as well as IRF-1 transcription factors \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
+Signalsome formation and activity is driven by ATRA-induced expression of CD38 and the putative heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
 BLR1, identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
-is necessary for MAPK activation and differentiation \cite{WANG2004}.
+is necessary for MAPK activation and differentiation \cite{WANG2004}, and is also involved with signalsome activity.
 Studies of the BLR1 promoter identified a 5' 17bp GT box approximately 1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
@@
 are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
-BLR1 overexpression enhanced RAF phosphorylation and accelerated terminal differentiation, while RAF inhibition reduced BLR1 expression and differentiation \cite{Wang2008}.
-BLR1 knock-out cells failed to activate RAF or differentiate in the presence of ATRA \cite{Wang2008}.
+BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and differentiation \cite{Wang2008}.
+BLR1 knock-out cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
 Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
-A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient to explain ATRA-induced sustained MAPK activation and the expression of differentiation markers \cite{Tasseff2011}.
+Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the siganlsome.
+A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
+to explain ATRA-induced sustained MAPK activation, and the expression of a limited number of functional differentiation markers \cite{Tasseff2011}.
 Model analysis also suggested that Raf was the most distinct of the MAPK proteins.
+However, this previous study developed and analyzed a complex model,
+thus leaving open the critical question of what is the minimal positive feedback circuit required to drive ATRA-induced differentiation.
 
-A critical question is what other components of the MAPK positive feedback circuit are required to drive ATRA-induced functional differentiation of HL-60 cells.
-Wang and Yen showed that ectopic expression of the constitutively active CR3 domain of Raf1 restored ATRA-induced G0 arrest and differentiation in BLR1 knock-out cells \cite{Wang2008}.
-However, ectopic expression of Raf1 CR3 alone, in the absence of ATRA, failed to induce arrest or differentiation.
-Thus, additional ATRA-inducible components must exist,
-which independently promote arrest and differentiation in the absence of BLR1.
-In this study, we developed a mathematical model of the key architectural feature of ATRA induced differentiation of HL-60 cells,
+In this study, we explored this question using a minimal mathematical model of the key architectural feature of ATRA induced differentiation of HL-60 cells,
 namely positive feedback between an ATRA-inducible signalsome complex and MAPK activation.
-Previously Tasseff et al. hypothesized that signalsome-MAPK positive feedback was essential for ATRA-induced
-cell cycle arrest and differentiation \cite{Tasseff2011}.
-We explored this hypothesis by constructing a minimal model containing only signalsome and MAPK components (Fig. \ref{fig:network}A).
-The effective model was developed using a novel framework which integrated logical rules with kinetic modeling.
-This formulation significantly reduced the size of the model compared to the previous study of Tasseff et al., while maintaining similar model performance \cite{Tasseff2011}.
+The ATRA responsive signalsome-MAPK circuit was then used to drive a downstream gene expression program which encoded for the expression of functional differentiation markers.
+The effective model used a novel framework which integrated logical rules with kinetic modeling to describe gene expression and protein regulation,
+while largely relying upon biophysical parameters from the literature.
+This formulation significantly reduced the size and complexity of the model compared to the previous study of Tasseff et al., while increasing the breadth of the biology described \cite{Tasseff2011}.
 The effective model, despite its simplicity, captured key features of ATRA induced differentiation of HL-60 cells.
-We identified an ensemble of effective model parameters using measurements taken from ATRA-induced HL-60 cells.
-Using these parameters, model analysis predicted the bistability of MAPK activation as a function of ATRA exposure.
-Conformational experiments supported ATRA-induced bistability.
-These findings, combined with other literature evidence, suggests that positive feedback architectures are central to
-many cell fate programs.
+Model analysis predicted the bistability of MAPK activation as a function of ATRA exposure; conformational experiments supported ATRA-induced bistability.
+Model simulations were also consistent with measurements of the influence of MAPK inhibitors, and the failure of BLR1 knockout cells to differentiate when exposed to ATRA.
+Lastly, we showed by through immunoprecipitation studies, that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the siganlsome complex.
+Taken together, these findings when combined with other literature evidence,
+suggested that positive feedback architectures are central to differentiation programs generally, and necessary for ATRA-induced differentiation.
+
+% These findings, combined with other literature evidence, suggests that positive feedback architectures are central to
+% many cell fate programs.
+%Wang and Yen showed that ectopic expression of the constitutively active CR3 domain of Raf1 restored ATRA-induced G0 arrest and differentiation in BLR1 knock-out cells \cite{Wang2008}.
+%However, ectopic expression of Raf1 CR3 alone, in the absence of ATRA, failed to induce arrest or differentiation in BLR1 knock-out cells.
+%We identified an ensemble of effective model parameters using measurements taken from ATRA-induced HL-60 cells.
+%Using these parameters,
 
@@
 presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
-Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf,
-suggesting a direct physical interaction was possible.
+Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf, suggesting a direct physical interaction was possible.
 However, only the associations between Raf and Vav1 and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
@@
 An ensemble of effective model parameters was estimated by minimizing the difference between simulations and time-series measurements of BLR1 mRNA and
-cRAF-pS621 following the addition of 1$\mu$M ATRA.
-We focused on the S621 phosphorylation site of cRAF since enhanced phosphorylation at this site is a defining characteristic of  sustained MAPK activation in HL-60.
-The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A) and sustained phosphorylation of cRAF-pS621 (Fig. \ref{fig:model-fitting}B)
+cRaf-pS621 following the addition of 1$\mu$M ATRA.
+We focused on the S621 phosphorylation site of cRaf since enhanced phosphorylation at this site is a defining characteristic of  sustained MAPK activation in HL-60.
+The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A) and sustained phosphorylation of cRaf-pS621 (Fig. \ref{fig:model-fitting}B)
 in a growing population of HL-60 cells. However, the effective model failed to capture the decline of BLR1 message after 48 hr of ATRA exposure.
@@
 Nullcline analysis predicted two stable steady-states and a single unstable state when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C).
-In the lower stable state, neither the signalsome nor cRAF-pS621 were present (thus, the differentiation program was deactivated).
-However, at the high stable state, both the signalsome and cRAF-pS621 were present, allowing for sustained activation and differentiation.
+In the lower stable state, neither the signalsome nor cRaf-pS621 were present (thus, the differentiation program was deactivated).
+However, at the high stable state, both the signalsome and cRaf-pS621 were present, allowing for sustained activation and differentiation.
 Interestingly, when ATRA was above a critical threshold, only the activated state was accessible (Fig. \ref{fig:model-fitting}D).
@@
 \ref{fig:model-predictions}A).
-In the presence of BLR1, the signalsome and cRAF-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, blue).
+In the presence of BLR1, the signalsome and cRaf-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, blue).
 On the other hand, BLR1 deletion removed the ability of the circuit to maintain a sustained MAPK response following the withdraw of ATRA (Fig. \ref{fig:model-predictions}B, gray).
@@
 The reduced order gene expression model described signal integration and ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
-The signalsome-MAPK model produced two outputs, Trigger and cRAF-S621 which drove the downstream differentiation program.
+The signalsome-MAPK model produced two outputs, Trigger and cRaf-S621 which drove the downstream differentiation program.
 In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex, regulated the expression of the transcription factors
@@
 % Vav1 is a member of an ATRA-inducible signalsome complex which propels sustained MAPK activation, arrest and differentiation (Fig. \ref{fig:network}B).
-% We conducted immunoprecipitation studies and identified a limited number of ATRA-dependent and -independent RAF interaction partners.
-% Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with cRAF,
+% We conducted immunoprecipitation studies and identified a limited number of ATRA-dependent and -independent Raf interaction partners.
+% Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with cRaf,
 % suggesting a direct physical interaction was possible.
-% However, only the associations between cRAF and Vav1 and Src were ATRA-inducible (Fig. \ref{fig:network}B).
-% Others proteins e.g., CK2, Akt and 14-3-3, generally bound cRAF regardless of phosphorylation status or ATRA treatment.
-% Treatment with the RAF kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 and Src with cRAF (Fig. \ref{fig:network}C),
+% However, only the associations between cRaf and Vav1 and Src were ATRA-inducible (Fig. \ref{fig:network}B).
+% Others proteins e.g., CK2, Akt and 14-3-3, generally bound cRaf regardless of phosphorylation status or ATRA treatment.
+% Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 and Src with cRaf (Fig. \ref{fig:network}C),
 % although the signal intensity for Src was weak.
-% However, GW5074 did not influence the association of CK2 or 14-3-3 with cRAF,
-% further demonstrating their independence from cRAF phosphorylation.
-% Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with cRAF activation following ATRA-treatment.
+% However, GW5074 did not influence the association of CK2 or 14-3-3 with cRaf,
+% further demonstrating their independence from cRaf phosphorylation.
+% Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with cRaf activation following ATRA-treatment.
 
@@
 Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
-The Vav1/RAF association correlated with RAF activity, was ATRA-inducible and decreased after treatment with GW5074.
-The presence of Vav1 in RAF/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
+The Vav1/Raf association correlated with Raf activity, was ATRA-inducible and decreased after treatment with GW5074.
+The presence of Vav1 in Raf/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
 Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
@@
 First, there was likely missing connectivity in the effective differentiation circuit.
-Decreasing BLR1 expression with simultaneously sustained cRAF-pS261 activation was not captured by the current network architecture.
-This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRAF-pS261 abundance.
+Decreasing BLR1 expression with simultaneously sustained cRaf-pS261 activation was not captured by the current network architecture.
+This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRaf-pS261 abundance.
 We could model this by separating siganlsome formation into an inactive precursor pool that is transformed to a long-lived activated siganlsome by MAPK activation.
@@
 100\% ethanol with a stock concentration of 5mM, and used at a final concentration of 1$\mu$M (unless otherwise noted).
-The cRAF inhibitor GW5074 from Sigma-Aldrich (St. Louis, MO) was dissolved in DMSO with a stock concentration of 10mM,
+The cRaf inhibitor GW5074 from Sigma-Aldrich (St. Louis, MO) was dissolved in DMSO with a stock concentration of 10mM,
 and used at a final concentration of 2$\mu$M.
@@
 lysate to a concentration of 1g/L in a total volume of 300$\mu$L (M-Per buffer was used for dilution).
-The anti-RAF antibody was added at 3$\mu$L.
+The anti-Raf antibody was added at 3$\mu$L.
 A negative control with no bait protein was also used to exclude the direct interaction of proteins with the A/G beads.
@@
 and resolved by SDS-PAGE and Western blotting.  Secondary HRP bound antibody was used for visualization.
-All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of $\alpha$-p621 RAF which was purchased from Biosource/Invitrogen (Carlsbad, CA),
+All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of $\alpha$-p621 Raf which was purchased from Biosource/Invitrogen (Carlsbad, CA),
 and $\alpha$-CK2 from BD Biosciences (San Jose, CA).
@@
 drives the differentiation program and signalsome formation.
-Both Trigger and activated cRAF-S621 drive a phenotype gene expression program responsible for differentiation.
-Trigger activates the expression of a series of transcription factors which in combination with cRAF-S621 result in phenotypic change.}\label{fig:network}
+Both Trigger and activated cRaf-S621 drive a phenotype gene expression program responsible for differentiation.
+Trigger activates the expression of a series of transcription factors which in combination with cRaf-S621 result in phenotypic change.}\label{fig:network}
 \end{figure}
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=0.90\textwidth]{./figs/Fig-2-cRAF-BLR1-Fit-Analysis.pdf}
+\includegraphics[width=0.90\textwidth]{./figs/Fig-2-cRaf-BLR1-Fit-Analysis.pdf}
 \caption{Model analysis for ATRA-induced HL-60 differentiation.
 A: BLR1 mRNA versus time following exposure to 1$\mu$M ATRA at t = 0 hr.
-B: cRAF-pS621 versus time following exposure to 1$\mu$M ATRA at t = 0 hr.
+B: cRaf-pS621 versus time following exposure to 1$\mu$M ATRA at t = 0 hr.
 Points denote experimental measurements, solid lines denote the mean model performance. Shaded regions denote the 99\% confidence interval calculated over the parameter ensemble.
-C: Signalsome and cRAF-pS621 nullclines for ATRA below the critical threshold.
+C: Signalsome and cRaf-pS621 nullclines for ATRA below the critical threshold.
 The model had two stable steady states and a single unstable state in this regime.
-D: Signalsome and cRAF-pS621 nullclines for ATRA above the critical threshold.
+D: Signalsome and cRaf-pS621 nullclines for ATRA above the critical threshold.
 In this regime the model had only a single stable steady state.
@@
 A: BLR1 mRNA versus time with and without MAPK inhibitor.
-B: cRAF-pS621 versus time following pulsed exposure to 1$\mu$M ATRA with and without BLR1.
+B: cRaf-pS621 versus time following pulsed exposure to 1$\mu$M ATRA with and without BLR1.
 Solid lines denote the mean model performance, while shaded regions denote the 99\% confidence interval calculated over the parameter ensemble.

### 2017-01-19 Paper_v2.tex
Working on results

@@
 \section*{Results}
+We constructed an effective model of the ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
+The model connectivity was developed from literature and the studies presented here.
+We decomposed the ATRA program into three modules;
+a signal initiation module that was responsible for sensing and transforming the ATRA signal into activated cRaf-S621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
+a signal integration module that controlled the expression of upstream transcription factors given cRaf-S621 and activated Trigger (Fig. \ref{fig:network}B); and
+a phenotype module which encoded the expression of functional differentiation markers (Fig. \ref{fig:network}C).
 
-Interrogation of the Raf interactome suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
-The architecture ultimately responsible for mediating ATRA-induced Raf activation is not clear.
-To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
-A panel of 19 possible Raf interaction partners (kinases, GTPases, scaffolding proteins etc)
-was constructed based upon known signaling pathways.
-We did not consider the most likely binding partner,
-the small GTPase RAS, as previous studies have ruled it out in MAPK activation in HL-60 cells \cite{Wang2008,Katagiri1994}.
-Total Raf was used as a bait protein for the immunoprecipitation studies.
-Western blot analysis using total Raf and pS621 Raf specific antibodies confirmed the
-presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
-Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf, suggesting a direct physical interaction was possible.
-However, only the associations between Raf and Vav1 and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
-Furthermore, the Vav1 and Src associations were correlated with pS621 Raf abundance in the precipitate.
-Others proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
-The remaining 14 proteins were expressed in whole cell lysate (Fig. \ref{fig:ip-data}B),
-but were not detectable in the precipitate of Raf IP.
-Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
-although the signal intensity for Src was notably weak.
-However, GW5074 did not influence the association of CK2 or 14-3-3 with Raf, further demonstrating their independence from Raf phosphorylation.
-Interestingly, the Raf-Akt interaction qualitatively increased following treatment with GW5074;
-however, it remained unaffected by treatment with ATRA.
-Src family kinases are known to be important in myeloid differentiation \cite{Miranda2007} and their role in HL-60 differentiation has been investigated elsewhere \cite{Congleton:2012fk}.
-Given the existing work and variable reproducibility in the context of the Raf immunoprecipitate,
-we will not investigate the role of Src further in this study.
-Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with Raf activation following ATRA-treatment.
-Previous studies demonstrated that a Vav1-Slp76-Cbl-CD38 complex plays an important role in ATRA-induced MAPK activation and differentiation of HL-60 cells \cite{Shen2009}.
-Here we did not observe direct interaction of Raf with Cbl or Slp76;
-however, this complex could be involved upstream.
-
-%For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{fig:ip-data}A.
-
-Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers following ATRA exposure (Fig. \ref{fig:ip-data}D-F).
-Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
-ATRA treatment alone statistically significantly increased the G1/G0 percentage over the untreated control,
-while GW5074 alone had a negligible effect on the cell cycle distribution (Fig. \ref{fig:ip-data}D).
-Surprisingly, the combination of GW5074 and ATRA statistically significantly increased the G1/G0 population (82 $\pm$ 1\%)
-compared with ATRA alone (61 $\pm$ 0.5\%).
-Increased G1/G0 arrest following the combined treatment with GW5074 and ATRA was unexpected,
-as the combination of ATRA and the MEK inhibitor (PD98059) has been shown previously to decrease ATRA-induced growth arrest \cite{Yen1998}.
-However, growth arrest is not the sole indication of functional differentiation.
-Expression of the cell surface marker CD11b has also been shown to coincide with HL-60 cells myeloid differentiation \cite{Hickstein1989}.
-We measured CD11b expression, for the various treatment groups, using immuno-fluorescence flow cytometry 48 hr post-treatment.
-As with G1/G0 arrest, ATRA alone increased CD11b expression over the untreated control,
-while GW5074 further enhanced ATRA-induced CD11b expression (Fig. \ref{fig:ip-data}E).
-GW5074 alone had no statistically significant effect on CD11b expression, compared with the untreated control.
-Lastly, the inducible reactive oxygen species (ROS) response was used as a functional marker of differentiated neutrophils \cite{Congleton2011}.
-We measured the ROS response induced by the phorbol ester 12-O-tetradecanoylphorbol-13-acetate (TPA) using flow cytometry.
-Untreated cells showed no discernible TPA response,
-with only 7.0 $\pm$ 3.0\% ROS induction (Fig. \ref{fig:ip-data}F).
-Cells treated with ATRA had a significantly increased TPA response, 53 $\pm$ 7\% ROS induction 48 hr post-treatment.
-Treatment with both ATRA and GW5074 statistically significantly reduced ROS induction (22 $\pm$ 0.6\%) compared to ATRA alone.
-Interestingly, Western blot analysis did not detect a GW5074 effect on ATRA-induced expression of p47phox,
-a required upstream component of the ROS response  (Fig. \ref{fig:ip-data}F, bottom).
-Thus, the inhibitory effect of GW5074 on inducible ROS might occur downstream of p47phox expression.
-However, the ROS producing complex is MAPK dependent,
-therefore it is also possible that GW5074 inhibited ROS production by interfering
-with MAPK activation (in which case the p47Phox marker might not accurately reflect phenotypic conversion and differentiation).
-
-We constructed an effective model of the ATRA-induced HL-60 differentiation circuit which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
-The model connectivity was developed from literature and the studies presented here.
-The signaling model recapitulated sustained signalsome/MAPK activation following exposure to 1$\mu$M ATRA (Fig. \ref{fig:model-fitting}A-B).
+The signal initiation module recapitulated sustained signalsome/MAPK activation following exposure to 1$\mu$M ATRA (Fig. \ref{fig:model-fitting}A-B).
 An ensemble of effective model parameters was estimated by minimizing the difference between simulations and time-series measurements of BLR1 mRNA and
-cRaf-pS621 following the addition of 1$\mu$M ATRA.
-We focused on the S621 phosphorylation site of cRaf since enhanced phosphorylation at this site is a defining characteristic of  sustained MAPK activation in HL-60.
-The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A) and sustained phosphorylation of cRaf-pS621 (Fig. \ref{fig:model-fitting}B)
+cRaf-S621 following the addition of 1$\mu$M ATRA. We focused on the S621 phosphorylation site of cRaf since enhanced phosphorylation at this site is a defining
+characteristic of sustained MAPK activation in HL-60. The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A)
+and sustained phosphorylation of cRaf-pS621 (Fig. \ref{fig:model-fitting}B)
 in a growing population of HL-60 cells. However, the effective model failed to capture the decline of BLR1 message after 48 hr of ATRA exposure.
-Next, we tested the response of the signalsome/MAPK signaling model to different ATRA dosages.
+Next, we tested the response of the signalsome/MAPK signaling module to different ATRA dosages.
 
@@
 % Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with cRaf activation following ATRA-treatment.
+
+The composition of the siganlsome, and the kinase ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
+To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
+A panel of 19 possible Raf interaction partners (kinases, GTPases, scaffolding proteins etc)
+was constructed based upon known signaling pathways.
+We did not consider the most likely binding partner,
+the small GTPase RAS, as previous studies have ruled it out in MAPK activation in HL-60 cells \cite{Wang2008,Katagiri1994}.
+Total Raf was used as a bait protein for the immunoprecipitation studies.
+Interrogation of the Raf interactome suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
+Western blot analysis using total Raf and pS621 Raf specific antibodies confirmed the
+presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
+Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf, suggesting a direct physical interaction was possible.
+However, only the associations between Raf and Vav1 and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
+Furthermore, the Vav1 and Src associations were correlated with pS621 Raf abundance in the precipitate.
+Others proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
+The remaining 14 proteins were expressed in whole cell lysate (Fig. \ref{fig:ip-data}B),
+but were not detectable in the precipitate of Raf IP.
+Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
+although the signal intensity for Src was notably weak.
+However, GW5074 did not influence the association of CK2 or 14-3-3 with Raf, further demonstrating their independence from Raf phosphorylation.
+Interestingly, the Raf-Akt interaction qualitatively increased following treatment with GW5074;
+however, it remained unaffected by treatment with ATRA.
+Src family kinases are known to be important in myeloid differentiation \cite{Miranda2007} and their role in HL-60 differentiation has been investigated elsewhere \cite{Congleton:2012fk}.
+Given the existing work and variable reproducibility in the context of the Raf immunoprecipitate,
+we did not investigate the role of Src further in this study.
+Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with Raf activation following ATRA-treatment.
+Previous studies demonstrated that a Vav1-Slp76-Cbl-CD38 complex plays an important role in ATRA-induced MAPK activation and differentiation of HL-60 cells \cite{Shen2009}.
+Here we did not observe direct interaction of Raf with Cbl or Slp76;
+however, this complex could be involved upstream.
+
+%For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{fig:ip-data}A.
+Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
+Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers following ATRA exposure (Fig. \ref{fig:ip-data}D-F).
+ATRA treatment alone statistically significantly increased the G1/G0 percentage over the untreated control,
+while GW5074 alone had a negligible effect on the cell cycle distribution (Fig. \ref{fig:ip-data}D).
+Surprisingly, the combination of GW5074 and ATRA statistically significantly increased the G1/G0 population (82 $\pm$ 1\%)
+compared with ATRA alone (61 $\pm$ 0.5\%).
+Increased G1/G0 arrest following the combined treatment with GW5074 and ATRA was unexpected,
+as the combination of ATRA and the MEK inhibitor (PD98059) has been shown previously to decrease ATRA-induced growth arrest \cite{Yen1998}.
+However, growth arrest is not the sole indication of functional differentiation.
+Expression of the cell surface marker CD11b has also been shown to coincide with HL-60 cells myeloid differentiation \cite{Hickstein1989}.
+We measured CD11b expression, for the various treatment groups, using immuno-fluorescence flow cytometry 48 hr post-treatment.
+As with G1/G0 arrest, ATRA alone increased CD11b expression over the untreated control,
+while GW5074 further enhanced ATRA-induced CD11b expression (Fig. \ref{fig:ip-data}E).
+GW5074 alone had no statistically significant effect on CD11b expression, compared with the untreated control.
+Lastly, the inducible reactive oxygen species (ROS) response was used as a functional marker of differentiated neutrophils \cite{Congleton2011}.
+We measured the ROS response induced by the phorbol ester 12-O-tetradecanoylphorbol-13-acetate (TPA) using flow cytometry.
+Untreated cells showed no discernible TPA response,
+with only 7.0 $\pm$ 3.0\% ROS induction (Fig. \ref{fig:ip-data}F).
+Cells treated with ATRA had a significantly increased TPA response, 53 $\pm$ 7\% ROS induction 48 hr post-treatment.
+Treatment with both ATRA and GW5074 statistically significantly reduced ROS induction (22 $\pm$ 0.6\%) compared to ATRA alone.
+Interestingly, Western blot analysis did not detect a GW5074 effect on ATRA-induced expression of p47phox,
+a required upstream component of the ROS response  (Fig. \ref{fig:ip-data}F, bottom).
+Thus, the inhibitory effect of GW5074 on inducible ROS might occur downstream of p47phox expression.
+However, the ROS producing complex is MAPK dependent,
+therefore it is also possible that GW5074 inhibited ROS production by interfering
+with MAPK activation (in which case the p47Phox marker might not accurately reflect phenotypic conversion and differentiation).
 
@@
 \begin{figure}[!t]
-\includegraphics[width=0.9\textwidth]{./figs/Fig-1-IP-ExperimentalStudies.pdf}
-\caption{Investigation of a panel of possible Raf interaction partners in the presence and absence of ATRA.
-A: Species identified to precipitate out with Raf:
-first column shows Western blot analysis on total Raf immunoprecipitation with and without 24 hr ATRA treatment
-and the second on total lysate.
-B: The expression of species considered that did not precipitate out with Raf at levels detectable by Western blot analysis on total lysate.
-C: Effect of the Raf inhibitor GW5074 on Raf interactions as determined by Western blot analysis of total Raf immunoprecipitation.
-The Authors note the the signal associated with Src was found to be weak.
-D: Cell Cycle distribution as determined by flow cytometry indicated arrest induced by ATRA, which was increased by the addition of GW5074.
-E: Expression of the cell surface marker CD11b as determined by flow cytometry indicated increased expression induced by ATRA,
-which was enhanced by the addition of GW5074.
-F: Inducible reactive oxygen species (ROS) as determined by DCF flow cytometry. The functional differentiation response of ATRA treated cells
-was mitigated by GW5074.}\label{fig:ip-data}
-\end{figure}
-
-\begin{figure}[!t]
 \includegraphics[width=1.0\textwidth]{./figs/Fig-1-Network_v3.pdf}
@@
 
+\begin{figure}[!t]
+\includegraphics[width=0.9\textwidth]{./figs/Fig-1-IP-ExperimentalStudies.pdf}
+\caption{Investigation of a panel of possible Raf interaction partners in the presence and absence of ATRA.
+A: Species identified to precipitate out with Raf:
+first column shows Western blot analysis on total Raf immunoprecipitation with and without 24 hr ATRA treatment
+and the second on total lysate.
+B: The expression of species considered that did not precipitate out with Raf at levels detectable by Western blot analysis on total lysate.
+C: Effect of the Raf inhibitor GW5074 on Raf interactions as determined by Western blot analysis of total Raf immunoprecipitation.
+The Authors note the the signal associated with Src was found to be weak.
+D: Cell Cycle distribution as determined by flow cytometry indicated arrest induced by ATRA, which was increased by the addition of GW5074.
+E: Expression of the cell surface marker CD11b as determined by flow cytometry indicated increased expression induced by ATRA,
+which was enhanced by the addition of GW5074.
+F: Inducible reactive oxygen species (ROS) as determined by DCF flow cytometry. The functional differentiation response of ATRA treated cells
+was mitigated by GW5074.}\label{fig:ip-data}
+\end{figure}
+
 \clearpage

### 2017-01-23 Paper_v2.tex
Updated results

@@
 \section*{Results}
-We constructed an effective model of the ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
-The model connectivity was developed from literature and the studies presented here.
+We constructed an effective model of ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
+The model connectivity was developed from literature and the studies presented here (Table ZZ).
 We decomposed the ATRA program into three modules;
-a signal initiation module that was responsible for sensing and transforming the ATRA signal into activated cRaf-S621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
-a signal integration module that controlled the expression of upstream transcription factors given cRaf-S621 and activated Trigger (Fig. \ref{fig:network}B); and
-a phenotype module which encoded the expression of functional differentiation markers (Fig. \ref{fig:network}C).
+a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-S621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
+a signal integration module that controlled the expression of upstream transcription factors given cRaf-S621 and activated Trigger signals (Fig. \ref{fig:network}B); and
+a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors (Fig. \ref{fig:network}C).
+Each component of these modules was described by a mRNA and protein balance equation.
+Additionally, the signal initiation module also described the abundance of activated species e.g., Trigger and cRaf-S621 whose values were derived from
+unactivated Trigger and cRaf protein levels.
+Lastly, because the population of HL-60 cells was dividing (at least before ATRA-induced cell cycle arrest), we also considered a dilution term in all balance equations.
+The signal initiation module contained nine differential equations, while the signal integration and phenotype modules were collectively encoded by 54 differential equations.
+Model parameters were taken literature, or estimated from experimental data taken from literature using heuristic optimization (see materials and methods).
 

### 2017-01-24 Paper_v2.tex
Updated results
Working thru the results

@@
 We decomposed the ATRA program into three modules;
-a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-S621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
-a signal integration module that controlled the expression of upstream transcription factors given cRaf-S621 and activated Trigger signals (Fig. \ref{fig:network}B); and
+a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
+a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals (Fig. \ref{fig:network}B); and
 a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors (Fig. \ref{fig:network}C).
 Each component of these modules was described by a mRNA and protein balance equation.
-Additionally, the signal initiation module also described the abundance of activated species e.g., Trigger and cRaf-S621 whose values were derived from
+Additionally, the signal initiation module also described the abundance of activated species e.g., Trigger and cRaf-pS621 whose values were derived from
 unactivated Trigger and cRaf protein levels.
@@
 
-The signal initiation module recapitulated sustained signalsome/MAPK activation following exposure to 1$\mu$M ATRA (Fig. \ref{fig:model-fitting}A-B).
+The signal initiation module recapitulated sustained signalsome and MAPK activation following exposure to 1$\mu$M ATRA (Fig. \ref{fig:model-fitting}A-B).
 An ensemble of effective model parameters was estimated by minimizing the difference between simulations and time-series measurements of BLR1 mRNA and
-cRaf-S621 following the addition of 1$\mu$M ATRA. We focused on the S621 phosphorylation site of cRaf since enhanced phosphorylation at this site is a defining
+cRaf-pS621 following the addition of 1$\mu$M ATRA. We focused on the S621 phosphorylation site of cRaf since enhanced phosphorylation at this site is a defining
 characteristic of sustained MAPK activation in HL-60. The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A)
-and sustained phosphorylation of cRaf-pS621 (Fig. \ref{fig:model-fitting}B)
-in a growing population of HL-60 cells. However, the effective model failed to capture the decline of BLR1 message after 48 hr of ATRA exposure.
-Next, we tested the response of the signalsome/MAPK signaling module to different ATRA dosages.
+and sustained phosphorylation of cRaf-pS621 (Fig. \ref{fig:model-fitting}B) in a growing population of HL-60 cells.
+Together, the reinforcing positive feedback between the signalsome and MAPK led to sustained activation over multiple cellular generations.
+However, the effective model failed to capture the decline of BLR1 message after 48 hr of ATRA exposure.
+This suggested that we captured the logic leading to the onset of differentiation, but failed to describe program shutdown.
+Next, we tested the response of the signal initiation module to different ATRA dosages.
 
@@
 
-The signalsome/MAPK signaling model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
-Nullcline analysis predicted two stable steady-states and a single unstable state when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C).
+The signal initiation model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
+Phaseplane analysis predicted two stable steady-states when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C).
 In the lower stable state, neither the signalsome nor cRaf-pS621 were present (thus, the differentiation program was deactivated).
@@
 Morphological changes associated with differentiation were visible for ATRA $\geq$ 0.25 $\mu$M, suggesting the critical ATRA
-threshold was near this concentration. Next, we conducted washout ATRA washout experiments to determine if activated cells remained activated even in the absence of ATRA.
+threshold was near this concentration. Next, we conducted ATRA washout experiments to determine if activated cells remained activated in the absence of ATRA.
 HL-60 cells locked into an activated state remained activated following ATRA withdraw (Fig. \ref{fig:model-predictions}).
-Sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
-Thus, following activation, if we inhibited or removed elements from the effective circuit we expected the siganlsome and MAPK signals to decay.
+This sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
+Thus, following activation, if we inhibited or removed elements from the signal initiation module we expected the siganlsome and MAPK signals to decay.
 We simulated ATRA induced activation in the presence of kinase inhibitors, and without key circuit elements.
@@
 \ref{fig:model-predictions}A).
-In the presence of BLR1, the signalsome and cRaf-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, blue).
-On the other hand, BLR1 deletion removed the ability of the circuit to maintain a sustained MAPK response following the withdraw of ATRA (Fig. \ref{fig:model-predictions}B, gray).
+In the presence of BLR1, the signalsome and cRaf-pS621 signals were maintained following ATRA withdraw (Fig. \ref{fig:model-predictions}B, gray).
+On the other hand, BLR1 deletion removed the ability of the circuit to maintain a sustained MAPK response following the withdraw of ATRA (Fig. \ref{fig:model-predictions}B, blue).
 Lastly, washout experiments in which cells were exposed to 1$\mu$M ATRA for 24 hr, and then transferred to fresh media without ATRA,
 confirmed the persistence of the self sustaining activated state for up to 144 hr (Fig. \ref{fig:model-predictions}C).
-Thus, these experiments and simulations confirmed that reinforcing positive feedback likely drives the ATRA-induced differentiation program.
+Thus, these experiments confirmed that reinforcing positive feedback likely drives the ATRA-induced differentiation program.
 Next, we analyzed the ATRA-induced downstream gene expression program following signalsome and cRaf activation.
 
-The reduced order gene expression model described signal integration and ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
-The signalsome-MAPK model produced two outputs, Trigger and cRaf-S621 which drove the downstream differentiation program.
-In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex, regulated the expression of the transcription factors
-CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1. In turn, these transcription factors, in combination with cRaf-S621, regulated
-the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
-We assembled the connectivity of the signal integration program driven by Trigger, and the phenotypic program from literature (supplemental materials).
-We estimated the parameters of the signal integration and phenotype programs from previous studies which contained both steady-state and dynamic measurements
-of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
-The model simulations captured the time dependent expression of both CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A), and
-steady-state values for upstream members of the signal integration unit (Fig. \ref{fig:model-grn-simulations}B).
+The signal integration and phenotype modules described ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
+The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
+which then in turn activated the phenotypic program. In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex,
+regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
+In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
+We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table ZZ).
+We estimated the parameters which appeared in the control laws regulating these programs
+from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
+However, the bulk of the remaining model parameters were taken from directly from literature [REFHERE] and were not estimated from data in this study (see materials and methods).
+The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
+and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
+Taken together, the signal integration and phenotypic simulations were consistent with measurements, thereby validating the assumed molecular connectivity.
 
@@
 drives the differentiation program and signalsome formation.
-Both Trigger and activated cRaf-S621 drive a phenotype gene expression program responsible for differentiation.
-Trigger activates the expression of a series of transcription factors which in combination with cRaf-S621 result in phenotypic change.}\label{fig:network}
+Both Trigger and activated cRaf-pS621 drive a phenotype gene expression program responsible for differentiation.
+Trigger activates the expression of a series of transcription factors which in combination with cRaf-pS621 result in phenotypic change.}\label{fig:network}
 \end{figure}

### 2017-01-24 Paper_v2.tex
Working thru the results

@@
 from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
-However, the bulk of the remaining model parameters were taken from directly from literature [REFHERE] and were not estimated from data in this study (see materials and methods).
+However, the bulk of the remaining model parameters were taken from directly from literature [REFHERE] and were not estimated in this study (see materials and methods).
 The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),

### 2017-01-24 Paper_v2.tex
updating the MM

@@
 
-\subsubsection*{Effective ATRA differentiation model.}
-ATRA induced signaling events were modeled using saturation kinetics within an ordinary differential equation (ODE) framework:
+\subsubsection*{Model equations.}
+We decomposed the ATRA-induced differentiation program into three modules;
+a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (activated Trigger) signals;
+a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
+a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
+The output of the signal initiation module was the input to the gene expression model.
+For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$) abundance:
+\begin{eqnarray}
+	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
+	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
+\end{eqnarray}
+The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
+the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
+The specific transcription rate $r_{T,j}$ was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
+abundance of transcription factors, or other regulators influenced the expression of gene $j$.
+The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
 \begin{equation}
-	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\left(\mu + k_{d}\right) x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}
+	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
+\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
+and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
+The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
+Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
+The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.
+Lastly, the $\lambda_{j}$ term denotes the constitutive rate of expression of gene $j$.
+
+The gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced the expression of gene $j$.
+If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
+$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
+where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$,
+and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule which combines the individual regulatory inputs for gene $j$ into a single control term.
+In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations that resulted in gene expression [REFHERE]:
+\begin{equation}
+	{u}_{j} = \frac{W_{R_{1},j}+\displaystyle\sum_{n}W_{nj}f_{nj}}{1+W_{R_{1},j}+\displaystyle\sum_{d}W_{dj}f_{dj}}
 \end{equation}
-The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model, respectively.
-The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
-The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of process $j$.
-Typically, process $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
-The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
-If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
-Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
-Lastly, $\mu$ denotes the specific growth rate, and $k_{d}$ denotes the rate constant controlling cell death.
-Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
+The numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
+The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using hill like functions), while
+the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
+Given this formulation, the rate of constitutive expression was then given by:
+\begin{equation}
+	\lambda_{j} = \bar{r}_{T,j}\left(\frac{W_{R_{1},j}}{1+W_{R_{1},j}}\right)
+\end{equation}
+If a gene expression process had no modifying factors, $u_{j}=1$.
+Lastly, the specific translation rate was modeled as:
+\begin{equation}
+	r_{X,j} = V^{max}_{X}\left(\frac{L_{X,o}}{L_{X,j}}\right)\left(\frac{m_{j}}{K_{X}+m_{j}}\right)
+\end{equation}
+where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, and $K_{X}$ denotes a translation saturation constant.
+The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
+and the Ribosome abundance ($R_{2}$), $V^{max}_{X} = k_{X}\left(R_{2}\right)$. As was the case for transcription,
+we corrected the characteristic translation rate by the ratio of the length of a characteristic transcription normalized by the length of transcript $j$.
 
-Signaling rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the HL-60 model.
-The rate of an enzyme catalyzed process was modeled using saturation kinetics:
-\begin{equation}\label{eqn:rate-saturation}
-	\bar{r}_{j} = k_{j}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\left(\frac{x_{s}}{K_{js} + x_{s}}\right)
+The signal initiation, and integration modules required the level of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
+However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
+To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger.
+The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
+\begin{equation}
+	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\times{\texttt{ATRA}}}{1+\alpha\times{ATRA}}\right)
 \end{equation}
-where $k_{j}$ denotes the catalytic rate constant for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
-and $K_{js}$ denotes the saturation constant for species $s$ and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$.
+On the other hand, the abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
+The abundance of an activated signaling species $i$ was given by:
+\begin{equation}
+	\frac{dx_{i}}{dt}  =  r_{+,i}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
+\end{equation}
+The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model,
+respectively. The term $r_{+,i}\left(\mathbf{x},\mathbf{k}\right)$ denotes the rate of generation of activated species $i$,
+while  $\mu$ denotes the specific growth rate, and $k_{d,i}$ denotes the rate constant controlling the non-specific degradation of $x_{i}$.
+We have neglected deactivation reactions e.g., phosphatase activities.
+We assumed that signaling processes were fast compared to gene expression; this allowed us to approximate the signaling balance as:
+\begin{equation}
+	x_{i}^{*} \simeq \frac{r_{+,i}\left(\mathbf{x},\mathbf{k}\right)}{\left(\mu+k_{d,i}\right)} \qquad{i=1,\hdots,\mathcal{M}}
+\end{equation}
+The generation rate was written as the product of a kinetic term ($\bar{r}_{+,i}$) and a control term ($v_{i}$).
 The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
-For each rate, we used a rule-based approach to select from competing control factors.
 If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
@@
 If a process had no modifying factors, $v_{j}=1$.
+The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was given as:
+\begin{equation}
+	\bar{r}_{+,cRaf} = k_{+,cRaf}x_{s}\left(\frac{x_{cRaf}}{K_{+,cRaf}+x_{cRaf}}\right)
+\end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
+The formation of cRaf-pS621 was regulated by a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
+\begin{equation}
+	v_{+,cRaf} = \left(1-\frac{I}{K_{D}+I}\right)
+\end{equation}where $I$ denotes the MAPK inhibitor abundance, and $K_{D}$ the affinity constant for inhibitor $I$.
+
+% ATRA induced signaling events were modeled using saturation kinetics within an ordinary differential equation (ODE) framework:
+% \begin{equation}
+% 	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\left(\mu + k_{d}\right) x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}
+% \end{equation}
+% The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model, respectively.
+% The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
+% The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of process $j$.
+% Typically, process $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
+% The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
+% If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
+% Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
+% Lastly, $\mu$ denotes the specific growth rate, and $k_{d}$ denotes the rate constant controlling cell death.
+% Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
+%
+% Signaling rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the HL-60 model.
+% The rate of an enzyme catalyzed process was modeled using saturation kinetics:
+% \begin{equation}\label{eqn:rate-saturation}
+% 	\bar{r}_{j} = k_{j}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\left(\frac{x_{s}}{K_{js} + x_{s}}\right)
+% \end{equation}
+% where $k_{j}$ denotes the catalytic rate constant for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
+% and $K_{js}$ denotes the saturation constant for species $s$ and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$.
+% The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
+% For each rate, we used a rule-based approach to select from competing control factors.
+% If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
+% $v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
+% where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
+% The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
+% variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
+% If a process had no modifying factors, $v_{j}=1$.
 
@@
 
-The HL-60 model described both signal transduction and gene expression events following the addition of ATRA.
-The output of the signal transduction model was the input to the gene expression model.
-For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$):
-\begin{eqnarray}
-	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
-	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
-\end{eqnarray}
-The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
-the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
-The specific transcription rate was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
-abundance of transcription factors, or other regulators influenced the expression of gene $j$.
-The kinetic rate of transcription was modeled as:
-\begin{equation}
-	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
-\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
-and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$. The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths,
-where $L_{T,o}$ is a characteristic gene length, and $L_{T,j}$ denotes the length of gene $j$.
-Thus, the $\left(L_{T,o}/L_{T,j}\right)$ term is gene specific correction to the characteristic transcription rate.
-The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.
+% The HL-60 model described both signal transduction and gene expression events following the addition of ATRA.
 
-The gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
-If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
-$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
-where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$,
-and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule.
-In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations resulting in gene expression.
-Thus, the control variable $u_{j}$ took the form:
-\begin{equation}
-	{u}_{j} = \frac{W_{R_{1},j}+\displaystyle\sum_{n}W_{nj}f_{nj}}{1+W_{R_{1},j}+\displaystyle\sum_{d}W_{dj}f_{dj}}
-\end{equation}where the numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
-The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using hill like functions), while
-the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
-If a gene expression process had no modifying factors, $u_{j}=1$.
-Lastly, the specific translation rate was modeled as:
-\begin{equation}
-	r_{X,j} = V^{max}_{X}\left(\frac{m_{j}}{K_{X}+m_{j}}\right)
-\end{equation}
-where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, and $K_{X}$ denotes a translation saturation constant.
-The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
-and the Ribosome abundance ($R_{2}$), $V^{max}_{X} = k_{X}\left(R_{2}\right)$.
-
-In this study, we estimated the $W_{ij}$ parameters, and the parameters in the transfer functions $f_{dj}$ from gene expression data sets.
+\noindent\subsubsection*{Estimation of signaling and gene expression model parameters.}
+We estimated the $W_{ij}$ parameters, and the parameters in the transfer functions $f_{dj}$ from gene expression data sets.
 On the other hand, we estimated $k_{T}, k_{X}, \theta_{m,j}, \theta_{p,j}, R_{1}$ and $R_{2}$ using estimates of transcription and translation rates,
@@
 
-\noindent\subsubsection*{Estimation of signaling and gene expression model parameters.}
 Signal and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental data set $j$:
@@
 
-The signaling and gene expression model equations were implemented in Julia and solved using the \texttt{CVODE} routine of the Sundials package \cite{Julia,Hindmarsh2005}.
+The signaling and gene expression model equations were implemented in the Julia programming language, and solved using the \texttt{CVODE} routine of the Sundials package \cite{Julia,Hindmarsh2005}.
 The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.

### 2017-01-24 Paper_v2.tex
Working on MM

@@
 We constructed an effective model of ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
-The model connectivity was developed from literature and the studies presented here (Table ZZ).
+The model connectivity was developed from literature and the studies presented here (Table \ref{tbl:TF-network-connectivity}).
 We decomposed the ATRA program into three modules;
@@
 In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
-We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table ZZ).
+We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
 We estimated the parameters which appeared in the control laws regulating these programs
 from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
-However, the bulk of the remaining model parameters were taken from directly from literature [REFHERE] and were not estimated in this study (see materials and methods).
+However, the bulk of the remaining model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
 The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
@@
 
-\subsubsection*{Model equations.}
+\subsubsection*{Gene expression model equations.}
 We decomposed the ATRA-induced differentiation program into three modules;
@@
 
+\subsubsection*{Signaling model equations.}
 The signal initiation, and integration modules required the level of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
@@
 \begin{equation}
-	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\times{\texttt{ATRA}}}{1+\alpha\times{ATRA}}\right)
+	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\cdot{\mathrm{ATRA}}}{1+\alpha\cdot\mathrm{ATRA}}\right)
 \end{equation}
-On the other hand, the abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
+where $\alpha$ denotes a gain parameter; $\alpha = 0.0$ if ATRA is less than a threshold, and $\alpha = 0.1$ if ATRA is greater than the differentiation threshold.
+The abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
 The abundance of an activated signaling species $i$ was given by:
@@
 while $\mu$ denotes the specific growth rate, and $k_{d,i}$ denotes the rate constant controlling the non-specific degradation of $x_{i}$.
-We have neglected deactivation reactions e.g., phosphatase activities.
+We neglected deactivation reactions e.g., phosphatase activities.
 We assumed that signaling processes were fast compared to gene expression; this allowed us to approximate the signaling balance as:
@@
 If a process had no modifying factors, $v_{j}=1$.
-The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was given as:
+The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was modeled as:
 \begin{equation}
@@
 \end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
-The formation of cRaf-pS621 was regulated by a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
+The formation of cRaf-pS621 was regulated by only a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
 \begin{equation}
-	v_{+,cRaf} = \left(1-\frac{I}{K_{D}+I}\right)
-\end{equation}where $I$ denotes the MAPK inhibitor abundance, and $K_{D}$ the affinity constant for inhibitor $I$.
+	v_{+,cRaf} = \left(1-\frac{\mathrm{I}}{K_{D}+\mathrm{I}}\right)
+\end{equation}where $\mathrm{I}$ denotes the abundance of the MAPK inhibitor, and $K_{D}$ denotes the inhibitor affinity.
 
@@
 
-\noindent\subsubsection*{Estimation of signaling and gene expression model parameters.}
-We estimated the $W_{ij}$ parameters, and the parameters in the transfer functions $f_{dj}$ from gene expression data sets.
-On the other hand, we estimated $k_{T}, k_{X}, \theta_{m,j}, \theta_{p,j}, R_{1}$ and $R_{2}$ using estimates of transcription and translation rates,
-the half-life of a typical mRNA and protein, and a typical value for the copies per cell of RNA polymerase and ribosomes from literature \cite{Milo:2010aa}.
+\noindent\subsubsection*{Estimation of model parameters.}
+We estimated parameters appearing in the mRNA and protein balances, and the abundance of polymerases and ribosomes, from estimates of transcription and translation rates,
+the half-life of a typical mRNA and protein, and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table ZZ).
+For the remaining parameters, e.g., the $W_{ij}$ appearing in the control laws,
+or parameters appearing in the transfer functions $f_{dj}$, were estimated from the gene expression and signaling data sets discussed here.
 The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
@@
 
+\begin{center}
+\renewcommand{\arraystretch}{0.8}
+\tablefirsthead{
+\hline
+\rowcolor[gray]{0.8}\textbf{Actor}  & \textbf{Effect} & \textbf{Target} & \textbf{Cite}  \\
+\hline}
+\topcaption{Myelomonocytic transcription factor connectivity used in the signal integration and phenotype modules.}\label{tbl:TF-network-connectivity}
+\begin{scriptsize}
+\begin{supertabular}{c|c|l|l}
+\hline
+RAR$\alpha$ &  + & RAR$\alpha$ & \cite{Rishi1996} \\
+ 	&  + & PU.1 & \cite{Mueller2006}\\
+ 	&  + & C/EBP$\alpha$ & \cite{Friedman2007}\\
+ 	&  + & IRF-1 & \cite{Luo2006}\\
+ 	&  - & Oct4 & \cite{Sylvester1994}\\
+ 	&  + & CD38 & \cite{Drach1994}\\
+ 	&  + & p21 & \cite{Liu1996}\\
+ 	&  + & AhR & \cite{Bunaciu2013}\\
+ 	&  + & EGR1 & \cite{Balmer2002}\\
+\hline
+PPAR$\gamma$ &  + & C/EBP$\alpha$ & \cite{Rosen2002}\\
+ &  + & IRF-1 & \cite{Varley2009}\\
+ &  + & Oct1 & \cite{Bruemmer2003}\\
+ &  - & AP-1 & \cite{Delerive1999}\\
+ &  - & E2F & \cite{Altiok1997}\\
+ &  - & EGR1 & \cite{Fei2011}\\
+ &  + & CD38 & \cite{Song2012}\\
+ &  + & CD14 & \cite{Szanto2005}\\
+ &  + & p21 & \cite{Han2004}\\
+ &  - & p47phox & \cite{Von-Knethen2002}\\
+\hline
+PU.1 &  - & PPAR$\gamma$ & \cite{Dispirito2013}\\
+&  + & PU.1 & \cite{Chen1995}\\
+&  + & AP-1 & \cite{Steidl2006}\\
+&  + & EGR1 & \cite{Laslo2006}\\
+&  + & CD11b & \cite{Pahl1993}\\
+&  + & p21 & \cite{Yuki2013}\\
+&  + & p47phox & \cite{Li1999}\\
+\hline
+C/EBP$\alpha$ &  + & PPAR$\gamma$ & \cite{Rosen2002}\\
+ &  + & PU.1 & \cite{Dahl2003}\\
+ &  + & C/EBP$\alpha$ & \cite{Timchenko1995}\\
+ &  + & Gfi-1 & \cite{Lidonnici2010}\\
+ &  - & E2F & \cite{DAlo2003}\\
+ &  + & CD14 & \cite{Pan1999}\\
+ &  + & p21 & \cite{Harris2001}\\
+\hline
+IRF-1 &  + & CD38 & \cite{Bauvois1999}\\
+ &  + & p21 & \cite{Passioura2005}\\
+ &  - & PU.1 & \cite{Dahl2007}\\
+ &  - & C/EBP$\alpha$ & \cite{Duan2003}\\
+ &  - & E2F & \cite{Duan2003}\\
+ &  - & EGR1 & \cite{Laslo2006}\\
+ &  - & p21 & \cite{Duan2003}\\
+\hline
+Oct1 &  + & PU.1 & \cite{Chen1996}\\
+\hline
+AP-1 &  - & PPAR$\gamma$  & \cite{Delerive1999}\\
+&  + & PU.1 &  \cite{Behre1999}\\
+&  + & p21 &  \cite{Kardassis1999}\\
+\hline
+E2F &  + & E2F & \cite{Johnson1994}\\
+\hline
+EGR1 &  + & PPAR$\gamma$ & \cite{Fu2002}\\
+&  - & Gfi-1 & \cite{Mak2011}\\
+&  + & CD14 & \cite{Chen2004}\\
+\hline
+AhR &  + & AP-1 & \cite{Suh2002}\\
+&  + & IRF-1 & \cite{Shen2011}\\
+& - & Oct4 & \cite{Bunaciu2011}\\
+& - & PU.1 & \\
+
+\hline
+\end{supertabular}
+\end{scriptsize}
+\end{center}
+
+\clearpage
+
 \begin{figure}[!t]

### 2017-01-27 Paper_v2.tex
Working on the paper

@@
 The signal initiation module contained nine differential equations, while the signal integration and phenotype modules were collectively encoded by 54 differential equations.
-Model parameters were taken literature, or estimated from experimental data taken from literature using heuristic optimization (see materials and methods).
+Model parameters were taken literature (Table \ref{tbl:model-parameters}), or estimated from experimental data using heuristic optimization (see materials and methods).
 
@@
 
+\subsubsection*{Cell cycle arrest.}
+A defining feature of ATRA-induced differentiation is cell cycle arrest.
+We did not include a description of the restriction point logic, however, we do have a model of E2F transcription.
+We developed a heuristic
+
 \subsubsection*{Signaling model equations.}
@@
 
-The signaling and gene expression model equations were implemented in the Julia programming language, and solved using the \texttt{CVODE} routine of the Sundials package \cite{Julia,Hindmarsh2005}.
+The signaling and gene expression model equations were implemented in the Julia programming language, and solved using the \texttt{ODE23s} routine of the ODE package \cite{Julia}.
 The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
+Model parameters estimated directly or derived from literature, or data presented in this study are given in Table \ref{tbl:model-parameters}.
 
@@
 \begin{center}
-\renewcommand{\arraystretch}{0.8}
+% \renewcommand{\arraystretch}{0.8}
 \tablefirsthead{
 \hline
-\rowcolor[gray]{0.8}\textbf{Actor}  & \textbf{Effect} & \textbf{Target} & \textbf{Cite}  \\
-\hline}
+\rowcolor[gray]{0.9} Effector & Effect & Target & Source \\}
 \topcaption{Myelomonocytic transcription factor connectivity used in the signal integration and phenotype modules.}\label{tbl:TF-network-connectivity}
@@
 \end{scriptsize}
+\end{center}
+
+\clearpage
+
+\begin{center}
+
+	\tablefirsthead{
+	\hline
+	\rowcolor[gray]{0.9} Symbol & Description & Value & Units & Source  \\}
+	\topcaption{Model parameters estimated from literature.}\label{tbl:model-parameters}
+	\begin{scriptsize}
+	\begin{supertabular}{c|l|l|l|l}
+	\hline
+
+	$R_{1}$ & RNA polymerase abundance & 75,000 & copies/cell & [REFHERE] \\
+	$R_{2}$ & Ribosome abundance & 1 x 10$^{6}$ & copies/cell & [REFHERE] \\
+	$G_{i}$ & Characteristic gene abundance & 2 & copies/cell & \\
+	$K_{X}$ & Saturation constant transcription & 4,600 & copies/cell & \\
+	$K_{T}$ & Saturation constant translation & 100,000 & copies/cell & \\
+	& & & & \\
+	$t_{1/2,m}$ & characteristic mRNA half-life & 2 & hr & [REFHERE] \\
+	$t_{1/2,p}$ & characteristic protein half-life & 10 & hr & [REFHERE] \\
+	$\theta_{m,j}$ & characteristic mRNA degradation constant & 0.34 & hr$^{-1}$ & derived \\
+	$\theta_{p,j}$ & characteristic protein degradation constant & 0.07 & hr$^{-1}$ & derived \\
+	& & & & \\
+	$t_{d}$ & HL-60 doubling time & 19.5 & hr & this study \\
+	$\mu$ & growth rate & 0.035 & hr$^{-1}$ & derived \\
+	$k_{d}$ & death rate & 0.10$\mu$ & hr$^{-1}$ & derived \\
+	& & & & \\
+	$e_{T}$ & elongation rate RNA polymerase & 6 & nt/s & [REFHERE] \\
+	$e_{X}$ & elongation rate Ribosome & 5 & aa/s & [REFHERE] \\
+	$L_{T,o}$ & characteristic gene length & 15,000 & nt & [REFHERE] \\
+	$L_{X,o}$ & characteristic transcript length & 5,000 & nt & [REFHERE] \\
+	$k_{T}$ & characteristic transcription rate & 1.44 & hr$^{-1}$ & derived \\
+	$k_{X}$ & characteristic translation rate & 3.60 & hr$^{-1}$ & derived \\
+	& & & & \\
+	$D$	& Diameter of an HL-60 cell & 12 & $\mu$m$^{3}$ & [REFHERE] \\
+	$f_{C}$ & cytoplasmic fraction & 0.51 & & [REFHERE] \\
+
+	\hline
+	\end{supertabular}
+	\end{scriptsize}
+
 \end{center}

### 2017-01-27 Paper_v2.tex
E2F cell cycle

@@
 
-\subsubsection*{Cell cycle arrest.}
-A defining feature of ATRA-induced differentiation is cell cycle arrest.
-We did not include a description of the restriction point logic, however, we do have a model of E2F transcription.
-We developed a heuristic
+\subsubsection*{Effective cell cycle arrest model.}
+A defining feature of ATRA-induced differentiation is G1/G0 cell cycle arrest followed by differentiation.
+While we included E2F transcription, we do not consider a detailed model of restriction point regulation.
+Instead, we formulated an effective model of the fraction of cells undergoing ATRA-induced cell cycle arrest, $\mathcal{A}$ as:
+\begin{equation}\label{eq:a-eqn}
+	\frac{d\mathcal{A}}{dt} = \epsilon_{1}\left(\mathrm{E2F}\right)^{\epsilon_{2}}
+\end{equation}
+The parameters $\epsilon_{1}$ and $\epsilon_{2}$ were estimated from experimental cell distribution measurements by \texttt{log} transforming Eq. \eqref{eq:a-eqn}:
+\begin{eqnarray}
+\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{1}\right) & = & \log{\dot{\mathcal{A}}_{21}} \\
+\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{2}\right) & = & \log{\dot{\mathcal{A}}_{32}}
+\end{eqnarray}
+where subscripts denote the experimental time point (1 = 24 hr, 2 = 48 hr, and 3 = 72 hr), $\mathrm{E2F}_{i}$ denotes the $\mathrm{E2F}$ abundance at time point $i$,
+and $\dot{\mathcal{A}}_{ij}$ denotes the time derivative of the fraction of arrested cells (estimated from data).
+We calculated candidate values for $\epsilon_{j}$ by solving the log-transformed system of equations for 10 biological replicates.
 

### 2017-03-08 Paper_v2.tex
Working on Table 2

--- 2017-01-24 Paper_v2.tex
@@
 
+\subsubsection*{Effective cell cycle arrest model.}
+We formulated an effective model of the fraction of cells undergoing ATRA-induced cell cycle arrest, $\mathcal{A}$ as:
+\begin{equation}\label{eq:a-eqn}
+	\frac{d\mathcal{A}}{dt} = \epsilon_{1}\left(\mathrm{E2F}\right)^{\epsilon_{2}}
+\end{equation}where $\epsilon_{i}$ were unknown parameters, and $\mathrm{E2F}$ denotes the abundance of the E2F transcription factor.
+The parameters $\epsilon_{1}$ and $\epsilon_{2}$ were estimated from experimental cell cycle distribution measurements, by discretizing and \texttt{log} transforming Eq. \eqref{eq:a-eqn}:
+\begin{eqnarray}
+\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{1}\right) & = & \log{\dot{\mathcal{A}}_{21}} \\
+\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{2}\right) & = & \log{\dot{\mathcal{A}}_{32}}
+\end{eqnarray}
+where subscripts denote the experimental time point (1 = 24 hr, 2 = 48 hr, and 3 = 72 hr), $\mathrm{E2F}_{i}$ denotes the $\mathrm{E2F}$ abundance at time point $i$,
+and $\dot{\mathcal{A}}_{ij}$ denotes the time derivative of the fraction of arrested cells estimated between time point $i$ and $j$ (estimated from data).
+We calculated candidate values for $\epsilon_{j}$ by solving the log-transformed system of equations for 10 biological replicates.
+
 \subsubsection*{Signaling model equations.}
@@
 We estimated parameters appearing in the mRNA and protein balances, and the abundance of polymerases and ribosomes, from estimates of transcription and translation rates,
-the half-life of a typical mRNA and protein, and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table ZZ).
+the half-life of a typical mRNA and protein, and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table \ref{tbl:model-parameters}).
 For the remaining parameters, e.g., the $W_{ij}$ appearing in the control laws,
@@
 \end{scriptsize}
+\end{center}
+
+\clearpage
+
+\begin{center}
+
+	\tablefirsthead{
+	\hline
+	\rowcolor[gray]{0.9} Symbol & Description & Value & Units & Source  \\}
+	\topcaption{Characteristic model parameters estimated from literature.}\label{tbl:model-parameters}
+	\begin{scriptsize}
+	\begin{supertabular}{c|l|l|l|l}
+	\hline
+
+	$R_{1}$ & RNA polymerase abundance & 75,000 & copies/cell & \cite{Jackson:2000aa,Zhao:2014aa} \\
+	$R_{2}$ & Ribosome abundance & 1 x 10$^{6}$ & copies/cell & \cite{Freitas2004} \\
+	$G_{i}$ & Characteristic gene abundance & 2 & copies/cell & this study\\
+	$K_{X}$ & Saturation constant transcription & 4,600 & copies/cell & this study\\
+	$K_{T}$ & Saturation constant translation & 100,000 & copies/cell & this study\\
+	& & & & \\
+	$t_{1/2,m}$ & characteristic mRNA half-life (transcription factor) & 2 & hr & \cite{Yang:2003aa} \\
+	$t_{1/2,p}$ & characteristic protein half-life & 10 & hr & \cite{Doherty:2009aa} \\
+	$\theta_{m,j}$ & characteristic mRNA degradation constant & 0.34 & hr$^{-1}$ & derived \\
+	$\theta_{p,j}$ & characteristic protein degradation constant & 0.07 & hr$^{-1}$ & derived \\
+	& & & & \\
+	$t_{d}$ & HL-60 doubling time & 19.5 & hr & this study \\
+	$\mu$ & growth rate & 0.035 & hr$^{-1}$ & derived \\
+	$k_{d}$ & death rate & 0.10$\mu$ & hr$^{-1}$ & derived \\
+	& & & & \\
+	$e_{T}$ & elongation rate RNA polymerase & 6 & nt/s & \cite{Darzacq:2007aa} \\
+	$e_{X}$ & elongation rate Ribosome & 5 & aa/s & \cite{Bostrom:1986aa} \\
+	$L_{T,o}$ & characteristic gene length & 15,000 & nt & [REFHERE] \\
+	$L_{X,o}$ & characteristic transcript length & 5,000 & nt & [REFHERE] \\
+	$k_{T}$ & characteristic transcription rate & 1.44 & hr$^{-1}$ & derived \\
+	$k_{X}$ & characteristic translation rate & 3.60 & hr$^{-1}$ & derived \\
+	& & & & \\
+	$D$	& Diameter of an HL-60 cell & 12.4 & $\mu$m$^{3}$ & \cite{Rosenbluth:2006aa} \\
+	$f_{C}$ & cytoplasmic fraction & 0.51 & dimensionless & \cite{Rosenbluth:2006aa} \\
+
+	\hline
+	\end{supertabular}
+	\end{scriptsize}
+
 \end{center}

### 2017-03-08 Paper_v2.tex
Completed table 2

@@
 	$e_{X}$ & elongation rate Ribosome & 5 & aa/s & \cite{Bostrom:1986aa} \\
-	$L_{T,o}$ & characteristic gene length & 15,000 & nt & [REFHERE] \\
-	$L_{X,o}$ & characteristic transcript length & 5,000 & nt & [REFHERE] \\
+	$L_{T,o}$ & characteristic gene length & 15,000 & nt & \cite{Meyers2004} \\
+	$L_{X,o}$ & characteristic transcript length & 5,000 & nt & derived \\
 	$k_{T}$ & characteristic transcription rate & 1.44 & hr$^{-1}$ & derived \\

### 2017-04-04 Paper_v2.tex
some updates

@@
 
-\subsubsection*{Gene expression model equations.}
+\subsubsection*{Effective gene expression model equations.}
 We decomposed the ATRA-induced differentiation program into three modules;
@@
 The output of the signal initiation module was the input to the gene expression model.
-For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$) abundance:
+For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$), protein ($p_{j}$) and signaling species abundance:
 \begin{eqnarray}
 	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
-	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
+	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}\\\label{eqn-signaling-balances}
+	\mathbf{g}\left(p_{1},\hdots,p_{\mathcal{G}},\kappa\right) &=& \mathbf{0}
 \end{eqnarray}
@@
 and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule which combines the individual regulatory inputs for gene $j$ into a single control term.
-In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations that resulted in gene expression [REFHERE]:
+In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations that resulted in gene expression \cite{Moon:2012ab}:
 \begin{equation}
@@
 
-\subsubsection*{Effective cell cycle arrest model.}
-We formulated an effective model of the fraction of cells undergoing ATRA-induced cell cycle arrest, $\mathcal{A}$ as:
-\begin{equation}\label{eq:a-eqn}
-	\frac{d\mathcal{A}}{dt} = \epsilon_{1}\left(\mathrm{E2F}\right)^{\epsilon_{2}}
-\end{equation}where $\epsilon_{i}$ were unknown parameters, and $\mathrm{E2F}$ denotes the abundance of the E2F transcription factor.
-The parameters $\epsilon_{1}$ and $\epsilon_{2}$ were estimated from experimental cell cycle distribution measurements, by discretizing and \texttt{log} transforming Eq. \eqref{eq:a-eqn}:
-\begin{eqnarray}
-\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{1}\right) & = & \log{\dot{\mathcal{A}}_{21}} \\
-\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{2}\right) & = & \log{\dot{\mathcal{A}}_{32}}
-\end{eqnarray}
-where subscripts denote the experimental time point (1 = 24 hr, 2 = 48 hr, and 3 = 72 hr), $\mathrm{E2F}_{i}$ denotes the $\mathrm{E2F}$ abundance at time point $i$,
-and $\dot{\mathcal{A}}_{ij}$ denotes the time derivative of the fraction of arrested cells estimated between time point $i$ and $j$ (estimated from data).
-We calculated candidate values for $\epsilon_{j}$ by solving the log-transformed system of equations for 10 biological replicates.
-
 \subsubsection*{Signaling model equations.}
@@
 However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
-To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger.
+To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger (shown generally as Eq \eqref{eqn-signaling-balances}).
 The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
@@
 \end{equation}where $\mathrm{I}$ denotes the abundance of the MAPK inhibitor, and $K_{D}$ denotes the inhibitor affinity.
+
+\subsubsection*{Effective cell cycle arrest model.}
+We formulated an effective model of the fraction of cells undergoing ATRA-induced cell cycle arrest, $\mathcal{A}$ as:
+\begin{equation}\label{eq:a-eqn}
+	\frac{d\mathcal{A}}{dt} = \epsilon_{1}\left(\mathrm{E2F}\right)^{\epsilon_{2}}
+\end{equation}where $\epsilon_{i}$ were unknown parameters, and $\mathrm{E2F}$ denotes the abundance of the E2F transcription factor.
+The parameters $\epsilon_{1}$ and $\epsilon_{2}$ were estimated from experimental cell cycle distribution measurements, by discretizing and \texttt{log} transforming Eq. \eqref{eq:a-eqn}:
+\begin{eqnarray}
+\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{1}\right) & = & \log{\dot{\mathcal{A}}_{21}} \\
+\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{2}\right) & = & \log{\dot{\mathcal{A}}_{32}}
+\end{eqnarray}
+where subscripts denote the experimental time point (1 = 24 hr, 2 = 48 hr, and 3 = 72 hr), $\mathrm{E2F}_{i}$ denotes the $\mathrm{E2F}$ abundance at time point $i$,
+and $\dot{\mathcal{A}}_{ij}$ denotes the time derivative of the fraction of arrested cells estimated between time point $i$ and $j$ (estimated from data).
+We calculated candidate values for $\epsilon_{j}$ by solving the log-transformed system of equations for 10 biological replicates.
 

### 2017-04-13 Paper_v2.tex
Final figures

@@
 \vspace{0.05in}
-{\par \centering \large{Ryan Tasseff, Holly A. Jensen, Wei Dai, Rodica P. Bunaciu$^{\dag}$, Johanna Congleton$^{\dag}$, Andrew Yen$^{\dag}$, and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Ryan Tasseff, Holly A.
+Jensen, Wei Dai, Katherine Rogers, Adithya Sagar, Rodica P.
+Bunaciu$^{\dag}$, Johanna Congleton$^{\dag}$, Andrew Yen$^{\dag}$, and Jeffrey D.
+Varner$^{*}$}}
 \vspace{0.10in}
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=0.85\textwidth]{./figs/Fig-5-GRN-Simulations.pdf}
+\includegraphics[width=0.85\textwidth]{./figs/Fig-4-GRN-Simulations.pdf}
+\caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
+A: CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
+B: Gene expression at t = 48 hr following ATRA exposure.
+Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-grn-simulations}
+\end{figure}
+
+\begin{figure}[!t]\centering
+\includegraphics[width=0.85\textwidth]{./figs/Fig-5-GRN-CellCycle-Predictions.pdf}
+\caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
+A: CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
+B: Gene expression at t = 48 hr following ATRA exposure.
+Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-grn-simulations}
+\end{figure}
+
+\begin{figure}[!t]\centering
+\includegraphics[width=0.85\textwidth]{./figs/Fig-6-GRN-Robustness.pdf}
 \caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
@@
 \begin{figure}[!t]
-\includegraphics[width=0.9\textwidth]{./figs/Fig-1-IP-ExperimentalStudies.pdf}
+\includegraphics[width=0.9\textwidth]{./figs/Fig-7-IP-ExperimentalStudies.pdf}
 \caption{Investigation of a panel of possible Raf interaction partners in the presence and absence of ATRA.

### 2017-04-14 Paper_v2.tex
Edits to materials and methods

@@
 The signal initiation model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
-Phaseplane analysis predicted two stable steady-states when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C).
+Phaseplane analysis predicted two stable steady-states when ATRA was present below a critical threshold, and only a single steady-state above the threshold (Fig. \ref{fig:model-fitting}C).
 In the lower stable state, neither the signalsome nor cRaf-pS621 were present (thus, the differentiation program was deactivated).
@@
 \subsubsection*{Signaling model equations.}
-The signal initiation, and integration modules required the level of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
+The signal initiation, and integration modules required the abundance of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
 However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
@@
 
-\subsubsection*{Effective cell cycle arrest model.}
-We formulated an effective model of the fraction of cells undergoing ATRA-induced cell cycle arrest, $\mathcal{A}$ as:
+\noindent\subsubsection*{Estimation of gene expression model parameters.}
+We estimated parameters appearing in the mRNA and protein balances, the abundance of polymerases and ribosomes,
+transcription and translation rates, the half-life of a typical mRNA and protein,
+and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table \ref{tbl:model-parameters}).
+The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
+Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
+Thus, the bulk of our gene expression model parameters were based directly upon literature values, and were not adjusted during model identification.
+However, the remaining parameters, e.g., the $W_{ij}$ appearing in the gene expression control laws,
+or parameters appearing in the transfer functions $f_{dj}$, were estimated from the protein expression and signaling data sets discussed here.
+
+Signaling and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental protein data set $j$.
+We measured the squared difference in the scale, fold change and shape for protein $j$:
+\begin{equation}\label{eqn:objective-function}
+	E_{j}(\mathbf{k}) = \biggl(\mathcal{M}_{j}\left(t_{-}\right)-\hat{y}_{j}\left(t_{-},\mathbf{k}\right)\biggr)^2+\sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\mathcal{M}^{\prime}_{ij}-y^{\prime}_{ij}(\mathbf{k})\biggr)^2
+\end{equation}
+The first term in Eqn. \eqref{eqn:objective-function} quantified the initial \textit{scale} error, directly before the addition of ATRA.
+In this case, $\mathcal{M}_{j}\left(t_{-}\right)$ (the approximate concentration of protein $j$ before the addition of ATRA) was estimated from literature.
+This term was required because the protein measurements were reported as the fold-change; thus, the data was normalized by a control
+value measured before the addition of ATRA. However, the model operated on a physical scale.
+The first term allowed the model to capture physically realistic changes following ATRA addition.
+The second term quantified the difference in the \textit{fold-change} of protein $j$ as a function of time.
+The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote the scaled experimental observations and simulation outputs (fold-change; protein normalized by control value directly before ATRA addition)
+at time $i$ from protein $j$,
+where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
+Lastly, the third term of the objective function measured the difference in the \textit{shape} of the measured and simulated protein levels.
+The scaled value $0\leq\mathcal{M}^{\prime}_{ij}\leq{1}$ was given by:
+\begin{equation}\label{norm_exp_data}
+\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
+\end{equation}
+where $\mathcal{M}^{\prime}_{ij}=0$ and $\mathcal{M}^{\prime}_{ij}=1$  describe the lowest (highest) intensity bands.
+A similar scaling was used for the simulation output.
+We minimized the total model residual $\sum_{j}E_{j}$ using a heuristic direct-search optimization procedure, subject to box constraints on the parameter values,
+starting from a random initial parameter guess. Each downhill step was archived and used for ensemble calculations.
+The optimization procedure have been reported previously [REFHERE].
+
+% Model parameters estimated directly or derived from literature, or data presented in this study are given in Table \ref{tbl:model-parameters
+\subsubsection*{Estimation of an effective cell cycle arrest model.}
+We formulated an effective N-order polynomial model of the fraction of cells undergoing ATRA-induced cell cycle arrest at time t, $\hat{\mathcal{A}}(t)$, as:
 \begin{equation}\label{eq:a-eqn}
-	\frac{d\mathcal{A}}{dt} = \epsilon_{1}\left(\mathrm{E2F}\right)^{\epsilon_{2}}
-\end{equation}where $\epsilon_{i}$ were unknown parameters, and $\mathrm{E2F}$ denotes the abundance of the E2F transcription factor.
-The parameters $\epsilon_{1}$ and $\epsilon_{2}$ were estimated from experimental cell cycle distribution measurements, by discretizing and \texttt{log} transforming Eq. \eqref{eq:a-eqn}:
-\begin{eqnarray}
-\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{1}\right) & = & \log{\dot{\mathcal{A}}_{21}} \\
-\log\epsilon_{1}+\epsilon_{2}\log\left(\mathrm{E2F}_{2}\right) & = & \log{\dot{\mathcal{A}}_{32}}
-\end{eqnarray}
-where subscripts denote the experimental time point (1 = 24 hr, 2 = 48 hr, and 3 = 72 hr), $\mathrm{E2F}_{i}$ denotes the $\mathrm{E2F}$ abundance at time point $i$,
-and $\dot{\mathcal{A}}_{ij}$ denotes the time derivative of the fraction of arrested cells estimated between time point $i$ and $j$ (estimated from data).
-We calculated candidate values for $\epsilon_{j}$ by solving the log-transformed system of equations for 10 biological replicates.
+	\hat{\mathcal{A}}\left(t\right)\simeq a_{0}+\sum_{i=1}^{N-1}a_{i}\phi_{i}\left(\mathbf{p}(t),t\right)
+\end{equation}where $a_{i}$ were unknown parameters, and $\phi_{i}\left(\mathbf{p}(t),t\right)$ denotes a basis function.
+The basis functions were dependent upon the system state; in this study, we assumed $N = 4$ and basis functions of the form:
+\begin{equation}
+	\phi_{i}\left(\mathbf{p}(t),t\right) = \left(\frac{t}{T}+\frac{p21}{E2F}\Bigr|_{t}\right)^{(i-1)}
+\end{equation}
+The parameters $a_{0},\hdots,a_{3}$ were estimated directly from cell-cycle measurements (biological replicates) using least-squares.
+% The abundances of the $p21$ and $E2F$ proteins were taken from the gene expression model solution. The least squares solution for the polynomial
+% coefficients was given by:
+% \begin{equation}
+% 	\mathbf{a} = \left[\left(\mathbf{V}^{T}\mathbf{V}\right)^{-1}\mathbf{V}^{T}\right]\mathcal{A}(t)
+% \end{equation}where $\mathbf{V}$ denotes the Vandermonde matrix constructed from the basis functions, and $\mathcal{A}(t)$ denotes the vector of
+% measured G0 fractions following ATRA exposure.
+
+\subsubsection*{Availability of model code.}
+The signaling and gene expression model equations, and the parameter estimation procedure, were implemented in the Julia programming language.
+The model equations were solved using the \texttt{ODE23s} routine of the ODE package \cite{Julia}.
+The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
 
@@
 
-\noindent\subsubsection*{Estimation of model parameters.}
-We estimated parameters appearing in the mRNA and protein balances, and the abundance of polymerases and ribosomes, from estimates of transcription and translation rates,
-the half-life of a typical mRNA and protein, and typical values for the copies per cell of RNA polymerase and ribosomes from literature (Table \ref{tbl:model-parameters}).
-For the remaining parameters, e.g., the $W_{ij}$ appearing in the control laws,
-or parameters appearing in the transfer functions $f_{dj}$, were estimated from the gene expression and signaling data sets discussed here.
-The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
-Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
-Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
-The values used for the characteristic transcription/translation parameters, degradation constants and
-macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.
-
-Signal and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental data set $j$:
-\begin{equation}\label{eqn:objective-function}
-	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
-\end{equation}
-The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs at time $i$ from training set $j$,
-where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
-The first term in Eqn. \eqref{eqn:objective-function} quantified the relative simulation error.
-We used immunoblot intensity measurements for model training.
-Thus, we trained the model on the \emph{relative} change between bands within each data set.
-Suppose we have the intensity of species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$.
-The scaled value $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ is given by:
-\begin{equation}\label{norm_exp_data}
-\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
-\end{equation}
-where $\hat{\mathcal{M}}_{ij}=0$ and $\hat{\mathcal{M}}_{ij}=1$  describe the lowest (highest) intensity bands.
-A similar scaling was used for the simulation output. The second term in the objective function ensured a realistic concentration scale was estimated by the model.
-We set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
-We minimized the total model residual $\sum_{j}E_{j}$ using heuristic optimization starting from a random initial parameter guess.
-
-The signaling and gene expression model equations were implemented in the Julia programming language, and solved using the \texttt{ODE23s} routine of the ODE package \cite{Julia}.
-The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
-Model parameters estimated directly or derived from literature, or data presented in this study are given in Table \ref{tbl:model-parameters}.
-
 % All code was implemented in the Octave programming language, and is available for
@@
 \section*{Author's contributions}
-J.V and A.Y directed the study. R.T, H.J and J.C conducted the cell culture measurements.
-J.V and W.D developed the reduced order HL-60 models and the parameter ensemble.
-W.D analyzed the model ensemble, and generated figures for the manuscript.
+J.V and A.Y directed the study. R.T, H.J, R.B and J.C conducted the cell culture measurements.
+J.V, R.B, W.D, K.R and A.S developed the reduced order HL-60 models and the parameter ensemble.
+W.D and J.V analyzed the model ensemble, and generated figures for the manuscript.
 The manuscript was prepared and edited for publication by W.D, A.Y and J.V.

### 2017-04-19 Paper_v2.tex
Some updates
Almost there ... one more paragraph

@@
 We estimated the parameters which appeared in the control laws regulating these programs
-from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA [REFHERE].
+from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
 However, the bulk of the remaining model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
@@
 and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
-Taken together, the signal integration and phenotypic simulations were consistent with measurements, thereby validating the assumed molecular connectivity.
+Lastly, we used the \textit{predicted} values of the P21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
+The phenotype module predicted P21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
+We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
+The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and wash robust to uncertainty in the measured data (Fig.
+\ref{fig:model-cell-cycle-simulations}B, gray).
+Taken together, the output of the signal integration and phenotypic modules was consistent with time-series and steady-state measurements, thereby validating the assumed molecular connectivity.
+Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
+Next, we explored which nodes and interactions between nodes in the signal integration module most influenced the system response.
 
@@
 Here we did not observe direct interaction of Raf with Cbl or Slp76;
-however, this complex could be involved upstream.
+however, this complex could be involved upstream. Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
 
 %For more details on GW5074 including dose response see Supplemental Materials including Fig. \ref{fig:ip-data}A.
-Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
 Inhibition of Raf kinase activity modulated MAPK activation and differentiation markers following ATRA exposure (Fig. \ref{fig:ip-data}D-F).
@@
 starting from a random initial parameter guess. Each downhill step was archived and used for ensemble calculations.
-The optimization procedure have been reported previously [REFHERE].
+The optimization procedure (a covariance matrix adaptation evolution strategy) has been reported previously \cite{Igel:2007aa}.
 
@@
 \includegraphics[width=0.85\textwidth]{./figs/Fig-5-GRN-CellCycle-Predictions.pdf}
-\caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
-A: CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
-B: Gene expression at t = 48 hr following ATRA exposure.
-Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-grn-simulations}
+\caption{Model simulation of HL-60 cell-cycle arrest following exposure to 1$\mu$M ATRA at t = 0 hr.
+A: Predicted P21 and E2F expression levels for the best parameter set following ATRA exposure at time t = 0 hr.
+B: Estimated fraction of HL-60 cells in G0 arrest following ATRA exposure at time t = 0 hr.
+The gray region denotes the 95\% confidence estimate of the polynomial model.
+Experimental data in panel B was reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-cell-cycle-simulations}
 \end{figure}
@@
 \includegraphics[width=0.85\textwidth]{./figs/Fig-6-GRN-Robustness.pdf}
-\caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
-A: CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
-B: Gene expression at t = 48 hr following ATRA exposure.
-Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-grn-simulations}
+\caption{Robustness of the HL-60 differentiation program following exposure to 1$\mu$M ATRA at t = 0 hr.
+A: Singular value decomposition of the system response ($l^{2}$-norm between the perturbed and nominal state) following pairwise gene knockout simulations using the best fit parameter set.
+The percentage at the top of each column describes the fraction of the variance in the system state captured by the node combinations in the rows.
+B: Singular value decomposition of the system response ($l^{2}$-norm between the perturbed and nominal state) following the pairwise removal of connections from the PU.1 and AP1 nodes.}\label{fig:model-response-ko}
 \end{figure}

### 2017-04-19 Paper_v2.tex
Paragraph for gene KOs

@@
 
-The signal integration and phenotype modules described ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
+The signal integration and phenotype modules described ATRA-induced gene expression in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
 The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
@@
 % Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with cRaf activation following ATRA-treatment.
+
+The PU.1 and AP1 proteins were critical regulators of the ATRA-induced signal integration and phenotypic programs (Fig. \ref{fig:model-response-ko}).
+We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
+The difference between the system state with and without the gene knockouts (encoded in a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
+All simulations were conducted using the best fit parameter set.
+The first four modes described $>$99\% of the gene knockout variance,
+with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent GFI1 and CEBPa (Fig. \ref{fig:model-response-ko}A).
+To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
+SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
+The most sensitive interactions for the PU.1 protein involved the regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
+On the other other, the most sensitive connections for AP1 involved the regulation of P21 expression, and the mutual activation of PU.1 and AP1 expression.
+Taken together, these results suggested that the PU.1 and AP1 proteins acted as important regulators for both the signal integration and phenotype modules.
+However, this analysis did not given insight into which components of the signal initiation module was important, and particularly the nature of the signalsome complex.
+Toward this question, we experimentally explored the composition and regulation of the signalsome complex.
 

### 2017-04-19 Paper_v2.tex
Fixed prot names in figs and text

--- 2017-04-19 Paper_v2.tex
@@
 Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
-Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the siganlsome.
+Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the signalsome.
 A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
@@
 Model simulations were also consistent with measurements of the influence of MAPK inhibitors, and the failure of BLR1 knockout cells to differentiate when exposed to ATRA.
-Lastly, we showed by through immunoprecipitation studies, that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the siganlsome complex.
+Lastly, we showed by through immunoprecipitation studies, that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the signalsome complex.
 Taken together, these findings when combined with other literature evidence,
@@
 This sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
-Thus, following activation, if we inhibited or removed elements from the signal initiation module we expected the siganlsome and MAPK signals to decay.
+Thus, following activation, if we inhibited or removed elements from the signal initiation module we expected the signalsome and MAPK signals to decay.
 We simulated ATRA induced activation in the presence of kinase inhibitors, and without key circuit elements.
@@
 and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
-Lastly, we used the \textit{predicted} values of the P21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
-The phenotype module predicted P21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
+Lastly, we used the \textit{predicted} values of the p21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
+The phenotype module predicted p21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
 We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
@@
 
-The composition of the siganlsome, and the kinase ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
+The PU.1 and AP1 proteins were important regulators of the ATRA-induced signal integration and phenotypic programs (Fig. \ref{fig:model-response-ko}).
+We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
+The difference between the system state with and without the gene knockouts (encoded as a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
+All simulations were conducted using the best fit parameter set.
+The first four modes described $>$99\% of the gene knockout variance,
+with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent Gfi1 and C/EBPa (Fig. \ref{fig:model-response-ko}A).
+To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
+SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
+The most sensitive interactions for the PU.1 protein involved the C/EBPa-dependent regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
+On the other other, the most sensitive connections for AP1 involved the C/EBPa-dependent regulation of p21 expression, and the mutual activation of PU.1 and AP1 expression.
+Taken together, these results suggested that the PU.1 and AP1 proteins acted as important self-reinforcing regulators for both the signal integration and phenotype modules.
+The analysis suggested that the PU.1 signaling axis promoted the formation of the neutrophil NADPH oxidase (through p47Phox), while AP1 was responsible for cell cycle arrest (through p21).
+However, this analysis did not give insight into which upstream components of the signal initiation module were important.
+Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.
+
+The composition of the signalsome, and the kinase ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
 To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
@@
 
-We conducted immunoprecipitation studies and identified a limited number of ATRA-dependent and -independent Raf interaction partners.
+Analysis of the signal integration and phenotype modules suggested that PU.1 and AP1 regulated distinct phenotypic axes following ATRA exposure.
+
+Immunoprecipitation studies identified a limited number of ATRA-dependent and -independent Raf interaction partners.
 While we were unable to detect the association of Raf with common kinases and GTPases such as PKC, PKA, p38, Rac and Rho,
@@
 This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRaf-pS261 abundance.
-We could model this by separating siganlsome formation into an inactive precursor pool that is transformed to a long-lived activated siganlsome by MAPK activation.
+We could model this by separating signalsome formation into an inactive precursor pool that is transformed to a long-lived activated signalsome by MAPK activation.
 We should also explore adding additional downstream biological modules to this skeleton model, for example the upregulation of reactive oxygen markers such as p47Phox or
@@
 \caption{Model simulation of HL-60 cell-cycle arrest following exposure to 1$\mu$M ATRA at t = 0 hr.
-A: Predicted P21 and E2F expression levels for the best parameter set following ATRA exposure at time t = 0 hr.
+A: Predicted p21 and E2F expression levels for the best parameter set following ATRA exposure at time t = 0 hr.
 B: Estimated fraction of HL-60 cells in G0 arrest following ATRA exposure at time t = 0 hr.

### 2017-04-19 Paper_v2.tex
Working on discussion

@@
 
-The PU.1 and AP1 proteins were important regulators of the ATRA-induced signal integration and phenotypic programs (Fig. \ref{fig:model-response-ko}).
+The PU.1 and AP1 proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
 We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
@@
 Analysis of the signal integration and phenotype modules suggested that PU.1 and AP1 regulated distinct phenotypic axes following ATRA exposure.
+In particular, deletion of PU.1 and AP1 was predicted to lead to largest shift in the system state.
 
@@
 Constitutively bound 14-3-3 may also stabilize the S621 phosphorylation, modulate the activity and/or up-regulate autophosphorylation.
-Activated Raf can then drive ERK activation, which in turn closes the positive feedback loop by activating Raf transcription factors,
-e.g. Sp1 and/or STAT1 \cite{Kim2009,Milanini-Mongiat2002,Zhang2004,Li2006}.
+Activated Raf can then drive ERK activation, which in turn closes the positive feedback loop by activating Raf transcription factors
+e.g., Sp1 and/or STAT1 \cite{Kim2009,Milanini-Mongiat2002,Zhang2004,Li2006}.
 We tested this working hypothesis using mathematical modeling.
@@
 
-Several engineered, or naturally occurring systems involved in cell fate decisions incorporate positive feedback and bistability \cite{Ferrell:2002aa}.
+Several engineered, or naturally occurring systems involved in cell fate decisions incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
 One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
@@
 The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
-This is similar to the differentiation circuit presented here; ATRA drives signalsome which activates MAPK, cell-cycle arrest, differentiation and signalsome.
-Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, they share similar cell fate decision architectures.
+This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
+Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
+Reinforcing feedback and bistability has also been implicated in hematopoietic stem cell fate determination.
+Laslo et al showed that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBPa), encodes a bistable switch that
+results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
 Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
 Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
-Thus, while positive feedback is sometimes not desirable in man made systems, it may be at the core of a diverse variety of
+Thus, while positive feedback is often undesirable in human engineered systems, it may be at the core of a diverse variety of
 cell fate programs and other networks important to human health.
 
-Model performance was impressive given its limited size. However, there were several issues to explore further.
+There were several issues that can be explored further with the effective ATRA differentiation model.
 First, there was likely missing connectivity in the effective differentiation circuit.

### 2017-04-20 Paper_v2.tex
working on discussion

@@
 
+% The presence of Vav1 in Raf/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
+% Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
+% of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
+% While its function in the signalsome is unclear, Vav1 has been shown to associate with a Cbl-Slp76-CD38 complex in an ATRA-dependent manner;
+% furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevented
+% ATRA-induced MAPK activation \cite{Shen2009}.
+% Thus, interaction of Cbl-Slp76-Vav1 and CD38 appears to be required for transmission of the ATRA signal by the signalsome.
+
 \section*{Discussion}
-In this study, we presented an effective model of ATRA-inducible differentiation of HL-60 cells which
-encoded positive feedback between the ATRA-inducible signalsome complex and the MAPK pathway.
-Despite its simplicity, the model captured key features of the ATRA induced differentiation such as
-sustained MAPK activation, and bistability with respect to ATRA exposure. We also reported a new ATRA-inducible component of the signalsome, Vav1.
+In this study, we presented an effective model of ATRA-inducible differentiation of HL-60 cells.
+The model consisted of three modules: a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (Trigger) signals;
+a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
+a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
+The model described the transcription and translation of genes in each module, and signaling events in each module in a growing population of HL-60 cells.
+Model parameters were largely taken from literature, however, unknown coefficients were estimated from protein measurements in HL-60 cells following ATRA exposure.
+Despite its simplicity, the effective model captured key features of the ATRA induced differentiation such as
+sustained MAPK activation, and bistability with respect to ATRA exposure.
+The model also described the expression of upstream transcription factors which regulated the expression of differentiation markers.
+Lastly, analysis of the response of the model to perturbations identified PU.1 and AP1 as master regulators of ATRA-induced differentiation.
+We also reported a new ATRA-inducible component of the signalsome, Vav1.
 Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
-The Vav1/Raf association correlated with Raf activity, was ATRA-inducible and decreased after treatment with GW5074.
-The presence of Vav1 in Raf/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
-Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
-of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
-While its function in the signalsome is unclear, Vav1 has been shown to associate with a Cbl-Slp76-CD38 complex in an ATRA-dependent manner;
-furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevented
-ATRA-induced MAPK activation \cite{Shen2009}. Thus, interaction of Cbl-Slp76-Vav1 and CD38 appears to be required for transmission of the ATRA signal by the
-signalsome.
+The Vav1/Raf association correlated with Raf activity, was ATRA-inducible and decreased after treatment with the Raf inhibitor GW5074.
+
+Man-made engineered systems, or naturally occurring cell fate decision programs incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
+One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
+This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
+The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
+This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
+Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
+Reinforcing feedback and bistability has also been implicated in hematopoietic cell fate determination.
+Laslo et al showed that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBPa), encodes a bistable switch that
+results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
+The current model contained the Gfi-1 and Egr-1/2 agonistic switch.
+Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
+Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
+Thus, while reinforcing feedback is often undesirable in human engineered systems, it is at the core of a diverse variety of
+cell fate programs and other networks important to human health.
 
 Analysis of the signal integration and phenotype modules suggested that PU.1 and AP1 regulated distinct phenotypic axes following ATRA exposure.
-In particular, deletion of PU.1 and AP1 was predicted to lead to largest shift in the system state.
+The PU.1 transcription factor, a member of the ets transcription factor family, is a well known regulatory protein in granulocyte and monocyte development \cite{Friedman2007}.
+The relative level of PU.1 and C/EBPa regulates macrophage versus neutrophil cell fate decisions in granulocytic macrophage progenitor cells \cite{Dahl2003}.
 
@@
 Such molecular mechanisms could underly aspects of cellular memory associated to consecutive ATRA treatments.
-
-Several engineered, or naturally occurring systems involved in cell fate decisions incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
-One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
-This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
-The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
-This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
-Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
-Reinforcing feedback and bistability has also been implicated in hematopoietic stem cell fate determination.
-Laslo et al showed that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBPa), encodes a bistable switch that
-results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
-Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
-Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
-Thus, while positive feedback is often undesirable in human engineered systems, it may be at the core of a diverse variety of
-cell fate programs and other networks important to human health.
 

### 2017-04-26 Paper_v2.tex
Working on discussion
Working on text

@@
 {\par \centering \large{Ryan Tasseff, Holly A.
-Jensen, Wei Dai, Katherine Rogers, Adithya Sagar, Rodica P.
-Bunaciu$^{\dag}$, Johanna Congleton$^{\dag}$, Andrew Yen$^{\dag}$, and Jeffrey D.
+Jensen, Johanna Congleton$^{\dag}$, Wei Dai, Katherine Rogers, Adithya Sagar, Rodica P.
+Bunaciu$^{\dag}$, Andrew Yen$^{\dag}$, and Jeffrey D.
 Varner$^{*}$}}
@@
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Frontiers~in~Systems~Biology}}
+{\par \centering \textbf{To be submitted:}~\emph{?}}
 \vspace{0.5in}
@@
 The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
-which then in turn activated the phenotypic program. In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex,
+which then in turn activated the phenotypic program. In particular, Trigger (a surrogate for the RAR$\alpha$/RXR transcriptional complex)
 regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
@@
 We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
-We estimated the parameters which appeared in the control laws regulating these programs
+We estimated the parameters for the signal initiation, and phenotype modules
 from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
-However, the bulk of the remaining model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
+However, the bulk of the model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
 The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
@@
 We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
-The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and wash robust to uncertainty in the measured data (Fig.
+The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and was robust to uncertainty in the measured data (Fig.
 \ref{fig:model-cell-cycle-simulations}B, gray).
@@
 The PU.1 and AP1 proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
-We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
+We conducted pairwise knockout simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
 The difference between the system state with and without the gene knockouts (encoded as a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
-All simulations were conducted using the best fit parameter set.
+A panel of twenty parameter sets was sampled, and the average displacement matrix was decomposed.
 The first four modes described $>$99\% of the gene knockout variance,
@@
 
-Man-made engineered systems, or naturally occurring cell fate decision programs incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
+Naturally occurring cell fate decisions often incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
 One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
@@
 results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
-The current model contained the Gfi-1 and Egr-1/2 agonistic switch.
+The current model contained the Gfi-1 and Egr-1/2 agonistic switch; however, its significance was unclear for HL-60 cells.
+The expression of Gfi-1, Egr-1/2, C/EBPa and PU.1 was not consistent with the canonical lineage pattern expected from literature.
+For example, Egr-1/2 expression (associated with a macrophage lineage) increased, while Gfi-1 expression (associated with a neutrophil lineage)
+remained constant following ATRA exposure.
+Literature evidence in nonmalignant myelomonocytic fate selection has shown that Gfi-1 and EGR-1/2 promote granulocytic and monocytic differentiation, respectively \cite{Laslo2006}.
+Thus, HL-60 cells, which are a less mature cancer cell line, exhibited a non-canonical expression pattern.
 Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.

### 2017-05-03 Paper_v2.tex
Are we there yet?

@@
 Analysis of the signal integration and phenotype modules suggested that PU.1 and AP1 regulated distinct phenotypic axes following ATRA exposure.
-The PU.1 transcription factor, a member of the ets transcription factor family, is a well known regulatory protein in granulocyte and monocyte development \cite{Friedman2007}.
-The relative level of PU.1 and C/EBPa regulates macrophage versus neutrophil cell fate decisions in granulocytic macrophage progenitor cells \cite{Dahl2003}.
+PU.1, a member of the ets transcription factor family, is a well known regulatory protein in granulocyte and monocyte development \cite{Friedman2007}.
+The relative level of PU.1 and C/EBPa is thought to regulate macrophage versus neutrophil cell fate decisions in granulocytic macrophage progenitor cells \cite{Dahl2003}.
+However, in previous studies we showed that PU.1 has additional non-canonical function as it is highly expressed following the addition of ATRA to HL-60 cells. 
+Analysis of the model in this study suggested that the C/EBPa dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes for the P47Phox protein, was the most sensitive PU.1 connection.
+P47Phox, also known as neutrophil cytosol factor 1, is one of four cytosolic subunits of the multi-protein NADPH oxidase complex found in neutrophils \cite{El-Benna:2009aa}.
+This enzyme is responsible for reactive oxygen species (ROS) production, a key component of the anti-microbial function of neutrophils.
+PU.1 also acted in concert with the activator protein 1 (AP-1) transcription factor to regulate p21 expression, a second phenotypic outcome of ATRA-induced differentiation.
+The AP-1 transcription factor regulates gene expression in response to a variety of stimuli.
+However, in the context of ATRA-induced differentiation, AP1 have been shown to
 
@@
 
-	$R_{1}$ & RNA polymerase abundance & 75,000 & copies/cell & \cite{Jackson:2000aa,Zhao:2014aa} \\
+	$R_{1}$ & RNA polymerase abundance & 85,000 & copies/cell & \cite{Jackson:2000aa,Zhao:2014aa} \\
 	$R_{2}$ & Ribosome abundance & 1 x 10$^{6}$ & copies/cell & \cite{Freitas2004} \\
 	$G_{i}$ & Characteristic gene abundance & 2 & copies/cell & this study\\
-	$K_{X}$ & Saturation constant transcription & 4,600 & copies/cell & this study\\
-	$K_{T}$ & Saturation constant translation & 100,000 & copies/cell & this study\\
+	$K_{X}$ & Saturation constant transcription & 600 & copies/cell & this study\\
+	$K_{T}$ & Saturation constant translation & 95,000 & copies/cell & this study\\
 	& & & & \\
-	$t_{1/2,m}$ & characteristic mRNA half-life (transcription factor) & 2 & hr & \cite{Yang:2003aa} \\
+	$t_{1/2,m}$ & characteristic mRNA half-life (transcription factor) & 2-4 & hr & \cite{Yang:2003aa} \\
 	$t_{1/2,p}$ & characteristic protein half-life & 10 & hr & \cite{Doherty:2009aa} \\
@@
 	& & & & \\
-	$e_{T}$ & elongation rate RNA polymerase & 6 & nt/s & \cite{Darzacq:2007aa} \\
+	$e_{T}$ & elongation rate RNA polymerase & 50-100 & nt/s & \cite{Sehgal:1976aa,Darzacq:2007aa,Kos:2010aa,Darnell:2013aa} \\
 	$e_{X}$ & elongation rate Ribosome & 5 & aa/s & \cite{Bostrom:1986aa} \\

### 2017-05-04 Paper_v2.tex
Some more changes in results and discussion

@@
 which then in turn activated the phenotypic program. In particular, Trigger (a surrogate for the RAR$\alpha$/RXR transcriptional complex)
-regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
-In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
+regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and Egr-1.
+In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or p47Phox.
 We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
@@
 
-The PU.1 and AP1 proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
-We conducted pairwise knockout simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
-The difference between the system state with and without the gene knockouts (encoded as a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
-A panel of twenty parameter sets was sampled, and the average displacement matrix was decomposed.
-The first four modes described $>$99\% of the gene knockout variance,
-with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent Gfi1 and C/EBPa (Fig. \ref{fig:model-response-ko}A).
-To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
-SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
-The most sensitive interactions for the PU.1 protein involved the C/EBPa-dependent regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
-On the other other, the most sensitive connections for AP1 involved the C/EBPa-dependent regulation of p21 expression, and the mutual activation of PU.1 and AP1 expression.
-Taken together, these results suggested that the PU.1 and AP1 proteins acted as important self-reinforcing regulators for both the signal integration and phenotype modules.
-The analysis suggested that the PU.1 signaling axis promoted the formation of the neutrophil NADPH oxidase (through p47Phox), while AP1 was responsible for cell cycle arrest (through p21).
-However, this analysis did not give insight into which upstream components of the signal initiation module were important.
+The Gfi-1 and PPARg proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
+We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
+The difference between the system state with and without the gene knockouts (encoded as a normalized state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
+A panel of ten parameter sets was sampled, and the average normalized displacement matrix was decomposed.
+The first six modes (approximately 36\% of the total) described $>$95\% of the gene knockout variance,
+with the most important components of these modes being the Gfi-1 and PPARg proteins, and to a lesser extent PU.1, C/EBPa and and AP1 (Fig. \ref{fig:model-response-ko}A).
+To better understand which protein-DNA connections were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
+SVD decomposition of the normalized state displacement matrix assembled from the pairwise connection deletions, suggested the first six modes (approximately 26\% of the total) accounted for $>$90\% of the variance.
+Globally, the most sensitive interactions controlled p21 and p47Phox expression, markers for cell-cycle arrest and reactive oxygen formation phenotypic axes activated following ATRA addition (Fig.
+\ref{fig:model-response-ko}B).
+Analysis of the modes suggested the action of PPARg, Gfi-1 and C/EBPa were consistently important over multiple target genes.
+The connection knockout analysis also revealed robustness in the network.
+For example, no pair of deletions qualitatively changed the expression of regulators such as PU.1, Oct1, Oct4 or PPARg.
+Thus, the expression of these species was robust to disturbance in the connectivity.
+To better understand the combined influence of the PPARg and Gfi-1 deletions, we computed the fold change in the protein levels in the
+single ($\Delta$Gfi-1 or $\Delta$PPARg) and double ($\Delta$Gfi-1 and $\Delta$PPARg) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
+Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBPa expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).
+On the other hand, deletion of PPARg led to $>$8 fold downregulation of CD38, p21, IRF1 and OCT1 (Fig. \ref{fig:model-response-ko-simulations},red).
+Both knockouts slightly increased E2F expression, but neither influenced the expression of p47Phox.
+The double mutant was qualitatively similar to the combined behavior of the two single mutant cases.
+Taken together, Gfi-1 and PPARg controlled the cell-cycle arrest and receptor signaling axes, with PPARg regulating CD38, IRF1 and p21 expression
+while Gfi-1 controlled CD11b expression. These simulations suggested deletion of PPARg and Gfi-1 would not interfere with reactive oxygen formation,
+but would limit the ability of HL-60 cells to arrest.
+However, this analysis did not give insight into which components upstream of the signal initiation module were important.
 Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.
+
+% Taken together, these results suggested that the PU.1 and AP1 proteins acted as important self-reinforcing regulators for both the signal integration and phenotype modules.
+% The analysis suggested that the PU.1 signaling axis promoted the formation of the neutrophil NADPH oxidase (through p47Phox), while AP1 was responsible for cell cycle arrest (through p21).
 
@@
 Treatment with both ATRA and GW5074 statistically significantly reduced ROS induction (22 $\pm$ 0.6\%) compared to ATRA alone.
-Interestingly, Western blot analysis did not detect a GW5074 effect on ATRA-induced expression of p47phox,
+Interestingly, Western blot analysis did not detect a GW5074 effect on ATRA-induced expression of p47Phox,
 a required upstream component of the ROS response  (Fig. \ref{fig:ip-data}F, bottom).
-Thus, the inhibitory effect of GW5074 on inducible ROS might occur downstream of p47phox expression.
+Thus, the inhibitory effect of GW5074 on inducible ROS might occur downstream of p47Phox expression.
 However, the ROS producing complex is MAPK dependent,
@@
 remained constant following ATRA exposure.
-Literature evidence in nonmalignant myelomonocytic fate selection has shown that Gfi-1 and EGR-1/2 promote granulocytic and monocytic differentiation, respectively \cite{Laslo2006}.
+Literature evidence in nonmalignant myelomonocytic fate selection has shown that Gfi-1 and Egr-1/2 promote granulocytic and monocytic differentiation, respectively \cite{Laslo2006}.
 Thus, HL-60 cells, which are a less mature cancer cell line, exhibited a non-canonical expression pattern.
@@
 
-Analysis of the signal integration and phenotype modules suggested that PU.1 and AP1 regulated distinct phenotypic axes following ATRA exposure.
+Analysis of the signal integration and phenotype modules suggested that Gfi-1 and PPARg regulated distinct phenotypic axes following ATRA exposure.
+The most significantly unregulated protein in the $\Delta$Gfi-1 mutant was the PU.1 protein.
 PU.1, a member of the ets transcription factor family, is a well known regulatory protein in granulocyte and monocyte development \cite{Friedman2007}.
@@
 However, in previous studies we showed that PU.1 has additional non-canonical function as it is highly expressed following the addition of ATRA to HL-60 cells.
-Analysis of the model in this study suggested that the C/EBPa dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes for the P47Phox protein, was the most sensitive PU.1 connection.
-P47Phox, also known as neutrophil cytosol factor 1, is one of four cytosolic subunits of the multi-protein NADPH oxidase complex found in neutrophils \cite{El-Benna:2009aa}.
+Analysis of the model in this study suggested that the C/EBPa dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes for the p47Phox protein, was the most sensitive PU.1 connection.
+p47Phox, also known as neutrophil cytosol factor 1, is one of four cytosolic subunits of the multi-protein NADPH oxidase complex found in neutrophils \cite{El-Benna:2009aa}.
 This enzyme is responsible for reactive oxygen species (ROS) production, a key component of the anti-microbial function of neutrophils.
@@
 The AP-1 transcription factor regulates gene expression in response to a variety of stimuli.
-However, in the context of ATRA-induced differentiation, AP1 have been shown to
+However, in the context of ATRA-induced differentiation, AP1 has been shown to be inhibited in
 
@@
  	&  + & AhR & \cite{Bunaciu2013}\\
- 	&  + & EGR1 & \cite{Balmer2002}\\
+ 	&  + & Egr-1 & \cite{Balmer2002}\\
 \hline
@@
  &  - & E2F & \cite{Altiok1997}\\
- &  - & EGR1 & \cite{Fei2011}\\
+ &  - & Egr-1 & \cite{Fei2011}\\
  &  + & CD38 & \cite{Song2012}\\
@@
  &  + & p21 & \cite{Han2004}\\
- &  - & p47phox & \cite{Von-Knethen2002}\\
+ &  - & p47Phox & \cite{Von-Knethen2002}\\
 \hline
@@
 &  + & AP-1 & \cite{Steidl2006}\\
-&  + & EGR1 & \cite{Laslo2006}\\
+&  + & Egr-1 & \cite{Laslo2006}\\
 &  + & CD11b & \cite{Pahl1993}\\
 &  + & p21 & \cite{Yuki2013}\\
-&  + & p47phox & \cite{Li1999}\\
+&  + & p47Phox & \cite{Li1999}\\
 \hline
@@
  &  - & E2F & \cite{Duan2003}\\
- &  - & EGR1 & \cite{Laslo2006}\\
+ &  - & Egr-1 & \cite{Laslo2006}\\
  &  - & p21 & \cite{Duan2003}\\
@@
 \hline
-EGR1 &  + & PPAR$\gamma$ & \cite{Fu2002}\\
+Egr-1 &  + & PPAR$\gamma$ & \cite{Fu2002}\\
 &  - & Gfi-1 & \cite{Mak2011}\\
@@
 The percentage at the top of each column describes the fraction of the variance in the system state captured by the node combinations in the rows.
-B: Singular value decomposition of the system response ($l^{2}$-norm between the perturbed and nominal state) following the pairwise removal of connections from the PU.1 and AP1 nodes.}\label{fig:model-response-ko}
+B: Singular value decomposition of the system response ($l^{2}$-norm between the perturbed and nominal state) following the pairwise removal of connections.}\label{fig:model-response-ko}
+\end{figure}
+
+\begin{figure}[!t]\centering
+\includegraphics[width=0.85\textwidth]{./figs/Fig-7-GeneKO-Simulations.pdf}
+\caption{Robustness of the HL-60 differentiation program following exposure to 1$\mu$M ATRA at t = 0 hr.
+Protein fold change at t = 48 hr (rows) in single and double knock-out mutants (columns) relative to wild-type HL-60 cells.
+The responses were grouped into $>$2,4 and 8 fold changes.
+The best fit parameter set was used to calculate the protein fold change.}\label{fig:model-response-ko-simulations}
 \end{figure}
@@
 \begin{figure}[!t]
-\includegraphics[width=0.9\textwidth]{./figs/Fig-7-IP-ExperimentalStudies.pdf}
+\includegraphics[width=0.9\textwidth]{./figs/Fig-8-IP-ExperimentalStudies.pdf}
 \caption{Investigation of a panel of possible Raf interaction partners in the presence and absence of ATRA.

### 2017-05-05 Paper_v2.tex
First draft

@@
 To better understand the combined influence of the PPARg and Gfi-1 deletions, we computed the fold change in the protein levels in the
-single ($\Delta$Gfi-1 or $\Delta$PPARg) and double ($\Delta$Gfi-1 and $\Delta$PPARg) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
+single (Gfi-1$^{-/-}$ or $\Delta$PPARg) and double (Gfi-1$^{-/-}$ and $\Delta$PPARg) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
 Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBPa expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).
@@
 The model described the transcription and translation of genes in each module, and signaling events in each module in a growing population of HL-60 cells.
-Model parameters were largely taken from literature, however, unknown coefficients were estimated from protein measurements in HL-60 cells following ATRA exposure.
+Model parameters were taken from literature, however, unknown coefficients
+that appear in the promoter logic models were estimated from protein measurements in HL-60 cells following ATRA exposure.
 Despite its simplicity, the effective model captured key features of the ATRA induced differentiation such as
@@
 The model also described the expression of upstream transcription factors which regulated the expression of differentiation markers.
-Lastly, analysis of the response of the model to perturbations identified PU.1 and AP1 as master regulators of ATRA-induced differentiation.
+Lastly, analysis of the response of the model to perturbations identified Gfi-1 and PPARg as master regulators of ATRA-induced differentiation.
 We also reported a new ATRA-inducible component of the signalsome, Vav1.
@@
 
-Analysis of the signal integration and phenotype modules suggested that Gfi-1 and PPARg regulated distinct phenotypic axes following ATRA exposure.
-The most significantly unregulated protein in the $\Delta$Gfi-1 mutant was the PU.1 protein.
-PU.1, a member of the ets transcription factor family, is a well known regulatory protein in granulocyte and monocyte development \cite{Friedman2007}.
-The relative level of PU.1 and C/EBPa is thought to regulate macrophage versus neutrophil cell fate decisions in granulocytic macrophage progenitor cells \cite{Dahl2003}.
-However, in previous studies we showed that PU.1 has additional non-canonical function as it is highly expressed following the addition of ATRA to HL-60 cells.
-Analysis of the model in this study suggested that the C/EBPa dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes for the p47Phox protein, was the most sensitive PU.1 connection.
+%However, in previous studies we have shown that PU.1 has additional non-canonical function, as it is highly expressed following the addition of ATRA to HL-60 cells.
+%
+Analysis of the signal integration and phenotype modules suggested that Gfi-1 and PPARg were required for ATRA-induced differentiation in HL-60 cells.
+Model analysis showed that PU.1, Egr-1 and C/EBPa expression increased in Gfi-1$^{-/-}$ mutants, where PU.1 expression was upregulated by greater than 8-fold.
+PU.1, a member of the \textit{ets} transcription factor family, is a well known regulator of granulocyte and monocyte development \cite{Friedman2007}.
+The relative level of PU.1 and C/EBPa is thought to control macrophage versus neutrophil cell fate decisions in granulocytic macrophage progenitor cells \cite{Dahl2003}.
+Simulations suggested that combined Gfi-1 + PPARg deletion crippled the ability of HL-60 cells to undergo neutrophilic differentiation following ATRA exposure.
+p21 expression decreased significantly, suggesting Gfi-1$^{-/-}$ + PPARg$^{-/-}$ mutants were less likely to G0-arrest following ATRA exposure.
+The expression of other neutrophilic markers, such as CD38, also decreased in Gfi-1$^{-/-}$ + PPARg$^{-/-}$ cells.
+On the other hand, the expression of reactive oxygen metabolic markers, or other important transcription factors such as OCT4 did not change.
+For example, model analysis suggested that the C/EBPa dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes the p47Phox protein, was the most sensitive PU.1 connection; deletion of this connection removed the ability of the system to express p47Phox.
 p47Phox, also known as neutrophil cytosol factor 1, is one of four cytosolic subunits of the multi-protein NADPH oxidase complex found in neutrophils \cite{El-Benna:2009aa}.
 This enzyme is responsible for reactive oxygen species (ROS) production, a key component of the anti-microbial function of neutrophils.
-PU.1 also acted in concert with the activator protein 1 (AP-1) transcription factor to regulate p21 expression, a second phenotypic outcome of ATRA-induced differentiation.
-The AP-1 transcription factor regulates gene expression in response to a variety of stimuli.
-However, in the context of ATRA-induced differentiation, AP1 has been shown to be inhibited in
+However, Gfi-1 deletion did not increase p47Phox expression, nor did the deletion of PPARg.
+This suggested that while p47Phox expression required C/EBPa and PU.1, it was saturated with respect these proteins, and simultaneously not sensitive to PPARg abundance.
+Together, Gfi-1$^{-/-}$ + PPARg$^{-/-}$ cells were predicted to exhibit
+some aspects of the ATRA response, but not other critical features such as cell cycle arrest.
+Hock et al showed that Gfi-1$^{-/-}$ mice lacked normal neutrophils, and were highly sensitive to bacterial infection \cite{Hock:2003aa}.
+Thus, the model analysis was consistent with this study.
+However, other predictions concerning the behavior of the Gfi-1$^{-/-}$ + PPARg$^{-/-}$ mutants have yet to be verified experimentally.
 
@@
 Although the active state was self-sustaining, the inactive state demonstrated considerable robustness to perturbation.
-For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on ppERK levels
-over the entire ensemble.
+For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on ppERK levels over the entire ensemble.
 CD38 expression correlated with the ppERK, suggesting its involvement in the signaling complex.
@@
 
-There were several issues that can be explored further with the effective ATRA differentiation model.
-First, there was likely missing connectivity in the effective differentiation circuit.
-Decreasing BLR1 expression with simultaneously sustained cRaf-pS261 activation was not captured by the current network architecture.
-This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRaf-pS261 abundance.
-We could model this by separating signalsome formation into an inactive precursor pool that is transformed to a long-lived activated signalsome by MAPK activation.
-We should also explore adding additional downstream biological modules to this skeleton model, for example the upregulation of reactive oxygen markers such as p47Phox or
-cell cycle arrest components to capture the switch from an actively proliferating population to a population in G0-arrest.
-Next, the choice of max/min integration rules or the particular form of the transfer functions could also be explored.
-Integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
-Alternative integration rules might have different properties which could influence model identification or performance.
-For example, a mean integration rule would be differentiable, allowing derivative-based optimization approaches to be used.
-The form of the transfer function could also be explored. We choose hill-like functions because of their
-prominence in the systems and synthetic biology community. However, many other transfer functions are possible.
+% There were several issues that can be explored further with the effective ATRA differentiation model.
+% First, there was likely missing connectivity in the effective differentiation circuit.
+% Decreasing BLR1 expression with simultaneously sustained cRaf-pS261 activation was not captured by the current network architecture.
+% This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRaf-pS261 abundance.
+% We could model this by separating signalsome formation into an inactive precursor pool that is transformed to a long-lived activated signalsome by MAPK activation.
+% We should also explore adding additional downstream biological modules to this skeleton model, for example the upregulation of reactive oxygen markers such as p47Phox or
+% cell cycle arrest components to capture the switch from an actively proliferating population to a population in G0-arrest.
+% Next, the choice of max/min integration rules or the particular form of the transfer functions could also be explored.
+% Integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
+% Alternative integration rules might have different properties which could influence model identification or performance.
+% For example, a mean integration rule would be differentiable, allowing derivative-based optimization approaches to be used.
+% The form of the transfer function could also be explored. We choose hill-like functions because of their
+% prominence in the systems and synthetic biology community. However, many other transfer functions are possible.
 

### 2017-05-05 Paper_v2.tex
First draft

@@
 To better understand the combined influence of the PPARg and Gfi-1 deletions, we computed the fold change in the protein levels in the
-single (Gfi-1$^{-/-}$ or $\Delta$PPARg) and double (Gfi-1$^{-/-}$ and $\Delta$PPARg) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
+single (Gfi-1$^{-/-}$ or PPARg$^{-/-}$) and double (Gfi-1$^{-/-}$ and PPARg$^{-/-}$) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
 Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBPa expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).

### 2017-05-05 Paper_v2.tex
First draft

@@
 Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the signalsome.
-A recent computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
+A previous computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
 to explain ATRA-induced sustained MAPK activation, and the expression of a limited number of functional differentiation markers \cite{Tasseff2011}.
@@
 namely positive feedback between an ATRA-inducible signalsome complex and MAPK activation.
-The ATRA responsive signalsome-MAPK circuit was then used to drive a downstream gene expression program which encoded for the expression of functional differentiation markers.
+The ATRA responsive signalsome-MAPK circuit was then used to drive a downstream gene expression program which encoded for the expression of intermediate and functional differentiation markers.
 The effective model used a novel framework which integrated logical rules with kinetic modeling to describe gene expression and protein regulation,
@@
 Model simulations were also consistent with measurements of the influence of MAPK inhibitors, and the failure of BLR1 knockout cells to differentiate when exposed to ATRA.
+In addition, the expression of intermediate and phenotypic differentiation markers as also captured following ATRA exposure.
 Lastly, we showed by through immunoprecipitation studies, that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the signalsome complex.

### 2017-05-08 Paper_v2.tex
David's chnages and sequence table

@@
 This suggested that we captured the logic leading to the onset of differentiation, but failed to describe program shutdown.
+Much of the focus in the literature has been on understanding the initiation of differentiation, with little attention paid to understanding how a program is
+terminated. This is a potential new direction that could be explored.
 Next, we tested the response of the signal initiation module to different ATRA dosages.
@@
 
-The Gfi-1 and PPARg proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
+The Gfi-1 and PPAR$\gamma$ proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
 We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
@@
 The first six modes (approximately 36\% of the total) described $>$95\% of the gene knockout variance,
-with the most important components of these modes being the Gfi-1 and PPARg proteins, and to a lesser extent PU.1, C/EBPa and and AP1 (Fig. \ref{fig:model-response-ko}A).
+with the most important components of these modes being the Gfi-1 and PPAR$\gamma$ proteins, and to a lesser extent PU.1, C/EBP$\alpha$ and and AP1 (Fig. \ref{fig:model-response-ko}A).
 To better understand which protein-DNA connections were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
-SVD decomposition of the normalized state displacement matrix assembled from the pairwise connection deletions, suggested the first six modes (approximately 26\% of the total) accounted for $>$90\% of the variance.
+Singular value decomposition of the normalized state displacement matrix assembled from the pairwise connection deletions, suggested the first six modes (approximately 26\% of the total) accounted for $>$90\% of the variance.
 Globally, the most sensitive interactions controlled p21 and p47Phox expression, markers for cell-cycle arrest and reactive oxygen formation phenotypic axes activated following ATRA addition (Fig.
 \ref{fig:model-response-ko}B).
-Analysis of the modes suggested the action of PPARg, Gfi-1 and C/EBPa were consistently important over multiple target genes.
+Analysis of the modes suggested the action of PPAR$\gamma$, Gfi-1 and C/EBP$\alpha$ were consistently important over multiple target genes.
 The connection knockout analysis also revealed robustness in the network.
-For example, no pair of deletions qualitatively changed the expression of regulators such as PU.1, Oct1, Oct4 or PPARg.
+For example, no pair of deletions qualitatively changed the expression of regulators such as PU.1, Oct1, Oct4 or PPAR$\gamma$.
 Thus, the expression of these species was robust to disturbance in the connectivity.
-To better understand the combined influence of the PPARg and Gfi-1 deletions, we computed the fold change in the protein levels in the
-single (Gfi-1$^{-/-}$ or PPARg$^{-/-}$) and double (Gfi-1$^{-/-}$ and PPARg$^{-/-}$) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
-Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBPa expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).
-On the other hand, deletion of PPARg led to $>$8 fold downregulation of CD38, p21, IRF1 and OCT1 (Fig. \ref{fig:model-response-ko-simulations},red).
+To better understand the combined influence of the PPAR$\gamma$ and Gfi-1 deletions, we computed the fold change in the protein levels in the
+single (Gfi-1$^{-/-}$ or PPAR$\gamma$$^{-/-}$) and double (Gfi-1$^{-/-}$ and PPAR$\gamma$$^{-/-}$) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
+Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBP$\alpha$ expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).
+On the other hand, deletion of PPAR$\gamma$ led to $>$8 fold downregulation of CD38, p21, IRF1 and OCT1 (Fig. \ref{fig:model-response-ko-simulations},red).
 Both knockouts slightly increased E2F expression, but neither influenced the expression of p47Phox.
 The double mutant was qualitatively similar to the combined behavior of the two single mutant cases.
-Taken together, Gfi-1 and PPARg controlled the cell-cycle arrest and receptor signaling axes, with PPARg regulating CD38, IRF1 and p21 expression
-while Gfi-1 controlled CD11b expression. These simulations suggested deletion of PPARg and Gfi-1 would not interfere with reactive oxygen formation,
+Taken together, Gfi-1 and PPAR$\gamma$ controlled the cell-cycle arrest and receptor signaling axes, with PPAR$\gamma$ regulating CD38, IRF1 and p21 expression
+while Gfi-1 controlled CD11b expression. These simulations suggested deletion of PPAR$\gamma$ and Gfi-1 would not interfere with reactive oxygen formation,
 but would limit the ability of HL-60 cells to arrest.
@@
 Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf, suggesting a direct physical interaction was possible.
-However, only the associations between Raf and Vav1 and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
+However, only the associations between Raf and Vav1, and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
 Furthermore, the Vav1 and Src associations were correlated with pS621 Raf abundance in the precipitate.
@@
 The model also described the expression of upstream transcription factors which regulated the expression of differentiation markers.
-Lastly, analysis of the response of the model to perturbations identified Gfi-1 and PPARg as master regulators of ATRA-induced differentiation.
+Lastly, analysis of the response of the model to perturbations identified Gfi-1 and PPAR$\gamma$ as master regulators of ATRA-induced differentiation.
 We also reported a new ATRA-inducible component of the signalsome, Vav1.
@@
 Reinforcing feedback and bistability has also been implicated in hematopoietic cell fate determination.
-Laslo et al showed that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBPa), encodes a bistable switch that
-results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
+Laslo et al showed in nonmalignant myelomonocytic cells that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBP$\alpha$), encode a bistable switch that
+results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBP$\alpha$ expression \cite{Laslo2006}.
 The current model contained the Gfi-1 and Egr-1/2 agonistic switch; however, its significance was unclear for HL-60 cells.
-The expression of Gfi-1, Egr-1/2, C/EBPa and PU.1 was not consistent with the canonical lineage pattern expected from literature.
+The expression of Gfi-1, Egr-1/2, C/EBP$\alpha$ and PU.1 was not consistent with the canonical lineage pattern expected from literature.
 For example, Egr-1/2 expression (associated with a macrophage lineage) increased, while Gfi-1 expression (associated with a neutrophil lineage)
 remained constant following ATRA exposure.
-Literature evidence in nonmalignant myelomonocytic fate selection has shown that Gfi-1 and Egr-1/2 promote granulocytic and monocytic differentiation, respectively \cite{Laslo2006}.
 Thus, HL-60 cells, which are a less mature cancer cell line, exhibited a non-canonical expression pattern.
@@
 %
-Analysis of the signal integration and phenotype modules suggested that Gfi-1 and PPARg were required for ATRA-induced differentiation in HL-60 cells.
-Model analysis showed that PU.1, Egr-1 and C/EBPa expression increased in Gfi-1$^{-/-}$ mutants, where PU.1 expression was upregulated by greater than 8-fold.
+Analysis of the signal integration and phenotype modules suggested that Gfi-1 and PPAR$\gamma$ were required for ATRA-induced differentiation in HL-60 cells.
+Model analysis showed that PU.1, Egr-1 and C/EBP$\alpha$ expression increased in Gfi-1$^{-/-}$ mutants, where PU.1 expression was upregulated by greater than 8-fold.
 PU.1, a member of the \textit{ets} transcription factor family, is a well known regulator of granulocyte and monocyte development \cite{Friedman2007}.
-The relative level of PU.1 and C/EBPa is thought to control macrophage versus neutrophil cell fate decisions in granulocytic macrophage progenitor cells \cite{Dahl2003}.
-Simulations suggested that combined Gfi-1 + PPARg deletion crippled the ability of HL-60 cells to undergo neutrophilic differentiation following ATRA exposure.
-p21 expression decreased significantly, suggesting Gfi-1$^{-/-}$ + PPARg$^{-/-}$ mutants were less likely to G0-arrest following ATRA exposure.
-The expression of other neutrophilic markers, such as CD38, also decreased in Gfi-1$^{-/-}$ + PPARg$^{-/-}$ cells.
+The relative level of PU.1 and C/EBP$\alpha$ is thought to control macrophage versus neutrophil cell fate decisions in granulocytic macrophage progenitor cells \cite{Dahl2003}.
+Simulations suggested that combined Gfi-1 + PPAR$\gamma$ deletion crippled the ability of HL-60 cells to undergo neutrophilic differentiation following ATRA exposure.
+p21 expression decreased significantly, suggesting Gfi-1$^{-/-}$ + PPAR$\gamma$$^{-/-}$ mutants were less likely to G0-arrest following ATRA exposure.
+The expression of other neutrophilic markers, such as CD38, also decreased in Gfi-1$^{-/-}$ + PPAR$\gamma$$^{-/-}$ cells.
 On the other hand, the expression of reactive oxygen metabolic markers, or other important transcription factors such as OCT4 did not change.
-For example, model analysis suggested that the C/EBPa dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes the p47Phox protein, was the most sensitive PU.1 connection; deletion of this connection removed the ability of the system to express p47Phox.
+For example, model analysis suggested that the C/EBP$\alpha$ dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes the p47Phox protein, was the most sensitive PU.1 connection; deletion of this connection removed the ability of the system to express p47Phox.
 p47Phox, also known as neutrophil cytosol factor 1, is one of four cytosolic subunits of the multi-protein NADPH oxidase complex found in neutrophils \cite{El-Benna:2009aa}.
 This enzyme is responsible for reactive oxygen species (ROS) production, a key component of the anti-microbial function of neutrophils.
-However, Gfi-1 deletion did not increase p47Phox expression, nor did the deletion of PPARg.
-This suggested that while p47Phox expression required C/EBPa and PU.1, it was saturated with respect these proteins, and simultaneously not sensitive to PPARg abundance.
-Together, Gfi-1$^{-/-}$ + PPARg$^{-/-}$ cells were predicted to exhibit
+While p47Phox expression required C/EBP$\alpha$ and PU.1, neither Gfi-1 nor PPAR$\gamma$ deletion increased expression.
+This suggested that p47Phox expression was saturated with respect to C/EBP$\alpha$ and PU.1, and simultaneously not sensitive to PPAR$\gamma$ abundance.
+Taken together, Gfi-1$^{-/-}$ + PPAR$\gamma$$^{-/-}$ cells were predicted to exhibit
 some aspects of the ATRA response, but not other critical features such as cell cycle arrest.
@@
 Thus, the model analysis was consistent with this study.
-However, other predictions concerning the behavior of the Gfi-1$^{-/-}$ + PPARg$^{-/-}$ mutants have yet to be verified experimentally.
+However, other predictions concerning the behavior of the Gfi-1$^{-/-}$ + PPAR$\gamma$$^{-/-}$ mutants have yet to be verified experimentally.
 
@@
 Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
-The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.
 Lastly, the $\lambda_{j}$ term denotes the constitutive rate of expression of gene $j$.
@@
 The numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
-The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using hill like functions), while
+The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using Hill like functions), while
 the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
@@
 However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
-To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger (shown generally as Eq \eqref{eqn-signaling-balances}).
+To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger.
 The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
@@
 \end{equation}
-The first term in Eqn. \eqref{eqn:objective-function} quantified the initial \textit{scale} error, directly before the addition of ATRA.
+The first term in Eq. \eqref{eqn:objective-function} quantified the initial \textit{scale} error, directly before the addition of ATRA.
 In this case, $\mathcal{M}_{j}\left(t_{-}\right)$ (the approximate concentration of protein $j$ before the addition of ATRA) was estimated from literature.
@@
 
+\begin{center}
+% \renewcommand{\arraystretch}{0.8}
+\tablefirsthead{
+\hline
+\rowcolor[gray]{0.9} Gene Name & Gene (bp) & RNA (bp) & Protein (AA) & Gene ID & Protein ID \\}
+\topcaption{Sequence lengths from NCBI RefSeq database was used in the signal integration and phenotype modules \cite{OLeary:2016aa}.
+The RNA sequence length used represents the total distance of transcription, and assume to be equal to the gene length. }\label{tbl:TF-sequences}
+\begin{scriptsize}
+\begin{supertabular}{r|l|l|l|l|l}
+\hline
+AP-1 &  10323 & 10323 & 331 & Gene ID: 3725 & NP\_002219\\
+AhR &  47530 & 47530 & 848 & Gene ID: 196 & NP\_001621\\
+CD11b &  72925 & 72925 & 1153 & Gene ID: 3684 & NP\_001139280 \\
+CD14 &  8974 & 8974 & 375 & Gene ID: 929 & NP\_001035110\\
+CD38 &  174978 & 74978 & 300 & Gene ID: 952 & NP\_001766\\
+C/EBP$\alpha$ &  2630 & 2630 & 393 & Gene ID: 1050 & NP\_001274353.1\\
+E2F &  17919 & 17919 & 437 & Gene ID: 1869 & NP\_005216\\
+Egr-1 &  10824 & 10824 & 543 & Gene ID: 1958 & NP\_001955\\
+Gfi-1 &  13833 & 13833 & 422 & Gene ID: 2672 & NP\_005254\\
+IRF-1 &  16165 & 16165 & 325 & Gene ID: 3659 & NP\_002189\\
+Oct1 &  206516 & 206516 & 741.33 & Gene ID: 5451 & NP\_002688.3, NP\_001185712.1, NP\_001185715.1\\
+Oct4 &  6356 & 6356 & 206.33 &Gene ID: 5460 & NP\_001167002, NP\_001167015, NP\_001167016\\
+P21 &  15651 & 15651 & 198 & NG\_009364.1 & NP\_001621\\
+P47 &  3074 & 3074 & 390 & GenBank: AF003533.1 & NP\_000256 \\
+PPAR$\gamma$ &  153507 & 153507 & 250 & Gene ID: 5468 &  NP\_001317544\\
+PU.1 &  40782 & 40782 & 270.5 & Gene ID: 6688 & NP\_001074016, NP\_003111 \\
+
+\hline
+\end{supertabular}
+\end{scriptsize}
+\end{center}
+
+\clearpage
+
 \begin{figure}[!t]

### 2017-05-12 Paper_v2.tex
Some changes ...

@@
 The model describes a key architectural feature of ATRA-induced differentiation,
-positive feedback between an ATRA-inducible signalsome complex involving many proteins including Vav1, a guanine nucleotide exchange factor,
+reinforcing feedback between an ATRA-inducible signalsome complex involving many proteins including Vav1, a guanine nucleotide exchange factor,
 and the activation of the mitogen activated protein kinase (MAPK) cascade.
+We decomposed the effective model into three modules;
+a signal initiation module that sensed and transformed an ATRA signal into program activation signals;
+a signal integration module that controlled the expression of upstream transcription factors; and
+a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
 The model, which was developed by integrating logical rules with kinetic modeling, was significantly smaller than previous models.
@@
 Conformational experiments supported ATRA-induced bistability.
-These findings, combined with other literature evidence, suggest that positive feedback is central to a diversity of cell fate programs.
+Additionally, the model captured intermediate and phenotypic gene expression data.
+Knockout analysis of the model suggested Gfi-1 and PPAR$\gamma$ were critical to the ATRA-induced differentiation program.
+These findings, combined with other literature evidence, suggested that reinforcing feedback is central to a diversity of cell fate programs.
 
@@
 is a promising approach for the treatment of many cancers \cite{Bushue:2010aa,Tang:2011aa,Cheung:2012aa}.
-For example, ATRA treatment induces remission in 80–90\% of promyelocytic leukemia (APL) PML-RAR$\alpha$-positive patients \cite{Nilsson:1984aa}, thereby transforming a fatal diagnosis into a manageable disease.
+For example, ATRA treatment induces remission in 80–90\% of promyelocytic leukemia (APL) PML-RAR$\alpha$-positive patients \cite{Nilsson:1984aa},
+thereby transforming a fatal diagnosis into a manageable disease \cite{Coombs:2015aa}.
 However, remission is sometimes not durable and relapsed cases exhibit emergent ATRA resistance \cite{Warrell:1993aa,Freemantle:2003aa}.
 To understand the basis of this resistance, we must first understand the ATRA-induced differentiation program.
-Toward this challenge, lessons learned in model systems, such as the lineage-uncommitted human myeloblastic cell line HL-60, could inform our analysis of the more
-complex differentiation programs occurring in patients. Patient derived HL-60 leukemia cells have been a durable experimental model since the 1970’s to study differentiation \cite{Breitman1980}.
+Toward this challenge, lessons learned in model systems, such as the lineage-uncommitted human myeloblastic cell line HL-60 reported to closely resemble patient derived cells \cite{Harrison:2016aa},
+could inform our analysis of the differentiation programs occurring in patients.
+Patient derived HL-60 leukemia cells have been a durable experimental model since the 1970’s to study differentiation \cite{Breitman1980}.
 HL-60 undergoes cell cycle arrest and either myeloid or monocytic differentiation following stimulation; ATRA induces G1/G0-arrest and myeloid differentiation in HL-60 cells,
@@
 Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
-ATRA drives sustained MEK-dependent activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
-MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate \cite{Hong2001}.
-ATRA (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
+ATRA drives sustained activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
+Betraying a feedback loop, MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate in response to ATRA \cite{Hong2001}.
+Retinoic acid (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
 RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation  \cite{Hong2001},
-and the formation of an ATRA-inducible signalsome complex at the membrane which drives MAPK activation through a yet to be identified kinase activity.
+and the formation of an ATRA-inducible signalsome complex at the membrane, which drives MAPK activation.
 While the makeup of the signalsome complex is not yet known,
-we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, as well as IRF-1 transcription factors \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
-Signalsome formation and activity is driven by ATRA-induced expression of CD38 and the putative heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
-BLR1, identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
-is necessary for MAPK activation and differentiation \cite{WANG2004}, and is also involved with signalsome activity.
-Studies of the BLR1 promoter identified a 5' 17bp GT box approximately 1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
+we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, plus transcription factors
+AhR and IRF1 \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
+Signalsome activity is driven by ATRA-induced expression of CD38 and putatively the heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
+BLR1 (also known as CXCR5), identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
+is necessary for MAPK activation and differentiation \cite{WANG2004}, and drives signalsome activity.
+Studies of the BLR1 promoter identified a non-canonical RARE site consisting of a 17 bp GT box approximately
+1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
 Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
 are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
-BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and differentiation \cite{Wang2008}.
+BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and ATRA-induced terminal differentiation \cite{Wang2008}.
+In particular, Raf phosphorylation of the NFATc3 transcription factors at the BLR1 promoter enables transcriptional activation at the RARE by ATRA bound to RAR/RXR \cite{Geil:2014aa}.
 BLR1 knock-out cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
@@
 Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the signalsome.
-A previous computational study of ATRA-induced differentiation in HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
+A previous computational study of ATRA-induced differentiation of HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
 to explain ATRA-induced sustained MAPK activation, and the expression of a limited number of functional differentiation markers \cite{Tasseff2011}.
@@
 In addition, the expression of intermediate and phenotypic differentiation markers as also captured following ATRA exposure.
-Lastly, we showed by through immunoprecipitation studies, that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the signalsome complex.
+Lastly, we showed through immunoprecipitation and inhibitor studies,
+that the guanine nucleotide exchange factor Vav1 is potentially a new ATRA-inducible member of the signalsome complex functioning as a regulator that contributes to signal amplification in the signalsome.
 Taken together, these findings when combined with other literature evidence,
 suggested that positive feedback architectures are central to differentiation programs generally, and necessary for ATRA-induced differentiation.
+The model answers a biologically important question that is not easily experimentally attacked, namely given the complexity of the signaling machine and the pathways it embodies, is there a critical small suite of molecules that are the action elements seminal to eliciting ATRA-induced cell differentiation and G0 arrest.
 
@@
 We decomposed the ATRA program into three modules;
-a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (Trigger) signals (Fig. \ref{fig:network}A);
+a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (Trigger) signals (Fig. \ref{fig:network}A);
 a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals (Fig. \ref{fig:network}B); and
@@
 unactivated Trigger and cRaf protein levels.
-Lastly, because the population of HL-60 cells was dividing (at least before ATRA-induced cell cycle arrest), we also considered a dilution term in all balance equations.
+Lastly, because the population of HL-60 cells was dividing, we also considered a dilution term in all balance equations.
 The signal initiation module contained nine differential equations, while the signal integration and phenotype modules were collectively encoded by 54 differential equations.
-Model parameters were taken literature (Table \ref{tbl:model-parameters}), or estimated from experimental data using heuristic optimization (see materials and methods).
+Model parameters were taken from literature (Table \ref{tbl:model-parameters}), or estimated from experimental data using heuristic optimization (see materials and methods).
 
@@
 cRaf-pS621 following the addition of 1$\mu$M ATRA. We focused on the S621 phosphorylation site of cRaf since enhanced phosphorylation at this site is a defining
-characteristic of sustained MAPK activation in HL-60. The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A)
+characteristic of sustained MAPK signaling activation in HL-60. The effective model captured both ATRA-induced BLR1 expression (Fig. \ref{fig:model-fitting}A)
 and sustained phosphorylation of cRaf-pS621 (Fig. \ref{fig:model-fitting}B) in a growing population of HL-60 cells.
-Together, the reinforcing positive feedback between the signalsome and MAPK led to sustained activation over multiple cellular generations.
+Together, the reinforcing feedback within the signalsome and its embedded MAPK signaling axis led to sustained activation over multiple cellular generations.
 However, the effective model failed to capture the decline of BLR1 message after 48 hr of ATRA exposure.
@@
 The signal initiation model was bistable with respect to ATRA induction (Fig. \ref{fig:model-fitting}C-D).
-Phaseplane analysis predicted two stable steady-states when ATRA was present below a critical threshold, and only a single steady-state above the threshold (Fig. \ref{fig:model-fitting}C).
-In the lower stable state, neither the signalsome nor cRaf-pS621 were present (thus, the differentiation program was deactivated).
-However, at the high stable state, both the signalsome and cRaf-pS621 were present, allowing for sustained activation and differentiation.
+Phaseplane analysis predicted two stable steady-states when ATRA was present below a critical threshold (Fig. \ref{fig:model-fitting}C),
+and only a single steady-state above the threshold (Fig. \ref{fig:model-fitting}D).
+In the lower stable state, neither the signalsome nor cRaf-pS621 were present (thus, the differentiation program was inactive).
+However, at the higher stable state, both the signalsome and cRaf-pS621 were present, allowing for sustained activation and differentiation.
 Interestingly, when ATRA was above a critical threshold, only the activated state was accessible (Fig. \ref{fig:model-fitting}D).
@@
 threshold was near this concentration. Next, we conducted ATRA washout experiments to determine if activated cells remained activated in the absence of ATRA.
-HL-60 cells locked into an activated state remained activated following ATRA withdraw (Fig. \ref{fig:model-predictions}).
+HL-60 cells locked into an activated state remained activated following ATRA withdraw (Fig. \ref{fig:model-predictions}C).
 This sustained activation resulted from reinforcing feedback between the signalsome and the MAPK pathway.
@@
 from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
-However, the bulk of the model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
+However, the bulk of the model parameters were taken from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
 The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
@@
 The Gfi-1 and PPAR$\gamma$ proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
-We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
+We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaf-S621 signals.
 The difference between the system state with and without the gene knockouts (encoded as a normalized state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
@@
 Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBP$\alpha$ expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).
-On the other hand, deletion of PPAR$\gamma$ led to $>$8 fold downregulation of CD38, p21, IRF1 and OCT1 (Fig. \ref{fig:model-response-ko-simulations},red).
+On the other hand, deletion of PPAR$\gamma$ led to $>$8 fold downregulation of CD38, p21, IRF1 and Oct1 (Fig. \ref{fig:model-response-ko-simulations},red).
 Both knockouts slightly increased E2F expression, but neither influenced the expression of p47Phox.
@@
 Interrogation of the Raf interactome suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
-Western blot analysis using total Raf and pS621 Raf specific antibodies confirmed the
+Western blot analysis using total Raf and Raf-pS621 specific antibodies confirmed the
 presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
@@
 However, only the associations between Raf and Vav1, and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
-Furthermore, the Vav1 and Src associations were correlated with pS621 Raf abundance in the precipitate.
+Furthermore, the Vav1 and Src associations were correlated with Raf-pS621 abundance in the precipitate.
 Others proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
@@
 In this study, we presented an effective model of ATRA-inducible differentiation of HL-60 cells.
-The model consisted of three modules: a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (Trigger) signals;
+The model consisted of three modules: a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (Trigger) signals;
 a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
@@
 The expression of other neutrophilic markers, such as CD38, also decreased in Gfi-1$^{-/-}$ + PPAR$\gamma$$^{-/-}$ cells.
-On the other hand, the expression of reactive oxygen metabolic markers, or other important transcription factors such as OCT4 did not change.
+On the other hand, the expression of reactive oxygen metabolic markers, or other important transcription factors such as Oct4 did not change.
 For example, model analysis suggested that the C/EBP$\alpha$ dependent interaction of PU.1 with the \textit{NCF1} gene, which encodes the p47Phox protein, was the most sensitive PU.1 connection; deletion of this connection removed the ability of the system to express p47Phox.
@@
 We decomposed the ATRA-induced differentiation program into three modules;
-a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (activated Trigger) signals;
+a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (activated Trigger) signals;
 a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
@@
 \subsubsection*{Signaling model equations.}
-The signal initiation, and integration modules required the abundance of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
+The signal initiation, and integration modules required the abundance of cRaf-pS621 and ATRA-RAR/RXR (activated Trigger) as inputs.
 However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
@@
 W.D and J.V analyzed the model ensemble, and generated figures for the manuscript.
-The manuscript was prepared and edited for publication by W.D, A.Y and J.V.
+The manuscript was prepared and edited for publication by W.D, R.B, A.Y and J.V.
 

### 2017-05-12 Paper_v2.tex
Changes

@@
 we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, plus transcription factors
-AhR and IRF1 \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
+AhR and IRF1 \cite{Congleton:2012fk,Shen:2011aa,Shen:2009aa,Yen2006,Marchisio:1998aa}.
 Signalsome activity is driven by ATRA-induced expression of CD38 and putatively the heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
@@
 a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors (Fig. \ref{fig:network}C).
+In particular, Trigger (a surrogate for the RAR$\alpha$/RXR transcriptional complex)
+regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and Egr-1.
+In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or p47Phox.
 Each component of these modules was described by a mRNA and protein balance equation.
@@
 The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
-which then in turn activated the phenotypic program. In particular, Trigger (a surrogate for the RAR$\alpha$/RXR transcriptional complex)
-regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and Egr-1.
-In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or p47Phox.
+which then in turn activated the phenotypic program.
 We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
@@
 Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
-Next, we explored which nodes and interactions between nodes in the signal integration module most influenced the system response.
+Next, we explored which proteins and protein interactions in the signal integration module most influenced the system response.
 
@@
 The Gfi-1 and PPAR$\gamma$ proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
-We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaf-S621 signals.
+We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which proteins
+controlled the processing of the Trigger and cRaf-S621 signals.
 The difference between the system state with and without the gene knockouts (encoded as a normalized state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
@@
 
-The composition of the signalsome, and the kinase ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
+The full composition of the signalsome, and the kinase therein ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
 To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
@@
 presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
-Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf, suggesting a direct physical interaction was possible.
+Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 co-precipitated with Raf, suggesting their co-existence in a complex was possible.
 However, only the associations between Raf and Vav1, and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
 Furthermore, the Vav1 and Src associations were correlated with Raf-pS621 abundance in the precipitate.
-Others proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
+Other proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
 The remaining 14 proteins were expressed in whole cell lysate (Fig. \ref{fig:ip-data}B),
-but were not detectable in the precipitate of Raf IP.
+but were not detectable in the immuno-precipitate with Raf IP; consistent with the potential importance of the Raf-Vav interaction for signaling, it paralleled Raf phosphorylation at S621, a putative telltale of the activated kinase.
 Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
@@
 Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with Raf activation following ATRA-treatment.
-Previous studies demonstrated that a Vav1-Slp76-Cbl-CD38 complex plays an important role in ATRA-induced MAPK activation and differentiation of HL-60 cells \cite{Shen2009}.
+Previous studies demonstrated that a Vav1-Slp76-Cbl-CD38 complex plays an important role in ATRA-induced MAPK activation and differentiation of HL-60 cells \cite{Shen:2009aa}.
 Here we did not observe direct interaction of Raf with Cbl or Slp76;
-however, this complex could be involved upstream. Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
+however, this interaction could could be involved upstream.
+Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.
 
@@
 % furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevented
-% ATRA-induced MAPK activation \cite{Shen2009}.
+% ATRA-induced MAPK activation \cite{Shen:2009aa}.
 % Thus, interaction of Cbl-Slp76-Vav1 and CD38 appears to be required for transmission of the ATRA signal by the signalsome.
@@
 Lastly, analysis of the response of the model to perturbations identified Gfi-1 and PPAR$\gamma$ as master regulators of ATRA-induced differentiation.
-We also reported a new ATRA-inducible component of the signalsome, Vav1.
+We also reported an ATRA-inducible component of the signalsome, Vav1.
 Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
@@
 For example, Egr-1/2 expression (associated with a macrophage lineage) increased, while Gfi-1 expression (associated with a neutrophil lineage)
-remained constant following ATRA exposure.
+was unchanged following ATRA exposure.
 Thus, HL-60 cells, which are a less mature cancer cell line, exhibited a non-canonical expression pattern.
@@
 %
-Analysis of the signal integration and phenotype modules suggested that Gfi-1 and PPAR$\gamma$ were required for ATRA-induced differentiation in HL-60 cells.
+Analysis of the signal integration and phenotype modules suggested Gfi-1 and PPAR$\gamma$ proteins were important regulators of ATRA-induced signal integration and phenotypic change.
 Model analysis showed that PU.1, Egr-1 and C/EBP$\alpha$ expression increased in Gfi-1$^{-/-}$ mutants, where PU.1 expression was upregulated by greater than 8-fold.
@@
 of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
-Interestingly, while an integrin ligand-induced ROS response was blocked in Vav1 knockout neutrophils, TPA was able to bypass the Vav1 requirement and stimulate both ERK phosphorylation
-and ROS induction \cite{Graham2007}.
+Interestingly, while an integrin ligand-induced ROS response was blocked in Vav1 knockout neutrophils,
+TPA was able to bypass the Vav1 requirement and stimulate both ERK phosphorylation and ROS induction \cite{Graham2007}.
 In this study, the TPA-induced ROS response was dependent upon Raf kinase activity, and was mitigated by the addition of GW5074.
@@
 furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevents
-ATRA-induced MAPK activation \cite{Shen2009}.
+ATRA-induced MAPK activation \cite{Shen:2009aa}.
 The literature suggest a variety of possible receptor-signaling pathways, which involve Vav1, for MAPK activation; moreover,
@@
 \end{eqnarray}
+where signaling species abundance was governed by the non-linear algebraic equations $\mathbf{g}\left(p_{1},\hdots,p_{\mathcal{G}},\kappa\right) = \mathbf{0}$.
+The model parameter vector is denoted by $\kappa$.
 The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
@@
 The parameters $a_{0},\hdots,a_{3}$ were estimated directly from cell-cycle measurements (biological replicates) using least-squares.
+The form of the basis function assumed p21 was directly proportional, and E2F inversely proportional to G0-arrest.
+However, this was only one form for the basis functions, many others are possible.
+
 % The abundances of the $p21$ and $E2F$ proteins were taken from the gene expression model solution. The least squares solution for the polynomial
@@
 
-\subsubsection*{Availability of model code.}
+\subsubsection*{Availability of the model code and parameters.}
 The signaling and gene expression model equations, and the parameter estimation procedure, were implemented in the Julia programming language.
 The model equations were solved using the \texttt{ODE23s} routine of the ODE package \cite{Julia}.
-The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
+The model code and parameter ensemble is available under an MIT software license and can be downloaded from http://www.varnerlab.org.
 
@@
 \subsubsection*{Immunoprecipitation and western blotting}
-Approximately $1.2 \times 10^7$ cells were lysed using 400$\mu$L of M-Per lysis buffer from Thermo Scientific (Waltham, MA).
-Lysates were cleared by centrifugation at 16,950 $\times$ g in a micro-centrifuge for 20 min at 4$^{o}$C.
-Lysates were pre-cleared using 100$\mu$L protein A/G Plus agarose beads from Santa Cruz Biotechnology (Santa Cruz, CA) by
-inverting overnight at 4$^{o}$C.  Beads were cleared by centrifugation and total protein concentration
-was determined by a BCA assay (Thermo Scientific, Waltham, MA).  Immunoprecipitations were setup by bringing
-lysate to a concentration of 1g/L in a total volume of 300$\mu$L (M-Per buffer was used for dilution).
-The anti-Raf antibody was added at 3$\mu$L.
-A negative control with no bait protein was also used to exclude the direct interaction of proteins with the A/G beads.
-After 1 hr of inversion at 4$^{o}$C, 20$\mu$L of agarose beads was added and samples were left
-to invert overnight at 4$^{o}$C. Samples were then washed three times with M-Per buffer by centrifugation.
-Finally proteins were eluted from agarose beads using a laemmli loading buffer.
-Eluted proteins were resolved by SDS-PAGE and Western blotting.  Total
-lysate samples were normalized by total protein concentration (20$\mu$g per sample)
-and resolved by SDS-PAGE and Western blotting.  Secondary HRP bound antibody was used for visualization.
+For immunoprecipitation experiments, approximately $1.2 \times 10^7$ cells were lysed as previously described.
+300$\mu$g protein (in 300 $\mu$L total volume) per sample was pre-cleared with Protein A/G beads.
+The beads were pelleted and supernatant was incubated with Raf antibody (3$\mu$L/sample) and beads overnight.
+All incubations included protease and phosphatase inhibitors in M-PER used for lysis with constant rotation at 4$^{o}$C.
+Bead/antibody/protein slurries were then washed and subjected to standard SDS-PAGE analysis as previously described \cite{Shen:2009aa}.
 All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of $\alpha$-p621 Raf which was purchased from Biosource/Invitrogen (Carlsbad, CA),
 and $\alpha$-CK2 from BD Biosciences (San Jose, CA).
+
+% (Shen M, Yen A. c-Cbl interacts with CD38 and promotes retinoic acid-induced differentiation and G0 arrest of human myeloblastic leukemia cells. Cancer Res. 2008 Nov 1;68(21):8761–8769.)
+
+% Approximately $1.2 \times 10^7$ cells were lysed using 400$\mu$L of M-Per lysis buffer from Thermo Scientific (Waltham, MA).
+% Lysates were cleared by centrifugation at 16,950 $\times$ g in a micro-centrifuge for 20 min at 4$^{o}$C.
+% Lysates were pre-cleared using 100$\mu$L protein A/G Plus agarose beads from Santa Cruz Biotechnology (Santa Cruz, CA) by
+% inverting overnight at 4$^{o}$C.  Beads were cleared by centrifugation and total protein concentration
+% was determined by a BCA assay (Thermo Scientific, Waltham, MA).  Immunoprecipitations were setup by bringing
+% lysate to a concentration of 1g/L in a total volume of 300$\mu$L (M-Per buffer was used for dilution).
+% The anti-Raf antibody was added at 3$\mu$L.
+% A negative control with no bait protein was also used to exclude the direct interaction of proteins with the A/G beads.
+% After 1 hr of inversion at 4$^{o}$C, 20$\mu$L of agarose beads was added and samples were left
+% to invert overnight at 4$^{o}$C. Samples were then washed three times with M-Per buffer by centrifugation.
+% Finally proteins were eluted from agarose beads using a laemmli loading buffer.
+% Eluted proteins were resolved by SDS-PAGE and Western blotting.  Total
+% lysate samples were normalized by total protein concentration (20$\mu$g per sample)
+% and resolved by SDS-PAGE and Western blotting.  Secondary HRP bound antibody was used for visualization.
+% All antibodies were purchased from Cell Signaling (Boston, MA) with the exception of $\alpha$-p621 Raf which was purchased from Biosource/Invitrogen (Carlsbad, CA),
+% and $\alpha$-CK2 from BD Biosciences (San Jose, CA).
 

### 2017-05-15 Paper_v2.tex
Fig 9

@@
 Lastly, analysis of the response of the model to perturbations identified Gfi-1 and PPAR$\gamma$ as master regulators of ATRA-induced differentiation.
-We also reported an ATRA-inducible component of the signalsome, Vav1.
+We also found evidence of a prominent regulatory role for a signaling molecule ATRA-inducible component of the signalsome, Vav1.
 Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
@@
 \subsubsection*{Effective gene expression model equations.}
+The ATRA differentiation model was encoded as a system of differential algebraic equations (DAEs).
 We decomposed the ATRA-induced differentiation program into three modules;
@@
 % References -
-\bibliography{Paper_v1}
+\bibliography{Paper_v2}
 
@@
 
+\begin{figure}[!t]
+\includegraphics[width=0.9\textwidth]{./figs/Fig-9-ATRA-Schematic.pdf}
+\caption{Schematic of the hypothetical principal pathways in the ATRA-induced
+signaling that results in cell differentiation in the HL-60 myeloid leukemia model
+\cite{wallace2016rrd,congleton2012src,congleton2014phosphorylation,Shen:2009aa,Bunaciu:2015aa,shen2008c}.
+It is based on modules and feedback loops. There are three main arms (shown top to bottom):
+1. Direct ATRA targeting of RAREs in genes such as CD38 or BLR1;
+2. Formation of a signalsome that has a
+regulatory module that includes Vav (a GEF), CBL and SLP-76 (adaptors), and Lyn (a SFK) that
+regulates a Raf/Mek/Erk axis that incorporates Erk to Raf feedback, where the regulators are
+modulated by AhR and CD38 receptors; and 3. Direct ATRA targeted up regulation of CDKI to
+control RB hypophosphorylation. The Raf/Mek/Erk axis is embedded in the signalsome and
+subject to modulation by the regulators. The output of the signalsome is discharge of the Raf
+from the cytosol to the nucleus where it binds (hyper)phospho-RB and other targets, including
+NFATc3, which enables activation of the RA bound RAR/RXR poised on the BLR1 promoter,
+and also GSK3, phosphorylation of which relieves its inhibitory effect on RAR$\alpha$. CDKI directed
+hypophosphorylation of RB releases Raf sequestered by RB to go to NFATc3, GSK3, and other
+targets. A significant consequence of the nuclear RAF is ergo ultimately to hyperactivate
+transcriptional activation by RAR$\alpha$ to drive differentiation.
+It might be noted that this general proposed model provides a mechanistic rationalization for
+why cell cycle arrest is historically oft times perceived as a feature predisposing phenotypic
+maturation.}
+\end{figure}
+
 \clearpage

### 2017-05-16 Paper_v2.tex
AY and RB changes

@@
 \vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{?}}
+{\par \centering \textbf{To be submitted:}~\emph{Oncotarget}}
 \vspace{0.5in}
@@
 Additionally, the model captured intermediate and phenotypic gene expression data.
-Knockout analysis of the model suggested Gfi-1 and PPAR$\gamma$ were critical to the ATRA-induced differentiation program.
-These findings, combined with other literature evidence, suggested that reinforcing feedback is central to a diversity of cell fate programs.
+Knockout analysis suggested Gfi-1 and PPAR$\gamma$ were critical to the ATRA-induced differentiation program.
+These findings, combined with other literature evidence, suggested that reinforcing feedback is central to hyperactive signaling in a diversity of cell fate programs.
 
@@
 
+%Understanding differentiation programs is an important therapeutic challenge.
+
 \section*{Introduction}
-Understanding the architecture of differentiation programs is an important therapeutic challenge.
 Differentiation induction chemotherapy (DIC), using agents such as the vitamin A derivative all-trans retinoic acid (ATRA),
-is a promising approach for the treatment of many cancers \cite{Bushue:2010aa,Tang:2011aa,Cheung:2012aa}.
+is a promising approach for the treatment of many cancers \cite{Bushue:2010aa,Tang:2011aa,Cheung:2012aa,Uray:2016aa,Li:2017aa}.
 For example, ATRA treatment induces remission in 80–90\% of promyelocytic leukemia (APL) PML-RAR$\alpha$-positive patients \cite{Nilsson:1984aa},
@@
 
-Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
+Sustained mitogen-activated protein kinase (MAPK) pathway activation is a defining feature of ATRA-induced HL-60 differentiation.
 ATRA drives sustained activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
-Betraying a feedback loop, MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate in response to ATRA \cite{Hong2001}.
+Betraying a feedback loop, MEK inhibition results in the loss of ERK as well as Raf phosphorylation and the failure to arrest and differentiate in response to ATRA \cite{Hong2001}.
 Retinoic acid (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
-RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation  \cite{Hong2001},
-and the formation of an ATRA-inducible signalsome complex at the membrane, which drives MAPK activation.
+RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation \cite{Hong2001} in concert with
+the formation of an ATRA-induced signalsome complex at the membrane, which drives MAPK activation.
 While the makeup of the signalsome complex is not yet known,
@@
 Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
-are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
+are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002,Li:2017ab}.
 BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and ATRA-induced terminal differentiation \cite{Wang2008}.
@@
 Taken together, these findings when combined with other literature evidence,
-suggested that positive feedback architectures are central to differentiation programs generally, and necessary for ATRA-induced differentiation.
+suggested that reinforcing feedback was central to differentiation programs generally, and necessary for ATRA-induced differentiation.
 The model answers a biologically important question that is not easily experimentally attacked, namely given the complexity of the signaling machine and the pathways it embodies, is there a critical small suite of molecules that are the action elements seminal to eliciting ATRA-induced cell differentiation and G0 arrest.
@@
 
-The full composition of the signalsome, and the kinase therein ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
-To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
+The full composition of the signalsome, and the kinase therein ultimately responsible for mediating ATRA-induced Raf activation is still not known.
+To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify interactions between Raf and 19 putative interaction partners.
 A panel of 19 possible Raf interaction partners (kinases, GTPases, scaffolding proteins etc)
@@
 Total Raf was used as a bait protein for the immunoprecipitation studies.
-Interrogation of the Raf interactome suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
+Interrogation of the Raf interactome thus suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
+
 Western blot analysis using total Raf and Raf-pS621 specific antibodies confirmed the
@@
 However, only the associations between Raf and Vav1, and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
+The interaction between Vav1 and Raf was one of the most prominent interactions in the panel, and it was crippled by inhibiting Raf.
 Furthermore, the Vav1 and Src associations were correlated with Raf-pS621 abundance in the precipitate.
@@
 but were not detectable in the immuno-precipitate with Raf IP; consistent with the potential importance of the Raf-Vav interaction for signaling, it paralleled Raf phosphorylation at S621, a putative telltale of the activated kinase.
-Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
+Furthermore, treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
 although the signal intensity for Src was notably weak.
@@
 Thus, the model analysis was consistent with this study.
-However, other predictions concerning the behavior of the Gfi-1$^{-/-}$ + PPAR$\gamma$$^{-/-}$ mutants have yet to be verified experimentally.
+However, other predictions concerning the behavior of the Gfi-1$^{-/-}$ + PPAR$\gamma$$^{-/-}$ mutants remain to be tested.
 
@@
 This suggested the proposed Raf-Vav1 architecture was at least consistent with the experimental studies.
-Further, analysis of the Raf-Vav1 model identified bistability in ppERK levels.
-Thus, two possible MAPK activation branches were possible for experimentally testable ATRA values. The analysis also suggested the ATRA-induced Raf-Vav1 architecture could be locked into a
-sustained signaling mode (high ppERK) even in the absence of a ATRA signal.
+Further, analysis of the Raf-Vav1 model identified bistability in phosphorylated ERK levels.
+Thus, two possible MAPK activation branches were possible for experimentally testable ATRA values.
+The analysis also suggested the ATRA-induced Raf-Vav1 architecture could be locked into a
+sustained signaling mode (high phosphorylated ERK) even in the absence of a ATRA signal.
 This locked-in property could give rise to an ATRA-induction memory.
@@
 Although the active state was self-sustaining, the inactive state demonstrated considerable robustness to perturbation.
-For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on ppERK levels over the entire ensemble.
-CD38 expression correlated with the ppERK, suggesting its involvement in the signaling complex.
+For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on phosphorylated ERK levels over the entire ensemble.
+CD38 expression correlated with the phosphorylated ERK, suggesting its involvement in the signaling complex.
 Our computational and experimental results showed that positive feedback,
@@
 \subsubsection*{Effective gene expression model equations.}
-The ATRA differentiation model was encoded as a system of differential algebraic equations (DAEs).
+The ATRA differentiation model was encoded as a system of differential algebraic equations (DAEs) which described signaling and gene expression processes.
+We modeled transcription and translation as Ordinary Differential Equations (ODEs), while signaling processes were assumed to quickly equilibrate and were
+treated as a pseudo steady state system of algebraic equations.  
 We decomposed the ATRA-induced differentiation program into three modules;
@@
 \end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
+In thus study, signalsome abundance was approximated by the abundance of the BLR1 protein;
+BLR1 expression is directly related to Raf nuclear translocation which in turn is related to activated signalsome.
+Thus, BLR1 is an indirect measure of the signalsome.
 The formation of cRaf-pS621 was regulated by only a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
@@
 \end{equation}where $a_{i}$ were unknown parameters, and $\phi_{i}\left(\mathbf{p}(t),t\right)$ denotes a basis function.
-The basis functions were dependent upon the system state; in this study, we assumed $N = 4$ and basis functions of the form:
+The basis functions were dependent upon the system state; in this study, we used N = 4 and basis functions of the form:
 \begin{equation}
@@
 The parameters $a_{0},\hdots,a_{3}$ were estimated directly from cell-cycle measurements (biological replicates) using least-squares.
-The form of the basis function assumed p21 was directly proportional, and E2F inversely proportional to G0-arrest.
-However, this was only one form for the basis functions, many others are possible.
+The form of the basis function assumed p21 was directly proportional, and E2F inversely proportional, to G0-arrest.
+However, this was one of many possible forms for the basis functions.
 
@@
 \subsubsection*{Immunoprecipitation and western blotting}
-For immunoprecipitation experiments, approximately $1.2 \times 10^7$ cells were lysed as previously described.
+For immunoprecipitation experiments, cells were lysed as previously described.
 300$\mu$g protein (in 300 $\mu$L total volume) per sample was pre-cleared with Protein A/G beads.
@@
 & - & Oct4 & \cite{Bunaciu2011}\\
-& - & PU.1 & \\
+& - & PU.1 & \cite{Bock:2017aa} \\
 
@@
 	$e_{X}$ & elongation rate Ribosome & 5 & aa/s & \cite{Bostrom:1986aa} \\
-	$L_{T,o}$ & characteristic gene length & 15,000 & nt & \cite{Meyers2004} \\
-	$L_{X,o}$ & characteristic transcript length & 5,000 & nt & derived \\
+	$L_{T,o}$ & characteristic gene length & 44,192 & nt & \cite{Meyers2004} \\
+	$L_{X,o}$ & characteristic transcript length & 1,374 & nt & derived \\
 	$k_{T}$ & characteristic transcription rate & 1.44 & hr$^{-1}$ & derived \\
@@
 \hline
-\rowcolor[gray]{0.9} Gene Name & Gene (bp) & RNA (bp) & Protein (AA) & Gene ID & Protein ID \\}
-\topcaption{Sequence lengths from NCBI RefSeq database was used in the signal integration and phenotype modules \cite{OLeary:2016aa}.
+\rowcolor[gray]{0.9} Gene Name & Gene (bp) & RNA (bp) & Protein (aa) & Gene ID & Protein ID \\}
+\topcaption{Sequence lengths from NCBI RefSeq database were used in the signal integration and phenotype modules \cite{OLeary:2016aa}.
 The RNA sequence length used represents the total distance of transcription, and assume to be equal to the gene length. }\label{tbl:TF-sequences}

### 2017-05-16 Paper_v2.tex
New Fig 9 edits

@@
 
-We hypothesized that Vav1 is a member of an ATRA-inducible complex which propels sustained MAPK activation, arrest and differentiation.
+We hypothesized that Vav1 is a member of an ATRA-inducible signalsome complex which propels sustained MAPK activation, arrest and differentiation (shown schematically in Fig.
+\ref{fig:signalsome-schematic}).
 Initially, ATRA-induced Vav1 expression drives increased association between Vav1 and Raf.
@@
 \subsubsection*{Effective gene expression model equations.}
-The ATRA differentiation model was encoded as a system of differential algebraic equations (DAEs) which described signaling and gene expression processes.
+The ATRA differentiation model was encoded as a system of differential algebraic equations (DAEs) which described both signaling and gene expression processes.
 We modeled transcription and translation as Ordinary Differential Equations (ODEs), while signaling processes were assumed to quickly equilibrate and were
@@
 abundance of transcription factors, or other regulators influenced the expression of gene $j$.
-The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
-\begin{equation}
-	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
-\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
-and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
-The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
-Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
-Lastly, the $\lambda_{j}$ term denotes the constitutive rate of expression of gene $j$.
 
@@
 \end{equation}
-The numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
-The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using Hill like functions), while
+The numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression,
+was normalized by all possible promoter configurations (denominator).
+The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using Hill functions), while
 the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
-Given this formulation, the rate of constitutive expression was then given by:
+Given the formulation of the control law, the $\lambda_{j}$ term (which denotes the constitutive rate of expression of gene $j$) was given by:
 \begin{equation}
@@
 \end{equation}
+
+The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
+\begin{equation}
+	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
+\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
+and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
+The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
+Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
 If a gene expression process had no modifying factors, $u_{j}=1$.
@@
 we corrected the characteristic translation rate by the ratio of the length of a characteristic transcription normalized by the length of transcript $j$.
+The sequence lengths used in this study are given in Table \ref{tbl:TF-sequences}.
 
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=0.85\textwidth]{./figs/Fig-4-GRN-Simulations.pdf}
+\includegraphics[width=0.70\textwidth]{./figs/Fig-4-GRN-Simulations.pdf}
 \caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
-A: CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
-B: Gene expression at t = 48 hr following ATRA exposure.
-Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-grn-simulations}
+A: Scaled CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
+B: Scaled Gene expression at t = 48 hr following ATRA exposure.
+Gene expression was normalized to expression in the absence of ATRA.
+Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.
+Model simulations were conducted using the ten best parameter sets collected during model identification.
+Solid lines (or bars) denote the mean model performance, while the shaded region (or error bars) denote the 95\% confidence
+interval calculated over the parameter ensemble.}\label{fig:model-grn-simulations}
 \end{figure}
@@
 B: Estimated fraction of HL-60 cells in G0 arrest following ATRA exposure at time t = 0 hr.
-The gray region denotes the 95\% confidence estimate of the polynomial model.
+Solid lines (or bars) denote the mean model performance, while the shaded region (or error bars) denotes the 95\% confidence estimate of the polynomial model.
 Experimental data in panel B was reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-cell-cycle-simulations}
@@
 \includegraphics[width=0.9\textwidth]{./figs/Fig-9-ATRA-Schematic.pdf}
-\caption{Schematic of the hypothetical principal pathways in the ATRA-induced
-signaling that results in cell differentiation in the HL-60 myeloid leukemia model
+\caption{Schematic of the hypothetical principal pathways implicated in ATRA-induced differentiation of HL-60 myeloid leukemia cells
 \cite{wallace2016rrd,congleton2012src,congleton2014phosphorylation,Shen:2009aa,Bunaciu:2015aa,shen2008c}.
-It is based on modules and feedback loops. There are three main arms (shown top to bottom):
-1. Direct ATRA targeting of RAREs in genes such as CD38 or BLR1;
-2. Formation of a signalsome that has a
-regulatory module that includes Vav (a GEF), CBL and SLP-76 (adaptors), and Lyn (a SFK) that
-regulates a Raf/Mek/Erk axis that incorporates Erk to Raf feedback, where the regulators are
-modulated by AhR and CD38 receptors; and 3. Direct ATRA targeted up regulation of CDKI to
-control RB hypophosphorylation. The Raf/Mek/Erk axis is embedded in the signalsome and
-subject to modulation by the regulators. The output of the signalsome is discharge of the Raf
+There are three main arms (shown top to bottom):
+(a) Direct ATRA targeting of RAREs in genes such as CD38 or BLR1;
+(b) Formation of a signalsome that has a regulatory module that includes Vav (a guanine nucleotide exchange factor),
+CBL and SLP-76 (adaptors), and Lyn (a Src family kinase) that
+regulates a Raf/Mek/Erk axis that incorporates Erk to Raf feedback, where the regulators are modulated by AhR and CD38 receptors; and
+(c) Direct ATRA upregulation of CDKIs to control RB hypophosphorylation.
+The Raf/Mek/Erk axis is embedded in the signalsome and
+subject to modulation by the regulators. The signalsome discharges Raf
 from the cytosol to the nucleus where it binds (hyper)phospho-RB and other targets, including
-NFATc3, which enables activation of the RA bound RAR/RXR poised on the BLR1 promoter,
-and also GSK3, phosphorylation of which relieves its inhibitory effect on RAR$\alpha$. CDKI directed
-hypophosphorylation of RB releases Raf sequestered by RB to go to NFATc3, GSK3, and other
-targets. A significant consequence of the nuclear RAF is ergo ultimately to hyperactivate
+NFATc3, which enables activation of the ATRA bound RAR/RXR poised on the BLR1 promoter,
+and GSK3, phosphorylation of which relieves the inhibitory effect on RAR$\alpha$.
+CDKI directed hypophosphorylation of RB releases Raf sequestered by RB to go to NFATc3, GSK3, and other targets.
+A significant consequence of the nuclear Raf is ergo ultimately to hyperactivate
 transcriptional activation by RAR$\alpha$ to drive differentiation.
@@
 why cell cycle arrest is historically oft times perceived as a feature predisposing phenotypic
-maturation.}
+maturation.}\label{fig:signalsome-schematic}
 \end{figure}
